### 2021-07-07 erc-B1/parts/eatr.tex
Initial commit

% ! TeX root = ../erc-B1.tex
\subsection{Bibliometrics and summary of publication activities}

\subsection{10 research highlights}

I now present my achievements through a selection of my publications, in
reverse chronological order. My name appears in bold, names of PhD
students I supervised in italics.

\begin{asparaitem}
\item \todo{Relevant publications, with context} 
% \fullcite can be used to fully cite a publication
\end{asparaitem}

\subsection{Invited presentations}

\noindent\hskip -.5cm\begin{tabular}{>{\bfseries}rll}
  \toprule
  \multicolumn{1}{c}{\bfseries Date}&{\bfseries Location}&{\bfseries
Context or inviting institution}\\
  \midrule
  ~\\
  \bottomrule
\end{tabular}

\subsection{Awards}

### 2021-07-07 article.tex
change template according to WOA guidelines

%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[
% twocolumn,
% hf,
]{ceurart}

%%
%% One can fix some overfulls
% \sloppy

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2020}
\copyrightclause{Copyright for this paper by its authors.
  Use permitted under Creative Commons License Attribution 4.0
  International (CC BY 4.0).}

%%
%% This command is for the conference information
\conference{Woodstock'20: Symposium on the irreproducible science,
  June 01--05, 2020, Woodstock, NY}

%%
%% The "title" command
\title{A better way to format your document for CEUR-WS}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1,2]{Dmitry S. Kulyabov}[%
orcid=0000-0002-0877-7063,
email=kulyabov-ds@rudn.ru,
url=https://yamadharma.github.io/,
]
\address[1]{Peoples' Friendship University of Russia (RUDN University),
  6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation}
\address[2]{Joint Institute for Nuclear Research,
  6 Joliot-Curie, Dubna, Moscow region, 141980, Russian Federation}

\author[3]{Ilaria Tiddi}[%
orcid=0000-0001-7116-9338,
email=i.tiddi@vu.nl,
url=https://kmitd.github.io/ilaria/,
]
\address[3]{Vrije Universiteit Amsterdam, De Boelelaan 1105, 1081 HV Amsterdam, The Netherlands}

\author[4]{Manfred Jeusfeld}[%
orcid=0000-0002-9421-8566,
email=Manfred.Jeusfeld@acm.org,
url=http://conceptbase.sourceforge.net/mjf/,
]
\address[4]{University of Skövde, Högskolevägen 1, 541 28 Skövde, Sweden}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX{} document is presented as an
  article formatted for publication by CEUR-WS in a conference
  proceedings. Based on the ``ceurart'' document class, this article
  presents and explains many of the common variations, as well as many
  of the formatting elements an author may use in the preparation of
  the documentation of their work.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  \LaTeX{} class \sep
  paper template \sep
  paper formatting \sep
  CEUR-WS
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

CEUR-WS's article template provides a consistent \LaTeX{} style for
use across CEUR-WS publications, and incorporates accessibility and
metadata-extraction functionality. This document will explain the
major features of the document class.

If you are new to publishing with CEUR-WS, this document is a valuable
guide to the process of preparing your work for publication.

The ``\verb|ceurart|'' document class can be used to prepare articles
for any CEUR-WS publication, and for any stage of publication, from
review to final ``camera-ready'' copy with {\itshape very} few changes
to the source.

This class depends on the following packages
for its proper functioning:

\begin{itemize}
\item \verb|natbib.sty| for citation processing;
\item \verb|geometry.sty| for margin settings;
\item \verb|graphicx.sty| for graphics inclusion;
\item \verb|hyperref.sty| optional package if hyperlinking is required in
  the document;
\item \verb|fontawesome5.sty| optional package for bells and whistles.
\end{itemize}

All the above packages are part of any
standard \LaTeX{} installation.
Therefore, the users need not be
bothered about downloading any extra packages.

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

\section{Template parameters}

There are a number of template
parameters which modify some part of the \verb|ceurart| document class.
This parameters are enclosed in square
brackets and are a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[parameter]{ceurart}
\end{verbatim}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\verb|twocolumn|}: Two column layout.
\item {\verb|hf|}: Enable header and footer.
\end{itemize}

\section{Front matter}

\subsection{Title Information}

The titles of papers should be either all use the emphasizing
capitalized style or they should all use the regular English (or
native language) style. It does not make a good impression if you or
your authors mix the styles.

Use the \verb|title| command to define the title of your work. Do not
insert line breaks in your title.

\subsection{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification. Multiple authors may share one affiliation. Authors'
names should not be abbreviated; use full first names wherever
possible. Include authors' e-mail addresses whenever possible.

\verb|\author| command have the below options:

\begin{itemize}
\item \verb|style|: Style of author name (chinese)
\item \verb|prefix|: Prefix
\item \verb|suffix|: Suffix
\item \verb|degree|: Degree
\item \verb|role|: Role
\item \verb|orcid|: ORCID
\item \verb|email|: E-mail
\item \verb|url|: URL
\end{itemize}

Author names can have some kinds of marks and notes:
\begin{itemize}
\item affiliation mark: \verb|\author[<num>]|.
% \item email: \verb|\ead{<email>}|,
% \item url: \verb|\ead[url]{<url>}|.
\end{itemize}

The author names and affiliations could be formatted in two ways:
\begin{enumerate}
\item Group the authors per affiliation.
\item Use an explicit mark to indicate the affiliations.
\end{enumerate}

Author block example:
\begin{verbatim}
\author[1,2]{Author Name}[%
    prefix=Prof.,
    degree=D.Sc.,
    role=Researcher,
    orcid=0000-0000-000-0000,
    email=name@example.com,
    url=https://name.example.com
]

\address[1]{Affiliation #1}
\address[2]{Affiliation #2}
\end{verbatim}

\subsection{Abstract and Keywords}

Abstract shall be entered in an environment that starts
with \verb|\begin{abstract}| and ends with
\verb|\end{abstract}|.

\begin{verbatim}
\begin{abstract}
  This is an abstract.
\end{abstract}
\end{verbatim}

The key words are enclosed in a \verb|{keyword}|
environment. Use \verb|\sep| to separate keywords.

\begin{verbatim}
\begin{keywords}
  First keyword \sep 
  Second keyword \sep 
  Third keyword \sep 
  Fourth keyword
\end{keywords}
\end{verbatim}

At the end of front matter add \verb|\maketitle| command.

\section{Sectioning Commands}

Your work should use standard \LaTeX{} sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is not allowed.

\section{Tables}

The ``\verb|ceurart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed \textit{above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \verb|table| to enclose the table's contents and the
table caption. The contents of the table itself must go in the
\verb|tabular| environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table*}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table*}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \verb|table*| to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table}

\section{Math Equations}

You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}

A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \verb|math| environment,
which can be invoked with the usual
\verb|\begin| \ldots \verb|\end| construction or with
the short form \verb|$| \ldots \verb|$|. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX};
this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty} \frac{1}{n} = 0,
\end{math}
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}

A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \verb|equation|
environment. An unnumbered display equation is produced by the
\verb|displaymath| environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX{}; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty} \frac{1}{n} = 0.
\end{equation}
Notice how it is formatted somewhat differently in
the \verb|displaymath|
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  S_{n} = \sum_{i=1}^{n} x_{i} ,
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \lim_{x \to 0} (1 + x)^{1/x} = e
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader. Figure captions go below the figure. Your figures should
also include a description suitable for screen readers, to
assist the visually-challenged to better understand your work.

Figure captions are placed below the figure.

\section{Citations and Bibliographies}

The use of Bib\TeX{} for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
\bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the Bib\TeX{} file.

\subsection{Some examples}

A paginated journal article \cite{Abril07}, an enumerated journal
article \cite{Cohen07}, a reference to an entire issue
\cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
monograph/whole book in a series (see 2a in spec. document)
\cite{Harel79}, a divisible-book such as an anthology or compilation
\cite{Editor00} followed by the same example, however we only output
the series if the volume number is given \cite{Editor00a} (so series
should not be present since it has no vol. no.), a chapter in a
divisible book \cite{Spector90}, a chapter in a divisible book in a
series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
an article in a proceedings (of a conference, symposium, workshop for
example) (paginated proceedings article) \cite{Andler79}, a
proceedings article with all possible elements \cite{Smith10}, an
example of an enumerated proceedings article \cite{VanGundy07}, an
informally published work \cite{Harel78}, a doctoral dissertation
\cite{Clarkson85}, a master's thesis: \cite{anisi03}, an online
document / world wide web resource \cite{Thornburg01, Ablamowicz07,
  Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2)
\cite{Novak03} and \cite{Lee05} and (Case 3) a patent
\cite{JoeScientist001}, work accepted for publication \cite{rous08},
prolific author \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other
cites might contain `duplicate' DOI and URLs (some SIAM articles)
\cite{Kirschmer:2010:AEI:1958016.1958018}. Multi-volume works as books
\cite{MR781536} and \cite{MR781537}. A couple of citations with DOIs:
\cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
citations: \cite{TUGInstmem, Thornburg01, R, UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
\begin{acknowledgments}
  These are different acknowledgments.
\end{acknowledgments}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered
{\verb|\section|}; please use the ``{\verb|acknowledgments|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
\appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered.

%%
%% The acknowledgments section is defined using the "acknowledgments" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acknowledgments}
  Thanks to the developers of ACM consolidated LaTeX styles
  \url{https://github.com/borisveytsman/acmart} and to the developers
  of Elsevier updated \LaTeX{} templates
  \url{https://www.ctan.org/tex-archive/macros/latex/contrib/els-cas-templates}.  
\end{acknowledgments}

%%
%% Define the bibliography file to be used
\bibliography{biblio}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Online Resources}

The sources for the ceur-art style are available via
\begin{itemize}
\item \href{https://github.com/yamadharma/ceurart}{GitHub},
\item \href{https://www.overleaf.com/project/5e76702c4acae70001d3bc87}{Overleaf},
\item
  \href{https://www.overleaf.com/latex/templates/template-for-submissions-to-ceur-workshop-proceedings-ceur-ws-dot-org/hpvjjzhjxzjk}{Overleaf
    template}.
\end{itemize}

\end{document}

%%
%% End of file

### 2021-07-07 article.tex
add requirement for action

@@
 %% This command is for the conference information
-\conference{Woodstock'20: Symposium on the irreproducible science,
-  June 01--05, 2020, Woodstock, NY}
+\conference{22nd Workshop "From Objects to Agents" 
+Bologna, Italy, 1-3 September 2021 }
 
@@
 %% The "title" command
-\title{A better way to format your document for CEUR-WS}
+\title{Towards a Neural Network model fro Aggregate Computing}
 

### 2021-07-07 article.tex
draft for abstract

@@
 ]{ceurart}
-
+\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
 %%
@@
 %% The "title" command
-\title{Towards a Neural Network model fro Aggregate Computing}
+\title{Towards a Neural Network model for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
 
@@
 %% the authors and their affiliations.
-\author[1,2]{Dmitry S. Kulyabov}[%
-orcid=0000-0002-0877-7063,
-email=kulyabov-ds@rudn.ru,
-url=https://yamadharma.github.io/,
+\author[1]{Gianluca Aguzzi}[%
+orcid=todo,
+email=gianluca.aguzzi@unibo.it,
+url=https://todo/,
 ]
-\address[1]{Peoples' Friendship University of Russia (RUDN University),
-  6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation}
-\address[2]{Joint Institute for Nuclear Research,
-  6 Joliot-Curie, Dubna, Moscow region, 141980, Russian Federation}
-
-\author[3]{Ilaria Tiddi}[%
-orcid=0000-0001-7116-9338,
-email=i.tiddi@vu.nl,
-url=https://kmitd.github.io/ilaria/,
+\author[2]{Roberto Casadei}[%
+orcid=todo,
+email=roby.casadei@unibo.it,
+url=todo,
 ]
-\address[3]{Vrije Universiteit Amsterdam, De Boelelaan 1105, 1081 HV Amsterdam, The Netherlands}
-
-\author[4]{Manfred Jeusfeld}[%
-orcid=0000-0002-9421-8566,
-email=Manfred.Jeusfeld@acm.org,
-url=http://conceptbase.sourceforge.net/mjf/,
+\author[2]{Mirko Viroli}[%
+orcid=todo,
+email=mirko.viroli@unibo.it,
+url=todo,
 ]
-\address[4]{University of Skövde, Högskolevägen 1, 541 28 Skövde, Sweden}
-
+\address[1]{Alma Mater Studiorum - Università di Bologna,
+  Cesena, Italy}
 %%
@@
 \begin{abstract}
-  A clear and well-documented \LaTeX{} document is presented as an
-  article formatted for publication by CEUR-WS in a conference
-  proceedings. Based on the ``ceurart'' document class, this article
-  presents and explains many of the common variations, as well as many
-  of the formatting elements an author may use in the preparation of
-  the documentation of their work.
+  Artificial intelligence is here to stay. 
+  In particular, Deep Learning techniques and models in the last years outperform hand-craft human solutions in different contexts, ranging from computer vision to videogames.
+  Currently, though, an all-encompassing Eeep learning model does not exist, so researchers explore the possibility to apply these outstanding techniques in different fields.
+  One of the most challenging is Cyber-Physical System. Here, engineers have to handle large scale, global-to-local behaviour mapping,
+  highly environment stochasticity. 
+  Historically, researchers favour handling them with self-organisation, by which a collective behaviour emerges from continuously local interaction between simple components.
+  A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing. Even though is promising for expressing high-level collective behaviour, 
+  it needs to build low-level blocks that tend to be tricky. 
+  On the opposite side, some research lines apply deep learning (supervised, reinforcement, non-supervised and self-supervised)
+  along with CPS. Currently, though, no one gain popularity and solution tend to be application-specific.
+  So, we outline a Neural Network model that could be applied in Aggregate Computing, making the possibility to combine declarative and automatic approaches that can easily scale in application complexity.
 \end{abstract}
@@
 \begin{keywords}
-  \LaTeX{} class \sep
-  paper template \sep
-  paper formatting \sep
-  CEUR-WS
+  Cyber-Physical System \sep
+  Aggregate Computing \sep
+  Deep Learning \sep
+  Reinforcement Learning \sep
 \end{keywords}
@@
 
-CEUR-WS's article template provides a consistent \LaTeX{} style for
-use across CEUR-WS publications, and incorporates accessibility and
-metadata-extraction functionality. This document will explain the
-major features of the document class.
+\section{Background and Related Work}
 
-If you are new to publishing with CEUR-WS, this document is a valuable
-guide to the process of preparing your work for publication.
-
-The ``\verb|ceurart|'' document class can be used to prepare articles
-for any CEUR-WS publication, and for any stage of publication, from
-review to final ``camera-ready'' copy with {\itshape very} few changes
-to the source.
-
-This class depends on the following packages
-for its proper functioning:
-
-\begin{itemize}
-\item \verb|natbib.sty| for citation processing;
-\item \verb|geometry.sty| for margin settings;
-\item \verb|graphicx.sty| for graphics inclusion;
-\item \verb|hyperref.sty| optional package if hyperlinking is required in
-  the document;
-\item \verb|fontawesome5.sty| optional package for bells and whistles.
-\end{itemize}
-
-All the above packages are part of any
-standard \LaTeX{} installation.
-Therefore, the users need not be
-bothered about downloading any extra packages.
-
-\section{Modifications}
-
-Modifying the template --- including but not limited to: adjusting
-margins, typeface sizes, line spacing, paragraph and list definitions,
-and the use of the \verb|\vspace| command to manually adjust the
-vertical spacing between elements of your work --- is not allowed.
-
-\section{Template parameters}
-
-There are a number of template
-parameters which modify some part of the \verb|ceurart| document class.
-This parameters are enclosed in square
-brackets and are a part of the {\verb|documentclass|} command:
-\begin{verbatim}
-  \documentclass[parameter]{ceurart}
-\end{verbatim}
-
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\verb|twocolumn|}: Two column layout.
-\item {\verb|hf|}: Enable header and footer.
-\end{itemize}
-
-\section{Front matter}
-
-\subsection{Title Information}
-
-The titles of papers should be either all use the emphasizing
-capitalized style or they should all use the regular English (or
-native language) style. It does not make a good impression if you or
-your authors mix the styles.
-
-Use the \verb|title| command to define the title of your work. Do not
-insert line breaks in your title.
-
-\subsection{Authors and Affiliations}
-
-Each author must be defined separately for accurate metadata
-identification. Multiple authors may share one affiliation. Authors'
-names should not be abbreviated; use full first names wherever
-possible. Include authors' e-mail addresses whenever possible.
-
-\verb|\author| command have the below options:
-
-\begin{itemize}
-\item \verb|style|: Style of author name (chinese)
-\item \verb|prefix|: Prefix
-\item \verb|suffix|: Suffix
-\item \verb|degree|: Degree
-\item \verb|role|: Role
-\item \verb|orcid|: ORCID
-\item \verb|email|: E-mail
-\item \verb|url|: URL
-\end{itemize}
-
-Author names can have some kinds of marks and notes:
-\begin{itemize}
-\item affiliation mark: \verb|\author[<num>]|.
-% \item email: \verb|\ead{<email>}|,
-% \item url: \verb|\ead[url]{<url>}|.
-\end{itemize}
-
-The author names and affiliations could be formatted in two ways:
-\begin{enumerate}
-\item Group the authors per affiliation.
-\item Use an explicit mark to indicate the affiliations.
-\end{enumerate}
-
-Author block example:
-\begin{verbatim}
-\author[1,2]{Author Name}[%
-    prefix=Prof.,
-    degree=D.Sc.,
-    role=Researcher,
-    orcid=0000-0000-000-0000,
-    email=name@example.com,
-    url=https://name.example.com
-]
-
-\address[1]{Affiliation #1}
-\address[2]{Affiliation #2}
-\end{verbatim}
-
-\subsection{Abstract and Keywords}
-
-Abstract shall be entered in an environment that starts
-with \verb|\begin{abstract}| and ends with
-\verb|\end{abstract}|.
-
-\begin{verbatim}
-\begin{abstract}
-  This is an abstract.
-\end{abstract}
-\end{verbatim}
-
-The key words are enclosed in a \verb|{keyword}|
-environment. Use \verb|\sep| to separate keywords.
-
-\begin{verbatim}
-\begin{keywords}
-  First keyword \sep 
-  Second keyword \sep 
-  Third keyword \sep 
-  Fourth keyword
-\end{keywords}
-\end{verbatim}
-
-At the end of front matter add \verb|\maketitle| command.
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX{} sectioning commands:
-\verb|section|, \verb|subsection|, \verb|subsubsection|, and
-\verb|paragraph|. They should be numbered; do not remove the numbering
-from the commands.
-
-Simulating a sectioning command by setting the first word or words of
-a paragraph in boldface or italicized text is not allowed.
-
-\section{Tables}
-
-The ``\verb|ceurart|'' document class includes the ``\verb|booktabs|''
-package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
-high-quality tables.
-
-Table captions are placed \textit{above} the table.
-
-Because tables cannot be split across pages, the best placement for
-them is typically the top of the page nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \verb|table| to enclose the table's contents and the
-table caption. The contents of the table itself must go in the
-\verb|tabular| environment, to be aligned properly in rows and
-columns, with the desired horizontal and vertical rules.
-
-Immediately following this sentence is the point at which
-Table~\ref{tab:freq} is included in the input file; compare the
-placement of the table here with the table in the printed output of
-this document.
-
-\begin{table*}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table*}
-
-To set a wider table, which takes up the whole width of the page's
-live area, use the environment \verb|table*| to enclose the table's
-contents and the table caption.  As with a single-column table, this
-wide table will ``float'' to a location deemed more
-desirable. Immediately following this sentence is the point at which
-Table~\ref{tab:commands} is included in the input file; again, it is
-instructive to compare the placement of the table here with the table
-in the printed output of this document.
-
-\begin{table}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-\section{Math Equations}
-
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of the three are
-discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-
-A formula that appears in the running text is called an inline or
-in-text formula.  It is produced by the \verb|math| environment,
-which can be invoked with the usual
-\verb|\begin| \ldots \verb|\end| construction or with
-the short form \verb|$| \ldots \verb|$|. You can use any of the symbols
-and structures, from $\alpha$ to $\omega$, available in
-\LaTeX~\cite{Lamport:LaTeX};
-this section will simply show a few
-examples of in-text equations in context. Notice how this equation:
-\begin{math}
-  \lim_{n\rightarrow \infty} \frac{1}{n} = 0,
-\end{math}
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \verb|equation|
-environment. An unnumbered display equation is produced by the
-\verb|displaymath| environment.
-
-Again, in either environment, you can use any of the symbols and
-structures available in \LaTeX{}; this section will just give a couple
-of examples of display equations in context.  First, consider the
-equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty} \frac{1}{n} = 0.
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \verb|displaymath|
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  S_{n} = \sum_{i=1}^{n} x_{i} ,
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \lim_{x \to 0} (1 + x)^{1/x} = e
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures. One or
-more images can be placed within a figure. If your figure contains
-third-party material, you must clearly identify it as such, as shown
-in the example below.
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
-    Ewing, Inc. [Public domain], via Wikimedia
-    Commons. (\url{https://goo.gl/VLCRBB}).}
-\end{figure}
-
-Your figures should contain a caption which describes the figure to
-the reader. Figure captions go below the figure. Your figures should
-also include a description suitable for screen readers, to
-assist the visually-challenged to better understand your work.
-
-Figure captions are placed below the figure.
-
-\section{Citations and Bibliographies}
-
-The use of Bib\TeX{} for the preparation and formatting of one's
-references is strongly recommended. Authors' names should be complete
---- use full first names (``Donald E. Knuth'') not initials
-(``D. E. Knuth'') --- and the salient identifying features of a
-reference should be included: title, year, volume, number, pages,
-article DOI, etc.
-
-The bibliography is included in your source document with these two
-commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
-\bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
-suffix, of the Bib\TeX{} file.
-
-\subsection{Some examples}
-
-A paginated journal article \cite{Abril07}, an enumerated journal
-article \cite{Cohen07}, a reference to an entire issue
-\cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
-monograph/whole book in a series (see 2a in spec. document)
-\cite{Harel79}, a divisible-book such as an anthology or compilation
-\cite{Editor00} followed by the same example, however we only output
-the series if the volume number is given \cite{Editor00a} (so series
-should not be present since it has no vol. no.), a chapter in a
-divisible book \cite{Spector90}, a chapter in a divisible book in a
-series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
-an article in a proceedings (of a conference, symposium, workshop for
-example) (paginated proceedings article) \cite{Andler79}, a
-proceedings article with all possible elements \cite{Smith10}, an
-example of an enumerated proceedings article \cite{VanGundy07}, an
-informally published work \cite{Harel78}, a doctoral dissertation
-\cite{Clarkson85}, a master's thesis: \cite{anisi03}, an online
-document / world wide web resource \cite{Thornburg01, Ablamowicz07,
-  Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2)
-\cite{Novak03} and \cite{Lee05} and (Case 3) a patent
-\cite{JoeScientist001}, work accepted for publication \cite{rous08},
-prolific author \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other
-cites might contain `duplicate' DOI and URLs (some SIAM articles)
-\cite{Kirschmer:2010:AEI:1958016.1958018}. Multi-volume works as books
-\cite{MR781536} and \cite{MR781537}. A couple of citations with DOIs:
-\cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
-citations: \cite{TUGInstmem, Thornburg01, R, UMassCitations}.
-
-\section{Acknowledgments}
-
-Identification of funding sources and other support, and thanks to
-individuals and groups that assisted in the research and the
-preparation of the work should be included in an acknowledgment
-section, which is placed just before the reference section in your
-document.
-
-This section has a special environment:
-\begin{verbatim}
-\begin{acknowledgments}
-  These are different acknowledgments.
-\end{acknowledgments}
-\end{verbatim}
-so that the information contained therein can be more easily collected
-during the article metadata extraction phase, and to ensure
-consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered
-{\verb|\section|}; please use the ``{\verb|acknowledgments|}'' environment.
-
-\section{Appendices}
-
-If your work needs an appendix, add it before the
-``\verb|\end{document}|'' command at the conclusion of your source
-document.
-
-Start the appendix with the ``\verb|appendix|'' command:
-\begin{verbatim}
-\appendix
-\end{verbatim}
-and note that in the appendix, sections are lettered, not
-numbered.
-
-%%
-%% The acknowledgments section is defined using the "acknowledgments" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acknowledgments}
-  Thanks to the developers of ACM consolidated LaTeX styles
-  \url{https://github.com/borisveytsman/acmart} and to the developers
-  of Elsevier updated \LaTeX{} templates
-  \url{https://www.ctan.org/tex-archive/macros/latex/contrib/els-cas-templates}.  
-\end{acknowledgments}
-
+\section{Conclusion and Future Work}
 %%
@@
 
-\section{Online Resources}
-
-The sources for the ceur-art style are available via
-\begin{itemize}
-\item \href{https://github.com/yamadharma/ceurart}{GitHub},
-\item \href{https://www.overleaf.com/project/5e76702c4acae70001d3bc87}{Overleaf},
-\item
-  \href{https://www.overleaf.com/latex/templates/template-for-submissions-to-ceur-workshop-proceedings-ceur-ws-dot-org/hpvjjzhjxzjk}{Overleaf
-    template}.
-\end{itemize}
-
 \end{document}

### 2021-07-08 article.tex
add citation
paper outline

@@
 ]{ceurart}
-\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+\usepackage[colorinlistoftodos,prependcaption]{todonotes}
+\usepackage{xcolor}
 %%
@@
   In particular, Deep Learning techniques and models in the last years outperform hand-craft human solutions in different contexts, ranging from computer vision to videogames.
-  Currently, though, an all-encompassing Eeep learning model does not exist, so researchers explore the possibility to apply these outstanding techniques in different fields.
+  Currently, though, an all-encompassing Deep learning model does not exist, so researchers explore the possibility to apply these outstanding techniques in different fields.
   One of the most challenging is Cyber-Physical System. Here, engineers have to handle large scale, global-to-local behaviour mapping,
@@
 \section{Introduction}
+Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}
+\todo[inline]{walkthrogh Cyber Physical Systems, why they are hard}
+\todo[inline]{Brief description of declarative approach, pros and cons}
+\todo[inline]{Brief description of machine learning approach (MARL and GNN for communication), pros and cons}
+\todo[inline]{Declare our direction: Combination to declarative and automatic approach}
+\todo[inline]{Outline}
 
 \section{Background and Related Work}
+\todo[inline]{walkthrogh traditional engineering way to design CPS}
+\todo[inline]{walkthrogh Aggregate Computing}
+\todo[inline]{walkthrogh Machine learning approach applied in MARL}
+\todo[inline]{current limitation (application specific, not large scala, centralised in some cases,..)}
+\todo[inline]{novel work on GNN -- agent learn to communicate -- node size indipendent~\cite{DBLP:conf/corl/TolstayaGPP0R19}}
+
+\section{Aggregate computing and Machine learning}
+\todo[inline]{description of execution model of AC}
+\todo[inline]{typical structure of AC algorithm : time evolution with hood operation}
+\todo[inline]{show an example of block G?}
+\todo[inline]{declare problems in applying traditional approach (variable-input size, feature hard-coded, environment non-stationarity, reward only know at the end)}
+\todo[inline]{define benefit, possible contrast (i.e. if AC is declarative, why we should use ML?)}
+
+\section{Toward a Neural Network Model}
+\todo[inline]{declare inspiration of GNN} %%ideally, we should use either reinforcement and supervised learning
+\todo[inline]{show the structure: state network, neighbour network, output network}
+\todo[inline]{possible limitations: non stabilisation and asynchrnous evaluation}
+\todo[inline]{learning algorithm: how to apply backprop here (similar to truncated backpropagation, currently does not work)}
+
+\section{Evaluation} %% optional section
+\todo[inline]{define experiment setup}
+\todo[inline]{verify if the system perform as traditional method}
 
 \section{Conclusion and Future Work}
+\todo[inline]{sharping the model}
+\todo[inline]{tune network in order to outperform traditional method}
+\todo[inline]{engineering hybrid application (when we need to use ML)}
+\todo[inline]{toward online learning?}
 %%

### 2021-07-13 paper-2021-woa-ac-ml.tex
Rename main tex file
change abstract a little bit

--- 2021-07-08 article.tex
@@
 %% The "title" command
-\title{Towards a Neural Network model for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
+\title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
 
@@
   Artificial intelligence is here to stay. 
-  In particular, Deep Learning techniques and models in the last years outperform hand-craft human solutions in different contexts, ranging from computer vision to videogames.
+  In particular, Deep Learning techniques in the last years outperform hand-craft human solutions in different contexts, ranging from computer vision to videogames.
   Currently, though, an all-encompassing Deep learning model does not exist, so researchers explore the possibility to apply these outstanding techniques in different fields.
-  One of the most challenging is Cyber-Physical System. Here, engineers have to handle large scale, global-to-local behaviour mapping,
+  One of the most challenging is Collective-Adaptive System (CAS). Here, engineers have to handle large scale, global-to-local behaviour mapping,
   highly environment stochasticity. 
@@
   it needs to build low-level blocks that tend to be tricky. 
-  On the opposite side, some research lines apply deep learning (supervised, reinforcement, non-supervised and self-supervised)
-  along with CPS. Currently, though, no one gain popularity and solution tend to be application-specific.
-  So, we outline a Neural Network model that could be applied in Aggregate Computing, making the possibility to combine declarative and automatic approaches that can easily scale in application complexity.
+  On the opposite side, some research lines apply Deep learning (supervised, reinforcement, non-supervised and self-supervised)
+  along with CAS. Currently, though, no one gain popularity and solution tend to be application-specific.
+  So, we outline a Machine Learning technique that could be applied in Aggregate Computing, making the possibility to combine declarative and automatic approaches that can easily scale in application complexity.
 \end{abstract}
@@
 \maketitle
-
+\todo[inline, color=red]{max pages : 16}
 \section{Introduction}

### 2021-07-16 paper-2021-woa-ac-ml.tex
minors
minors on section outline

@@
 \begin{keywords}
-  Cyber-Physical System \sep
+  Collective-Adaptive System \sep
   Aggregate Computing \sep
@@
 \todo[inline]{walkthrogh Cyber Physical Systems, why they are hard}
-\todo[inline]{Brief description of declarative approach, pros and cons}
-\todo[inline]{Brief description of machine learning approach (MARL and GNN for communication), pros and cons}
+\todo[inline]{Brief description of declarative approaches, pros and cons}
+\todo[inline]{Brief description of machine learning approaches (MARL and GNN for communication), pros and cons}
 \todo[inline]{Declare our direction: Combination to declarative and automatic approach}
@@
 \todo[inline]{walkthrogh Machine learning approach applied in MARL}
-\todo[inline]{current limitation (application specific, not large scala, centralised in some cases,..)}
+\todo[inline]{current limitation (application specific, not large scale, centralised in some cases,..)}
 \todo[inline]{novel work on GNN -- agent learn to communicate -- node size indipendent~\cite{DBLP:conf/corl/TolstayaGPP0R19}}
@@
 
-\section{Toward a Neural Network Model}
+\section{Towards Machine Learning approaches}
 \todo[inline]{declare inspiration of GNN} %%ideally, we should use either reinforcement and supervised learning
@@
 \todo[inline]{learning algorithm: how to apply backprop here (similar to truncated backpropagation, currently does not work)}
-
+\todo[inline]{when makes sense to apply RL? (goal-oriented application?)}
 \section{Evaluation} %% optional section

### 2021-07-20 paper-2021-woa-ac-ml.tex
cps to cas
more on introduction

@@
 Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}
-\todo[inline]{walkthrogh Cyber Physical Systems, why they are hard}
+\todo[inline]{walkthrogh CAS, why they are hard}
+Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs) due to the increasing interconnected computational devices placed in ever-changing environments.
+ Examples of those include robot swarm, smart cities and augmenting crowds.
+A CAS is characterised by 
+\begin{itemize}
+\item a global order that emerges from local node interactions, 
+\item  decentralised control, and 
+\item adaptive behaviours towards environmental changes to pursuit a collective goal.
+\end{itemize}
+CAS are notoriously hard to design due as part of complex systems. 
+%
+So, over the years, in literature, various approaches aim at handling this complexity. 
+%
+Some of them take inspiration from the natural system composed of social animals -- like bees, fish and ants. 
+%
+Here, a sort of hive mind arises -- something referred to as swarm intelligence -- born from local interactions between agents. 
+%
+This kind of behaviour is called self-organisation, namely a global order that emerges from continuous interactions between simple entities.
+
+Initially, researchers try to bring those properties in the computer science world by miming the collective behaviour bottom-up, 
+ namely trying to find the single node behaviour to reach a global behaviour, e.g. flocking, foraging, etc.
+%
+Though, this mapping isn't easily definable -- since we handle complex systems -- therefore novel techniques move the design focus by the ensemble level,
+ specifying the system behaviour as a whole.
+
 \todo[inline]{Brief description of declarative approaches, pros and cons}
@@
 \section{Background and Related Work}
-\todo[inline]{walkthrogh traditional engineering way to design CPS}
+\todo[inline]{walkthrogh traditional engineering way to design CAS}
 \todo[inline]{walkthrogh Aggregate Computing}

### 2021-08-11 paper-2021-woa-ac-ml.tex
Some characteristics of AC in ML-perspective

@@
 \documentclass[
-% twocolumn,
+twocolumn,
 % hf,
@@
 %% One can fix some overfulls
-% \sloppy
+\sloppy
+
+\newcommand{\meta}[1]{{\color{blue}#1}}
 
@@
 
+In this section,
+ we analyse the characteristics of the aggregate computing paradigm,
+ map these to relevant machine learning contributions in literature,
+ and summarise the analysis into a roadmap.
+
+\paragraph{Multi- and many-agent system}
+%
+An aggregate system is a multi-agent system
+ and, often, a \emph{many}-agent system
+ where a large number (hundreds or more)
+ of autonomous entities are programmed to achieve 
+ some collective behaviour by means of \emph{repeated} 
+ sensing, computation, communication, and actuation steps.
+%
+\meta{TODO present approaches for learning in MAS, e.g., through prominent surveys}
+
+\paragraph{Neighbour-based or indirect (environment-mediated) interaction}
+%
+In aggregate systems,
+ a device can only directly interact with neighbours.
+%
+Data flows may be implemented
+ to support indirect communication across multiple hops;
+ however, information from devices far away tends to be obsolete.
+%
+The environment can also be used, via stigmergy,
+ for indirect communication;
+ however, a device can typically only access 
+  its very local portion of the environment.
+%
+In other words, the system state is only \emph{partially observable}.
+%
+\meta{What ML in these kinds of systems?}
+
+\paragraph{Learning goal: collective behaviour}
+%
+The devices of an aggregate 
+ must cooperatively learn the ``aggregate program'',
+ namely the collective behaviour 
+ that achieves a particular \emph{global goal}
+ in a decentralised, resilient way.
+%
+
+\paragraph{Self-organisation, swarm robotics}
+%
+\meta{TODO}
+
+\paragraph{Emergent behaviour}
+%
+The collective behaviour of an aggregate system
+ \emph{emerges} 
+ from a dense network of computations and interactions
+ in an evolving environment.
+%
+
+\paragraph{Delayed, global rewards}
+%
+In the light of the above properties,
+ if we consider a learning approach based on rewards,
+ we must observe that such rewards would be
+ delayed and mostly global---i.e.,
+ it may not be easy to solve the credit-assignment problem
+ to distinguish good from bad individual contributions
+ to the global outcome.
+
+\paragraph{Transient phase}
+%
+When an aggregate behaviour is carried out,
+ we often focus on the stabilised outcomes,
+ tolerating outcomes that are being generated
+ during the transient phase.
+%
+However, intermediate results are often important,
+ and invariants should hold throughout the entire computation.
+
+\paragraph{Dynamic topology}
+%
+If we admit \emph{mobility} and \emph{failure},
+ then we must also consider 
+ the possibility of changes in neighbourhoods
+ and hence the need of dealing with dynamic topologies.
+
+\subsection{Roadmap}
+
 \section{Towards Machine Learning approaches}

### 2021-08-17 paper-2021-woa-ac-ml.tex
more on multi'and many agent

@@
 %
+Due to the high stochasticity of the environment,
+it is almost impossible to know and
+program the optimal behaviour for all agents in advance.
+This results in the need to create intelligent agents
+so that they can learn the optimal behaviour and adapt to environmental changes.
+%
+In recent decades there has been an emerging trend in the use of reinforcement learning
+ in multi-agent settings, as a powerful, robust and adaptive learning paradigm.
+ Progress has been considerable and a wide range of algorithms are now available.
+
 \meta{TODO present approaches for learning in MAS, e.g., through prominent surveys}

### 2021-08-18 paper-2021-woa-ac-ml.tex
more on methods on mas

@@
  Progress has been considerable and a wide range of algorithms are now available.
+% to expand and and references
+In general, there is not one technical solution in the multi-agent context.
+ There are several (even orthogonal) viewpoints on which research has focused.
+ One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning.
+ Others have focused on exploiting coordination between agents. 
+ Others have applied single RL (such as Histeric Q-leraning) to each agent and achieved good results (even if no formal proof exist).
+%
+ The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
 

### 2021-08-20 paper-2021-woa-ac-ml.tex
minor on nash and hysteretic q learning

@@
  There are several (even orthogonal) viewpoints on which research has focused.
- One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning.
+ One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning~\cite{nash-q}.
  Others have focused on exploiting coordination between agents. 
- Others have applied single RL (such as Histeric Q-leraning) to each agent and achieved good results (even if no formal proof exist).
+ Others have applied single RL (such as Hysteretic Q-learning~\cite{hysteretic-q}) to each agent and achieved good results (even if no formal proof exist).
 %

### 2021-10-05 paper-2021-woa-ac-ml.tex
on introduction

@@
 Artificial intelligence is here to stay. 
-  In particular, Deep Learning techniques in the last years outperform hand-craft human solutions in different contexts, ranging from computer vision to videogames.
-  Currently, though, an all-encompassing Deep learning model does not exist, so researchers explore the possibility to apply these outstanding techniques in different fields.
-  One of the most challenging is Collective-Adaptive System (CAS). Here, engineers have to handle large scale, global-to-local behaviour mapping,
+ In particular, Deep Learning techniques in the last years outperform 
+ hand-craft human solutions in different contexts, ranging from computer vision to videogames.
+%
+Currently, though, an all-encompassing Deep learning model does not exist, 
+ so researchers explore the possibility to apply these techniques in different fields.
+
+One of the most challenging is Collective-Adaptive System (CAS). % Or we want to use Cyber-Physical Swarm??
+ Here, engineers have to handle large scale, global-to-local behaviour mapping,
  highly environment stochasticity. 
-  Historically, researchers favour handling them with self-organisation, by which a collective behaviour emerges from continuously local interaction between simple components.
-  A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing. Even though is promising for expressing high-level collective behaviour, 
+%
+Historically, researchers favour handling them with self-organisation,
+ by which a collective behaviour emerges from continuously local interaction between simple components.
+%
+A novel technique to specify self-organising behaviour 
+ in a composable and functional manner is Aggregate Computing. 
+ Even though it is promising for expressing high-level collective behaviour, 
  it needs to build low-level blocks that tend to be tricky.
+
 On the opposite side, some research lines apply Deep learning (supervised, reinforcement, non-supervised and self-supervised)
  along with CAS. Currently, though, no one gain popularity and solution tend to be application-specific.
-  So, we outline a Machine Learning technique that could be applied in Aggregate Computing, making the possibility to combine declarative and automatic approaches that can easily scale in application complexity.
+%
+So, we outline a Machine Learning technique that could be applied in Aggregate Computing, 
+ making the possibility to combine specification-based and automatic approaches 
+ that can easily scale in application complexity and to any system size.
 \end{abstract}
@@
 %% information and builds the first part of the formatted document.
+
+%%%%% LEGENDA FOR NOTES %%%%%
+%% RED: constraints
+%% YELLOW: suggestions
+%% CYAN: todos
 \maketitle
@@
 \section{Introduction}
-Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}
 \todo[inline]{walkthrogh CAS, why they are hard}
-Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs) due to the increasing interconnected computational devices placed in ever-changing environments.
+Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs)
+ due to the increasing interconnected computational devices placed in ever-changing environments.
  Examples of those include robot swarm, smart cities and augmenting crowds.
-A CAS is characterised by 
+
+CASs are characterised by 
 \begin{itemize}
@@
 %
-Some of them take inspiration from the natural system composed of social animals -- like bees, fish and ants. 
+Some of them take inspiration from the natural system
+ composed of social animals --- like bees, fish and ants. 
 %
-Here, a sort of hive mind arises -- something referred to as swarm intelligence -- born from local interactions between agents. 
+Here, a sort of hive mind arises -- 
+ something referred to as swarm intelligence -- 
+born from local interactions between agents. 
 %
-This kind of behaviour is called self-organisation, namely a global order that emerges from continuous interactions between simple entities.
+This kind of behaviour is called self-organisation, 
+ namely a global order that emerges from continuous 
+ interactions between simple entities.
 
-Initially, researchers try to bring those properties in the computer science world by miming the collective behaviour bottom-up, 
- namely trying to find the single node behaviour to reach a global behaviour, e.g. flocking, foraging, etc.
+Initially, researchers try to bring those properties in the computer science world 
+ by miming the collective behaviour bottom-up, 
+ namely trying to find the right single node behaviour 
+ to reach a global behaviour, e.g. flocking, foraging, etc.
 %
-Though, this mapping isn't easily definable -- since we handle complex systems -- therefore novel techniques move the design focus by the ensemble level,
- specifying the system behaviour as a whole.
+Though, this mapping isn't easily definable ---
+ since we handle complex systems.
+%
+Ideally we would like to express collective behaviours focusing only at the ensemble level,
+ abstracting over non-functional aspect such as the network topology, environment uncertetines and 
+ so on.
 
-\todo[inline]{Brief description of declarative approaches, pros and cons}
-\todo[inline]{Brief description of machine learning approaches (MARL and GNN for communication), pros and cons}
-\todo[inline]{Declare our direction: Combination to declarative and automatic approach}
-\todo[inline]{Outline}
+In this direction, Aggregate Computing is an innovative solution by which 
+ developers can express collective self-organising behaviours through a functional program specification.
+%
+Indeed, an aggregate blueprint consist of a manipulation of \textit{computational fields}, a distributed
+ data structure where at each point in the space time is associated the result of a computation
+ perform by a node.
+%
+These manipulation are defined in \textit{Field-calcus} --- the root of Aggregate Computing. 
+ This calculus describe a set of minimal constructurs to express any spatiotemporal computation.
+%
+Thanks to the ``composable'' nature of Aggregate Computing, we find common abstraction 
+ in collective behaviour specification and then we incapsulate it in so-called ``building-blocks''.
+%
+With those, different properties have been proven (such as \textit{eventual consistency}) and various
+ application have been build (such as swarm robotics, crowds engineering and IoT like application).
+%
+Anyway, building block synthesis concerns different stuff related not only to the behaviour itself
+ but also about the "ensamble" specifcations, that is some some are orthogonal to the
+ beheviour itself. For instance, it is quite difficult to make building block
+ that easly scale with the node movements rapidity.
+%
+
+So, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
+ to developers of defining the right collective specifications.
+%
+This lead to an "hibrid" approach where some behaviour part are still express in the of Aggregate Computing
+ API, and another part is distilled through learning.
+
+\todo[inline, color=yellow]{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be usefull in the paper flow}
+\todo[inline, color=cyan]{Outline}
 
@@
 %% Define the bibliography file to be used
+\nocite{*}
 \bibliography{biblio}

### 2021-10-05 paper-2021-woa-ac-ml.tex
change paper structure

@@
   twocolumn,
-% hf,
+  % hf
 ]{ceurart}
@@
 \usepackage{xcolor}
-%%
-%% One can fix some overfulls
 \sloppy
@@
 
-%%
-%% end of the preamble, start of the body of the document source.
 \begin{document}
+\conference{ - X - }
 
-%%
-%% Rights management information.
-%% CC-BY is default license.
-\copyrightyear{2020}
-\copyrightclause{Copyright for this paper by its authors.
-  Use permitted under Creative Commons License Attribution 4.0
-  International (CC BY 4.0).}
-
-%%
-%% This command is for the conference information
-\conference{22nd Workshop "From Objects to Agents" 
-Bologna, Italy, 1-3 September 2021 }
-
-%%
-%% The "title" command
 \title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
 \author[1]{Gianluca Aguzzi}[%
@@
 ]
+
 \author[2]{Roberto Casadei}[%
@@
 ]
+
 \author[2]{Mirko Viroli}[%
@@
 ]
-\address[1]{Alma Mater Studiorum - Università di Bologna,
-  Cesena, Italy}
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
+\address[1]{Alma Mater Studiorum - Università di Bologna, Cesena, Italy}
+%%%%% LEGENDA FOR NOTES %%%%%
+%% RED: constraints
+%% YELLOW: suggestions
+%% CYAN: todos
+%% GREEN: outline
+\newcommand{\constraints}[1]{\todo[inline, color=red]{#1}}
+\newcommand{\suggestions}[1]{\todo[inline, color=yellow]{#1}}
+\newcommand{\todos}[1]{\todo[inline, color=cyan]{#1}}
+\newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
+
 \begin{abstract}
@@
 
-%%%%% LEGENDA FOR NOTES %%%%%
-%% RED: constraints
-%% YELLOW: suggestions
-%% CYAN: todos
 \maketitle
-\todo[inline, color=red]{max pages : 16}
+\suggestions{Perhaps we could change the narration directly towards "reinforcement learning". I think that I cannot continue to explore multi solution..}
 \section{Introduction}
@@
  application have been build (such as swarm robotics, crowds engineering and IoT like application).
-%
+
 Anyway, building block synthesis concerns different stuff related not only to the behaviour itself
- but also about the "ensamble" specifcations, that is some some are orthogonal to the
+ but also about the "ensamble" specifcations, that are orthogonal to the
  beheviour itself. For instance, it is quite difficult to make building block
  that easly scale with the node movements rapidity.
-%
 
@@
 %
-This lead to an "hibrid" approach where some behaviour part are still express in the of Aggregate Computing
+This lead to an "hybrid" approach where some behaviour part are still express in the of Aggregate Computing
  API, and another part is distilled through learning.
+%
+In this paper, as a first step towards a trurly hybrid approach, we investigate how Aggregate Computing can be
+ extended with Reinforcement Learning (and in particolar with Independent Reinforcement Learning) through a cases
+ study where we outline the performance improve of the "hop count" program.
 
-\todo[inline, color=yellow]{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be usefull in the paper flow}
-\todo[inline, color=cyan]{Outline}
+\suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be usefull in the paper flow}
+\todos{Outline}
 
 \section{Background and Related Work}
-\todo[inline]{walkthrogh traditional engineering way to design CAS}
-\todo[inline]{walkthrogh Aggregate Computing}
-\todo[inline]{walkthrogh Machine learning approach applied in MARL}
-\todo[inline]{current limitation (application specific, not large scale, centralised in some cases,..)}
-\todo[inline]{novel work on GNN -- agent learn to communicate -- node size indipendent~\cite{DBLP:conf/corl/TolstayaGPP0R19}}
+\outline{Probably, for a "short" paper, this section can be avoided.}
+\outline{walkthrogh traditional engineering way to design CAS}
+\outline{walkthrogh Aggregate Computing}
+
+\outline{I do not know, perhaps it is better to avoid this }
+\outline{walkthrogh Machine learning approach applied in MARL}
+\outline{current limitation (application specific, not large scale, centralised in some cases,..)}
 
 \section{Aggregate computing and Machine learning}
-\todo[inline]{description of execution model of AC}
-\todo[inline]{typical structure of AC algorithm : time evolution with hood operation}
-\todo[inline]{show an example of block G?}
-\todo[inline]{declare problems in applying traditional approach (variable-input size, feature hard-coded, environment non-stationarity, reward only know at the end)}
-\todo[inline]{define benefit, possible contrast (i.e. if AC is declarative, why we should use ML?)}
 
@@
  map these to relevant machine learning contributions in literature,
+ pointing out different learning settings in which aggregate computing can me integrated,
  and summarise the analysis into a roadmap.
@@
 
-\meta{TODO present approaches for learning in MAS, e.g., through prominent surveys}
+\todos{present approaches for learning in MAS, e.g., through prominent surveys}
 
@@
 %
-\meta{What ML in these kinds of systems?}
-
+\suggestions{ What kind of ML algorithms can be used here?}
 \paragraph{Learning goal: collective behaviour}
@@
 %
-\meta{TODO}
+\suggestions{ You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
 
@@
 
+\subsection{Machine Learning settings}
+\todo[inline, color=green]{"application" perprective: building block, framework, aggregate computing to guide learning}
+\todo[inline, color=green]{learning perspective: homogeneous team learning}
+\todo[inline, color=green]{learning mode perspective: offline through simulation, offline with local reward, offline then online, full-online}
+\todo[inline, color=green]{method perspective: Supervised, reinforcement, multi-agent reinforcement}
 \subsection{Roadmap}
+\todo[inline, color=green]{AC + ML for building block refining -- offline}
+\todo[inline, color=green]{AC + ML for building block refining -- online}
+\todo[inline, color=green]{AC + ML for application refining -- offline/offline}
+\todo[inline, color=green]{AC + ML for building block creation -- offline} %% why not?
+\todo[inline, color=green]{AC + ML for building block creation -- online}
+\todo[inline, color=green]{AC + ML for framework level -- online}
 
-\section{Towards Machine Learning approaches}
-\todo[inline]{declare inspiration of GNN} %%ideally, we should use either reinforcement and supervised learning
-\todo[inline]{show the structure: state network, neighbour network, output network}
-\todo[inline]{possible limitations: non stabilisation and asynchrnous evaluation}
-\todo[inline]{learning algorithm: how to apply backprop here (similar to truncated backpropagation, currently does not work)}
-\todo[inline]{when makes sense to apply RL? (goal-oriented application?)}
+\section{Aggregate Computing with Reinforcement}
+In this section we describe our first result in Aggregate Computing and Machine learning combination,
+ in particular, we briefly describe how nodes compute to follow the global specification,
+ how reinforcement learning can be injected in the computation loop,
+ and what kind of Reinforcement Learning can be used with Aggregate Computing.
+\subsection{Computatational Model}
+\subsection{Reinforcement Learnin integration}
+\subsection{Independent Reinforcement Learning}
+As a first step, we investigate t
 \section{Evaluation} %% optional section
-\todo[inline]{define experiment setup}
+\todo[inline]{define experiment setup, hop count problem}
+
 \todo[inline]{verify if the system perform as traditional method}

### 2021-10-06 paper-2021-woa-ac-ml.tex
more on AC + ML

@@
 \section{Introduction}
-\todo[inline]{walkthrogh CAS, why they are hard}
 Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs)
@@
 
-Initially, researchers try to bring those properties in the computer science world 
+Initially, researchers try to bring those properties into  the computer science world 
  by miming the collective behaviour bottom-up, 
@@
 %
-Ideally we would like to express collective behaviours focusing only at the ensemble level,
- abstracting over non-functional aspect such as the network topology, environment uncertetines and 
+Ideally, we would like to express collective behaviours focusing only at the ensemble level,
+ abstracting over non-functional aspects such as the network topology, environment uncertainties and 
  so on.
@@
 %
-Indeed, an aggregate blueprint consist of a manipulation of \textit{computational fields}, a distributed
- data structure where at each point in the space time is associated the result of a computation
+Indeed, an aggregate blueprint consists of manipulation of \textit{computational fields}, a distributed
+ data structure where at each point in the spacetime is associated the result of a computation
  perform by a node.
 %
-These manipulation are defined in \textit{Field-calcus} --- the root of Aggregate Computing. 
- This calculus describe a set of minimal constructurs to express any spatiotemporal computation.
+These manipulations are defined in \textit{Field-calcus} --- the root of Aggregate Computing. 
+ This calculus describes a set of minimal constructs to express any spatiotemporal computation.
 %
-Thanks to the ``composable'' nature of Aggregate Computing, we find common abstraction 
- in collective behaviour specification and then we incapsulate it in so-called ``building-blocks''.
+Thanks to the ``composable'' nature of Aggregate Computing, we find common abstractions 
+ in collective behaviour specification, and then we encapsulate them in so-called ``building-blocks''.
 %
 With those, different properties have been proven (such as \textit{eventual consistency}) and various
- application have been build (such as swarm robotics, crowds engineering and IoT like application).
+ applications have been build (such as swarm robotics, crowds engineering).
 
-Anyway, building block synthesis concerns different stuff related not only to the behaviour itself
- but also about the "ensamble" specifcations, that are orthogonal to the
- beheviour itself. For instance, it is quite difficult to make building block
- that easly scale with the node movements rapidity.
+Anyway, building block synthesis concerns different stuff related not only to the behaviour itself,
+ but also about the "ensemble" specifications. 
+ For instance, it is quite difficult to craft building blocks
+ that easily adapt w.r.t the node movements rapidity.
 
@@
 %
-This lead to an "hybrid" approach where some behaviour part are still express in the of Aggregate Computing
- API, and another part is distilled through learning.
+This lead to a "hybrid" approach where a part of behaviour is still expressed using common 
+ Aggregate Computing abstractions and another part is distilled through learning.
 %
-In this paper, as a first step towards a trurly hybrid approach, we investigate how Aggregate Computing can be
- extended with Reinforcement Learning (and in particolar with Independent Reinforcement Learning) through a cases
- study where we outline the performance improve of the "hop count" program.
-
+In this paper, as the first steps towards \textit{Hybrid Aggregate Computing}, we investigate how Aggregate Computing can be
+ extended with Reinforcement Learning (and in particular with Independent Reinforcement Learning) through a case
+ study where we outline the performance improvement of the "hop count" program.
+ \suggestions{If we leave the part about "roadmap" and "machine learning similar setting", we need to add some description here.}
 \suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be usefull in the paper flow}
@@
 \outline{walkthrogh Machine learning approach applied in MARL}
-\outline{current limitation (application specific, not large scale, centralised in some cases,..)}
+\outline{current limitation (application-specific, not large scale, centralised in some cases,..)}
 
@@
  map these to relevant machine learning contributions in literature,
- pointing out different learning settings in which aggregate computing can me integrated,
+ pointing out different learning perspective about the integration of Aggregate Computing with Machine Learning,
  and summarise the analysis into a roadmap.
@@
 %
-In recent decades there has been an emerging trend in the use of reinforcement learning
+In recent decades there has been an emerging trend in the use of Reinforcement Learning
  in multi-agent settings, as a powerful, robust and adaptive learning paradigm.
@@
 In general, there is not one technical solution in the multi-agent context.
- There are several (even orthogonal) viewpoints on which research has focused.
+ There are several (even orthogonal) viewpoints on which researchers have focused.
  One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning~\cite{nash-q}.
@@
 %
-\suggestions{ What kind of ML algorithms can be used here?}
+\suggestions{What kind of ML algorithms can be used here?}
 \paragraph{Learning goal: collective behaviour}
@@
 
-\subsection{Machine Learning settings}
-\todo[inline, color=green]{"application" perprective: building block, framework, aggregate computing to guide learning}
+\subsection{Intergration perspectives}
+Here we outline the different perspectives about Hybrid Aggregate Computing with the aim
+ to clarify the current research gap and to help us in the definition of a research roadmap.
+%
+\paragraph{Application} Learning can be useful at different levels in our framework.
+ Surely, what for us is most important is to improve our building-block definition.
+%
+ Indeed, most of our results and applications are based on them.
+ The building block improvement can succeed either by learning a correction factor 
+ or by learning the behaviour entirely.
+
+On the other side, learning can be useful also for framework related stuff,
+ mainly aim at "non-functional" aspects. Trust mechanisms, evaluation frequency,
+ energy consumption manager, package storage are all the things 
+ that currently need to be handle by hand.
+
+\suggestions{Move the definition of team learning and concurrent learning in the previous section}
+\paragraph{Learning problem} As reference to \cite{mas-anc-cooperation}, Multi agent learning can be
+ divided in \textit{concurrent} and \textit{team} learning. In the foster, each agent perform the 
+ learning process and the in the latter, the learning processo is perform only by one agent an then 
+ shared to the collective.
+ Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour -- i.e. each
+ agent perform the same program. Here, a common reward signal exist and evaluate the overall collective behaviour.
+ Aggregate Computing is quite near to this kind of settings.
+ On the other hand, Concurrent Learning, in \cite{csas-and-marl} seems to be usually applied in Collective
+ Adaptive System, so again a field near to Aggregate Computing. In particular, Indipendedet Reinforcement Learning
+ seems to reach good result in this kind of system, even if no thereom exist as the single agent context.
+ Concurrent learning does not presume a global collective behaviour convernce. Indeed using a common reward signal,
+ the collective can reach the same behaviour, as in the team learning settings as done here~\cite{iima2008swarm}.
+\paragraph{Learning mode}
+Aside of what kind of problem we setup, we can use Reinforcement Learning, Supervised Learning.
+ Superivsed learning is quite rare in CSAS -- or in the swarm system in general. Novel work leverage Graph Neural network
+ to learn how agent should communicate, supposing to know the right behaviour (i.e. produced via global vision simulation).
+ In Aggregate Computing, Supervised Learning makes sense if we want to craft new building block by zero. 
+ Indeed, we know eventually the right state of the system, but we do not know how the system can reach it
+
+ Reinforcement Leanring is quite used in CAS, 
+ due its flessibility and thanks to the plethora of apporaches used in this scenarion.
+ In our scenario, Reinforcement Learning can be used both at the building block level and the framework level.
+ For the forster, the simplest scenario can be that agents learn a correction factor to improve a long-term
+ reward signal.
+
+\paragraph{Learning time}
+ Finally, learning can be perform offline, offline and then online, or fully-online.
+ The foster is the most easy scenario. Here indeed, we can leverage simulation to know the right 
+ collective value and then synthesis the right value. Supervised, Reinforcement Learning and even Evolutionary 
+ computing can be used here. offline + online instead can be a mid-level complexity direction.
+ Here, simulation can always leveraged to reach a good behavior, but then at online in can be
+ adapted in front of new environmental situation. We know that it is quite problematic scenario. Reinforcement 
+ Learning is the most suitable mode here --- also the "meta" learning can help.
+ Finally, we can image that learning will be performed fully online. It is a common situation where
+ we do not know the environment and also simulation can be very hard to accomplish. Here, again, Reinforcement learning
+ seems to be the most suitable approach.
+
+\todo[inline, color=green]{"application" pesprective: building block, framework, aggregate computing to guide learning}
 \todo[inline, color=green]{learning perspective: homogeneous team learning}
@@
 \todo[inline, color=green]{method perspective: Supervised, reinforcement, multi-agent reinforcement}
+
 \subsection{Roadmap}
@@
 \section{Aggregate Computing with Reinforcement}
-In this section we describe our first result in Aggregate Computing and Machine learning combination,
+In this section, we describe our first result in Aggregate Computing and Machine learning combination,
  in particular, we briefly describe how nodes compute to follow the global specification,
- how reinforcement learning can be injected in the computation loop,
+ how Reinforcement Learning can be injected in the computation loop,
  and what kind of Reinforcement Learning can be used with Aggregate Computing.
 \subsection{Computatational Model}
-\subsection{Reinforcement Learnin integration}
+\subsection{Reinforcement Learning integration}
 \subsection{Independent Reinforcement Learning}

### 2021-10-06 paper-2021-woa-ac-ml.tex
more on the roadmap

@@
 \subsection{Roadmap}
-\todo[inline, color=green]{AC + ML for building block refining -- offline}
-\todo[inline, color=green]{AC + ML for building block refining -- online}
-\todo[inline, color=green]{AC + ML for application refining -- offline/offline}
-\todo[inline, color=green]{AC + ML for building block creation -- offline} %% why not?
+The path towards a fully integration of Machine Learning and Aggregate Computing is long
+ and probably we do not know the implications that it can bring. Currently, we program to:
+
+\begin{enumerate}
+  \item Enhance building block through learning -- \textbf{offline}: in this first step, we 
+  need to understand what kind of learning mode is most suitable for us. To decrease the complexity,
+  we aim at perform learning \textit{offline} and at the \textit{building block} level.
+  \item Enhance building blocks through learning -- \textbf{online}: here, the most challenging part
+  in bringing the offline results could be in the reward function definition. Indeed, at runtime
+  we cannot use a global view of the system. Novel way in describe a collective reward by Aggregate Computing
+  might be taken in consideration.
+  \item Create building block through learning: some building block currenlty lack of a proper implementation -- 
+  such the the distributed leader election -- so learning can help us also in crating them. 
+  Here, though, the research space increase drastically, so possibility it cannot be easy to bring our 
+  result there.
+  \item Improve framework aspects through learning: when we successufly apply learning in the first two steps, 
+  we can also consider to move learning at the framework level, to manage non-functional aspect. In this way,
+  developers can really focus on the "high level" behaviour aspect.
+\end{enumerate}
 \todo[inline, color=green]{AC + ML for building block creation -- online}

### 2021-10-07 paper-2021-woa-ac-ml.tex
more on RL

@@
 \usepackage{xcolor}
+\usepackage{listings}
 \sloppy
@@
 \begin{abstract}
+\suggestions{}
 Artificial intelligence is here to stay. 
@@
 \outline{walkthrogh Aggregate Computing}
-
-\outline{I do not know, perhaps it is better to avoid this }
+\subsection{Field calcus}
+\subsection{Machine Learning}
+\suggestions{I do not know, perhaps it is better to avoid this }
 \outline{walkthrogh Machine learning approach applied in MARL}
@@
  map these to relevant machine learning contributions in literature,
- pointing out different learning perspective about the integration of Aggregate Computing with Machine Learning,
+ pointing out different learning perspectives about the integration of Aggregate Computing with Machine Learning,
  and summarise the analysis into a roadmap.
@@
  divided in \textit{concurrent} and \textit{team} learning. In the foster, each agent perform the 
- learning process and the in the latter, the learning processo is perform only by one agent an then 
+ learning process and the in the latter, the learning process is performed only by one agent an then 
  shared to the collective.
@@
  and what kind of Reinforcement Learning can be used with Aggregate Computing.
+
 \subsection{Computatational Model}
+An aggregate program execution consists in a continuously node-point execution of a \textit{round}, that is
+ the atomical part of aggregate computing.
+%
+Each node can interact only with-in its neighbourhoods sharing the \textit{export}, namely the result produced by
+ the evaluation of a round.
+
+The main round steps consist in:
+\begin{enumerate}
+  \item execution context creation: where node collect information from the sensors, the neighbourhoods exports and the old
+  program output
+  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. This produce 
+  the export.
+  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
+  \item actuations: giving the export produced, node might be perform some actuation (e.g. motor activations, steering, ...)
+\end{enumerate}
+
+The rounds executiong is completly asynchrouns, so do not exist a global clock to coordinate the aggregate. This makes possible
+ to scala easly with the node number size.
+
+This collective, proactive and periodical execution will eventually bring to the system-wide specification.
+
 \subsection{Reinforcement Learning integration}
+By a local view, a program execution can be see as:
+$$
+f : CONTEXT => OUTPUT
+$$
+Where the context contains:
+\begin{itemize}
+  \item $S(name) => value$: a function from sensor name to the perceived value
+  \item $\phi(ID) => EXPORT$ a function from id to neighbourhoods data
+\end{itemize}
+The computational field $\theta$ contains all output produced by the system.
+
+Speaking by Reinforcement Learning settings, we need to define the state functions
+and the reward functions.
+
+Here the execution need to be distributed, so state will be a function from the local node context:
+$$
+STATE := \Omega(CONTEXT)
+$$
+
+Reward function instead, might be depens on the learning mode. Indeed, if we perform learning offline, 
+it could be defined from the overall computational field:
+$$
+REWARD_{global} := \psi({\theta})
+$$
+If we use a online mode, it need to be defined as a function to the local context:
+$$
+REWARD_{local} := \psi(\phi)
+$$
+
+Finally, actions will be something that influece the current aggregate program output, changing the program evaluation as:
+$$
+f_{rl} = CONTEXT => \lambda(OUTPUT, ACTION)
+$$
+
+As remarked before, the policy $\pi$ need to be distributed so each agent to choose an action perfom:
+$$
+ACTION = \pi(STATE)
+$$
+
 \subsection{Independent Reinforcement Learning}

### 2021-10-07 paper-2021-woa-ac-ml.tex
a grammar pass, a semantic pass is needed

--- 2021-10-06 paper-2021-woa-ac-ml.tex
@@
 %% Options:
-%% twocolumn : Two column layout.
 %% hf: enable header and footer.
@@
 ]{ceurart}
+\usepackage{tcolorbox}
 \usepackage[colorinlistoftodos,prependcaption]{todonotes}
 \usepackage{xcolor}
+\usepackage{subcaption}
+\usepackage{minted}
+\usepackage{listings}
+\usepackage{amsmath}
+\usepackage{systeme}
+
+\setminted{frame=single}
 \sloppy
@@
 \newcommand{\meta}[1]{{\color{blue}#1}}
-
 \begin{document}
@@
 %
-Currently, though, an all-encompassing Deep learning model does not exist, 
+Currently, though, an all-encompassing Deep learning algorithm does not exist, 
  so researchers explore the possibility to apply these techniques in different fields.
 
-One of the most challenging is Collective-Adaptive System (CAS). % Or we want to use Cyber-Physical Swarm??
+One of the most challenging is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
  Here, engineers have to handle large scale, global-to-local behaviour mapping,
@@
 %
-A novel technique to specify self-organising behaviour 
- in a composable and functional manner is Aggregate Computing. 
+A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing. 
  Even though it is promising for expressing high-level collective behaviour, 
@@
 So, we outline a Machine Learning technique that could be applied in Aggregate Computing, 
- making the possibility to combine specification-based and automatic approaches 
- that can easily scale in application complexity and to any system size.
+ making the possibility to combine specification-based and automatic approaches that can easily scale in application complexity and to any system size.
 \end{abstract}
@@
 \maketitle
-\suggestions{Perhaps we could change the narration directly towards "reinforcement learning". I think that I cannot continue to explore multi solution..}
+\suggestions{Perhaps we could change the narration directly towards "reinforcement learning". I think that I cannot continue to explore multiple solutions..}
 \section{Introduction}
@@
 %
-Indeed, an aggregate blueprint consists of manipulation of \textit{computational fields}, a distributed
- data structure where at each point in the spacetime is associated the result of a computation
- perform by a node.
+Indeed, an aggregate blueprint consists of the manipulation of \textit{computational fields}, a distributed
+ data structure where each point in the spacetime is associated with the result of a computation
+ performed by a node.
 %
-These manipulations are defined in \textit{Field-calcus} --- the root of Aggregate Computing. 
+These operations are defined in \textit{Field-calculus} --- the root of Aggregate Computing. 
  This calculus describes a set of minimal constructs to express any spatiotemporal computation.
@@
 Thanks to the ``composable'' nature of Aggregate Computing, we find common abstractions 
- in collective behaviour specification, and then we encapsulate them in so-called ``building-blocks''.
+ in collective behaviour specification, and then we encapsulate them in so-called ``building blocks''.
 %
 With those, different properties have been proven (such as \textit{eventual consistency}) and various
- applications have been build (such as swarm robotics, crowds engineering).
+ applications have been built (such as swarm robotics, crowds engineering).
 
-Anyway, building block synthesis concerns different stuff related not only to the behaviour itself,
- but also about the "ensemble" specifications. 
+Anyway, building block synthesis concerns different stuff related not only to the behaviour itself
+ but also to the "ensemble" specifications. 
  For instance, it is quite difficult to craft building blocks
@@
  Progress has been considerable and a wide range of algorithms are now available.
-% to expand and and references
+% to expand and references
 In general, there is not one technical solution in the multi-agent context.
@@
  however, a device can typically only access 
-  its very local portion of the environment.
+  to a very local portion of the environment.
 %
@@
  delayed and mostly global---i.e.,
- it may not be easy to solve the credit-assignment problem
+ it may not be easy to solve the credit assignment problem
  to distinguish good from bad individual contributions
@@
 \paragraph{Application} Learning can be useful at different levels in our framework.
- Surely, what for us is most important is to improve our building-block definition.
+ Surely, what for us is most important is to improve our building block definition.
 %
  Indeed, most of our results and applications are based on them.
- The building block improvement can succeed either by learning a correction factor 
- or by learning the behaviour entirely.
+ The building block improvement can succeed either by learning a correction factor or by learning the behaviour entirely.
 
@@
  mainly aim at "non-functional" aspects. Trust mechanisms, evaluation frequency,
- energy consumption manager, package storage are all the things 
- that currently need to be handle by hand.
+ energy consumption manager, package storage are all the things that currently need to be handle by hand.
 
 \suggestions{Move the definition of team learning and concurrent learning in the previous section}
-\paragraph{Learning problem} As reference to \cite{mas-anc-cooperation}, Multi agent learning can be
+\paragraph{Learning problem} Multi-agent learning can be
  divided in \textit{concurrent} and \textit{team} learning. In the foster, each agent perform the 
- learning process and the in the latter, the learning processo is perform only by one agent an then 
+ the learning process and the in the latter, the learning process is performed only by one agent an then 
  shared to the collective.
- Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour -- i.e. each
- agent perform the same program. Here, a common reward signal exist and evaluate the overall collective behaviour.
- Aggregate Computing is quite near to this kind of settings.
+ Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour -- i.e. each agent performs the same program. 
+ Here, a common reward signal exists and evaluate the overall collective behaviour.
+ Aggregate Computing is quite near to this kind of setting.
  On the other hand, Concurrent Learning, in \cite{csas-and-marl} seems to be usually applied in Collective
- Adaptive System, so again a field near to Aggregate Computing. In particular, Indipendedet Reinforcement Learning
- seems to reach good result in this kind of system, even if no thereom exist as the single agent context.
- Concurrent learning does not presume a global collective behaviour convernce. Indeed using a common reward signal,
+ Adaptive System, so again a field near to Aggregate Computing. 
+ In particular, Indipendedet Reinforcement Learning seems to reach good results in this kind of system, even if no theorems exist in the single-agent context.
+ Concurrent learning does not presume a global collective behaviour convergence. Indeed using a common reward signal,
  the collective can reach the same behaviour, as in the team learning settings as done here~\cite{iima2008swarm}.
-\paragraph{Learning mode}
-Aside of what kind of problem we setup, we can use Reinforcement Learning, Supervised Learning.
- Superivsed learning is quite rare in CSAS -- or in the swarm system in general. Novel work leverage Graph Neural network
- to learn how agent should communicate, supposing to know the right behaviour (i.e. produced via global vision simulation).
- In Aggregate Computing, Supervised Learning makes sense if we want to craft new building block by zero. 
+
+\paragraph{learning modality}
+Aside from what kind of problem we set up, we can use Reinforcement Learning, Supervised Learning.
+ Supervised learning is quite rare in CSAS -- or in the swarm system in general.
+Novel work leverage Graph Neural network to learn how agents should communicate, supposing to know the right behaviour (i.e. produced via global vision simulation).
+ In Aggregate Computing, Supervised Learning makes sense if we want to craft new building blocks by zero. 
  Indeed, we know eventually the right state of the system, but we do not know how the system can reach it
 
- Reinforcement Leanring is quite used in CAS, 
- due its flessibility and thanks to the plethora of apporaches used in this scenarion.
+ Reinforcement Learning is quite used in CAS, 
+ due to its flexibility and thanks to the plethora of approaches used in this scenario.
  In our scenario, Reinforcement Learning can be used both at the building block level and the framework level.
- For the forster, the simplest scenario can be that agents learn a correction factor to improve a long-term
+ For the first, the simplest scenario can be that agents learn a correction factor to improve a long-term
  reward signal.
@@
 \paragraph{Learning time}
- Finally, learning can be perform offline, offline and then online, or fully-online.
- The foster is the most easy scenario. Here indeed, we can leverage simulation to know the right 
- collective value and then synthesis the right value. Supervised, Reinforcement Learning and even Evolutionary 
- computing can be used here. offline + online instead can be a mid-level complexity direction.
- Here, simulation can always leveraged to reach a good behavior, but then at online in can be
- adapted in front of new environmental situation. We know that it is quite problematic scenario. Reinforcement 
- Learning is the most suitable mode here --- also the "meta" learning can help.
- Finally, we can image that learning will be performed fully online. It is a common situation where
- we do not know the environment and also simulation can be very hard to accomplish. Here, again, Reinforcement learning
- seems to be the most suitable approach.
+ Finally, learning can be performed \texttt{offline}, \texttt{offline and then online}, or \texttt{fully online}.
+ The foster is the easiest scenario.
+ Here indeed, we can leverage simulation to know the right collective value and then synthesis the right value. 
+ Supervised, Reinforcement Learning and even Evolutionary computing can be used here. 
+ offline and then online instead can be a mid-level complexity direction.
+ Here, simulation can always be leveraged to reach a good behaviour, but then online in can be adapted in front of new environmental situations. 
+ We know that it is quite a problematic scenario. 
+ Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.
+ Finally, we can imagine that learning will be performed fully online. 
+ It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
+ Here, again, Reinforcement learning seems to be the most suitable approach.
 
@@
 \todo[inline, color=green]{learning perspective: homogeneous team learning}
-\todo[inline, color=green]{learning mode perspective: offline through simulation, offline with local reward, offline then online, full-online}
+\todo[inline, color=green]{learning modality perspective: offline through simulation, offline with local reward, offline then online, full-online}
 \todo[inline, color=green]{method perspective: Supervised, reinforcement, multi-agent reinforcement}
@@
 \subsection{Roadmap}
-\todo[inline, color=green]{AC + ML for building block refining -- offline}
-\todo[inline, color=green]{AC + ML for building block refining -- online}
-\todo[inline, color=green]{AC + ML for application refining -- offline/offline}
-\todo[inline, color=green]{AC + ML for building block creation -- offline} %% why not?
+The path towards full integration of Machine Learning and Aggregate Computing is long and probably we do not know the implications that it can bring. 
+Currently, we program to:
+\begin{enumerate}
+  \item Enhance building block through learning -- \textbf{offline}: in this first step, we need to understand what kind of learning modality is most suitable for us. 
+  To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
+  \item Enhance building blocks through learning -- \textbf{online}: here, the most challenging part in bringing the offline results could be in the reward function definition. 
+  Indeed, at runtime, we cannot use a global view of the system. 
+  A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
+  \item Create building block through learning: some building blocks currently lack a proper implementation -- 
+  such as the distributed leader election -- so learning can help us also in crafting them. 
+  Here, though, the research space increase drastically, so it is possible it cannot be easy to bring our result there.
+  \item Improve framework aspects through learning: when we successfully apply to learn in the first two steps, 
+  we can also consider moving learning at the framework level, to manage non-functional aspects. In this way,
+  developers can focus on the "high level" behaviour aspect.
+\end{enumerate}
 \todo[inline, color=green]{AC + ML for building block creation -- online}
@@
  and what kind of Reinforcement Learning can be used with Aggregate Computing.
+
 \subsection{Computatational Model}
+An aggregate program execution consists in a continuously node-point execution of a \texttt{round}, that is
+ the atomical part of aggregate computing.
+%
+Each node can interact only with-in its neighbourhoods sharing the \texttt{export}, namely the result produced by
+ the evaluation of a \texttt{round}.
+
+The main round steps consist in:
+\begin{enumerate}
+  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
+  program output
+  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
+  This produces the \texttt{export}.
+  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
+  \item actuations: giving the export produced, nodes may performing some actuation (e.g. motor activations, steering, ...)
+\end{enumerate}
+
+The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. This makes possible
+ to scale easily with the node number size.
+
+This collective, proactive and periodical execution will eventually bring to the system-wide specification.
+
 \subsection{Reinforcement Learning integration}
-\subsection{Independent Reinforcement Learning}
-As a first step, we investigate t
+From a local view, program execution can be seen as:
+$$
+f : CONTEXT => OUTPUT
+$$
+Where the context contains:
+\begin{itemize}
+  \item $S(name) => value$: a function from sensor name to the perceived value
+  \item $\phi(ID) => EXPORT$ a function from id to neighbourhoods data
+\end{itemize}
+The computational field $\theta$ contains all output produced by the system.
+
+Speaking by Reinforcement Learning settings, we need to define the state functions
+and the reward functions.
+
+Here the execution needs to be distributed, so the state will be a function from the local node context:
+$$
+STATE := \Omega(CONTEXT)
+$$
+
+Reward function instead, might be depending on the learning modality. Indeed, if we perform learning offline, 
+it could be defined from the overall computational field:
+$$
+REWARD_{global} := \psi({\theta})
+$$
+If we use an online mode, it needs to be defined as a function to the local context:
+$$
+REWARD_{local} := \psi(\phi)
+$$
+
+Finally, actions will be something that influences the current aggregate program output, changing the program evaluation as:
+$$
+f_{rl} = CONTEXT => \lambda(OUTPUT, ACTION)
+$$
+
+As remarked before, the policy $\pi$ need to be distributed so each agent to choose the action that will be performed:
+$$
+ACTION = \pi(STATE)
+$$
+
+For our first tests, we decide to test a concurrent learning technique by which each agent try to improve our current local policy.
+In these settings, the environment became non-stationary making the learning process quite unstable.
+But, as we observe in learning application in collective self-adaptive systems, practically these solutions led to good results.
+
 \section{Evaluation} %% optional section
-\todo[inline]{define experiment setup, hop count problem}
+In this section, we outline the results of the Reinforcement Learning application in one simple
+ aggregate program: the hop count.
 
-\todo[inline]{verify if the system perform as traditional method}
+Here, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes.
 
-\section{Conclusion and Future Work}
-\todo[inline]{sharping the model}
-\todo[inline]{tune network in order to outperform traditional method}
-\todo[inline]{engineering hybrid application (when we need to use ML)}
-\todo[inline]{toward online learning?}
+A naive implementation of this kind of problem can be describe in Aggregate Computing as:
+\begin{minted}{scala}
+rep(inf) { hopCounts => 
+ mux(source) 
+  { 0 } 
+  { minHood(hopCounts) + 1} 
+}
+\end{minted}
+This solution though suffers from the slow-rising problem. It consists of a slow convergence time when an equilibrium situation is broken. 
+\todos{expand with examples}
+
+So, our first intuition is to increase the speed of convergence through learning. In particular, the previous solution can be rewrite as:
+
+ \begin{minted}{scala}
+rep(inf) { hopCounts => 
+  val currentState = state(hopCounts)
+  val action = localPolicy(state)
+  mux(source) 
+  { 0 } 
+  { minHood(hopCounts) + 1 + action} 
+}
+\end{minted}
+
+Where the actions are discrete values (0, 1 in the simplest scenario) that correct the local result following a policy refined through learning.
+
+The state instead, is built upon previous output and of the neighbourhoods output.
+
+In particular, in this case, it is built from a temporal window in order to encode the hop count increasing speed:
+\begin{minted}{scala}
+  def speed(hopCount) {
+    val min = minHood(hopCounts)
+    val window = recentValues(min, k)
+    min - window.head 
+  }
+\end{minted}
+Then, due to the partially observability of the environment, we mantain a 
+ list of speed, to understand the hop count direction:
+
+ \begin{minted}{scala}
+  def space(hopCount) {
+    val currentSpeed = speed(hopCount)
+    recentValues(currentSpeed, w)
+  }
+\end{minted}
+
+Where k and w are two parameters of this algorithm.
+
+The policy is improved locally, during simulation, using the Q-Learning algorithm. 
+The reward function is then evaluated locally and consist in: 
+\[
+\systeme*{0 : \textit{hop count output is right}, -1 : \textit{otherwise}}
+\]
+This is a common reward function used to maintain the nodes in a stable situation or to
+fasten the time needed to reach a stable situation.
+
+So, the learnin loop, can be defined as:
+\begin{minted}{scala}
+rep(initialState) {
+  oldState =>
+  val output = rep(inf) { hopCounts => 
+   val currentState = state(hopCounts)
+   val action = localPolicy(state)
+   mux(source) 
+    { 0 } 
+    { minHood(hopCounts) + 1 + action} 
+  }
+  val currentReward = reward(output)
+  qLearning(oldState, 
+   state, 
+   action, 
+   reward
+  )
+}
+\end{minted}
+
+To verify if the aggregate can improve the convergence time using reinforcement learning,
+we simulate a situation where the rising problem emerges:
+in the beginning, two source nodes exist in the network, ($\alpha$ and $\beta$).
+At the time $t$, $\beta$ disappears. So the node near to $\beta$ start a slow process
+to eventually reach the distance towards $\alpha$.
+
+We use x node displacement into an irregular grid. Each simulation least y time unit.
+ we run the simulation for 100 episodes.
+ We use also a time-decay policy both for $\epsilon$ and $\alpha$ as:
+$$
+put a formula
+$$
+
+Figure x shows the error for both naive implementation and Reinforcement Learning based implementation as the time changes.
+
+As you can see, at the time 100 using the greedy policy, the collective learning to recover to a stable situation
+faster than the standard naive implementation.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/1}
+      \label{fig:y equals x}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.3\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/50}
+      \label{fig:three sin x}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.3\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/100}
+      \label{fig:five over x}
+  \end{subfigure}
+     \caption{The error evolution compared with naive implementation.}
+\end{figure}
+\todos{add blinking scenarios}
+\todos{add movement scenarios}
+
+\section{Conclusion and Discussion}
+In this article, we briefly overview how Aggregate Computing can be mixed with Machine Learning -- in particular Reinforcement 
+Learning. This leads us to compare Aggregate Computing problems with current Machine Learning solutions in MASs systems.
+We then define a possible roadmap towards full support of a so-called "Hybrid Aggregate Computing", where Machine Learning
+and Aggregate Computing collaborate.
+\todos{discussion about the result}
+\todos{limitations: it does not easily scale with the complex application, it cannot be applied online, it cannot be used in different scenarios}
+\todos{future work: integrate deep learning solution, integrate actor-critic method (team learning)}
 %%

### 2021-10-08 paper-2021-woa-ac-ml.tex
change gitignore
draft for ac + rl combination

@@
 \usepackage{systeme}
+\usepackage{cleveref}
 
@@
 
-\newcommand{\meta}[1]{{\color{blue}#1}}
 \begin{document}
@@
 
-\author[2]{Mirko Viroli}[%
+\author[3]{Mirko Viroli}[%
  orcid=todo,
@@
 \newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
-
+%% Utility commands
+\newcommand{\hybridaggregate}{\textit{Hybrid Aggregate Computing}}
+\newcommand{\round}{\texttt{round}}
+\newcommand{\export}{\texttt{export}}
+%% Minted helpers
+\definecolor{mblue}{rgb}{0.27,0.33,0.53}
 \begin{abstract}
@@
 %
-In this paper, as the first steps towards \textit{Hybrid Aggregate Computing}, we investigate how Aggregate Computing can be
+In this paper, as the first steps towards \hybridaggregate{}, we investigate how Aggregate Computing can be
  extended with Reinforcement Learning (and in particular with Independent Reinforcement Learning) through a case
@@
 \suggestions{If we leave the part about "roadmap" and "machine learning similar setting", we need to add some description here.}
-\suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be usefull in the paper flow}
+\suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be useful in the paper flow}
 \todos{Outline}
@@
 \subsection{Field calcus}
+\todos{I think it can be helpful to introduce fields calculus through ScaFi...}
 \subsection{Machine Learning}
@@
 In general, there is not one technical solution in the multi-agent context.
- There are several (even orthogonal) viewpoints on which researchers have focused.
+ There are several (even orthogonal) viewpoints on which researchers have been focused.
  One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning~\cite{nash-q}.
- Others have focused on exploiting coordination between agents. 
+ Others exploiting coordination between agents. 
  Others have applied single RL (such as Hysteretic Q-learning~\cite{hysteretic-q}) to each agent and achieved good results (even if no formal proof exist).
@@
 %
-In aggregate systems,
- a device can only directly interact with neighbours.
+In aggregate systems, a device can only directly interact with neighbours.
 %
@@
 %
-\suggestions{ You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
+\suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
 
@@
 
-\subsection{Intergration perspectives}
-Here we outline the different perspectives about Hybrid Aggregate Computing with the aim
- to clarify the current research gap and to help us in the definition of a research roadmap.
+\subsection{Integration perspectives}
+Here we outline the different perspectives about \hybridaggregate{} with the aim
+ of clarifying the current research gap and to help us in the definition of a research roadmap.
 %
 \paragraph{Application} Learning can be useful at different levels in our framework.
- Surely, what for us is most important is to improve our building block definition.
+ Surely, what for us is most important is to improve our building blocks definition.
 % 
 Indeed, most of our results and applications are based on them.
- The building block improvement can succeed either by learning a correction factor or by learning the behaviour entirely.
+%
+The building block improvement may consist either of learning a correction factor or by learning the behaviour entirely.
 
 On the other side, learning can be useful also for framework related stuff,
- mainly aim at "non-functional" aspects. Trust mechanisms, evaluation frequency,
- energy consumption manager, package storage are all the things that currently need to be handle by hand.
+ mainly aim at "non-functional" aspects. 
+ Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that currently need to be handle by hand.
 
 \suggestions{Move the definition of team learning and concurrent learning in the previous section}
-\paragraph{Learning problem} Multi-agent learning can be
- divided in \textit{concurrent} and \textit{team} learning. In the foster, each agent perform the 
- the learning process and the in the latter, the learning process is performed only by one agent an then 
- shared to the collective.
- Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour -- i.e. each agent performs the same program. 
+\paragraph{Learning problem} 
+Multi-agent learning can be divided in \textit{Concurrent} and \textit{Team} learning. 
+In the foster, each agent perform the learning process and in the latter,
+ the learning process is performed only by one agent and then shared with the collective.
+% 
+Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
+ --- i.e. each agent performs the same program. 
  Here, a common reward signal exists and evaluate the overall collective behaviour.
  Aggregate Computing is quite near to this kind of setting.
+% 
+Indeed, each agent follows the \textit{same} aggregate program shared within the entire system.
+
 On the other hand, Concurrent Learning, in \cite{csas-and-marl} seems to be usually applied in Collective
- Adaptive System, so again a field near to Aggregate Computing. 
- In particular, Indipendedet Reinforcement Learning seems to reach good results in this kind of system, even if no theorems exist in the single-agent context.
- Concurrent learning does not presume a global collective behaviour convergence. Indeed using a common reward signal,
- the collective can reach the same behaviour, as in the team learning settings as done here~\cite{iima2008swarm}.
+ (Self-)Adaptive System, so again a field near to Aggregate Computing. 
+ In particular, Indipendedet Reinforcement Learning seems to reach good results in this kind of system, 
+ even if no theorems exist as in the single-agent context.
+%
+With Concurrent learning though is still possible to reach a global collective behaviour convergence. 
+ Indeed using a common reward signal, the collective may reach the same behaviour as in the team learning settings as done here~\cite{iima2008swarm}.
 
-\paragraph{learning modality}
-Aside from what kind of problem we set up, we can use Reinforcement Learning, Supervised Learning.
- Supervised learning is quite rare in CSAS -- or in the swarm system in general.
-Novel work leverage Graph Neural network to learn how agents should communicate, supposing to know the right behaviour (i.e. produced via global vision simulation).
+\paragraph{Learning technique}
+Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
+ such as Reinforcement Learning or Supervised Learning.
+%
+Supervised learning is quite rare in CSAS -- or in the ``swarm" like system in general.
+%
+Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, supposing to know the right behaviour (i.e.
+produced via global vision simulation).
+%
 In Aggregate Computing, Supervised Learning makes sense if we want to craft new building blocks by zero. 
- Indeed, we know eventually the right state of the system, but we do not know how the system can reach it
+% 
+Indeed, we know eventually the right state of the system, but we do not know how the system can reach it.
 
- Reinforcement Learning is quite used in CAS, 
+Reinforcement Learning is predominantly used in CAS, 
  due to its flexibility and thanks to the plethora of approaches used in this scenario.
- In our scenario, Reinforcement Learning can be used both at the building block level and the framework level.
- For the first, the simplest scenario can be that agents learn a correction factor to improve a long-term
- reward signal.
+% 
+In our case, Reinforcement Learning can be used both at the building block level and the framework level.
+%
+For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.
 
 \paragraph{Learning time}
- Finally, learning can be performed \texttt{offline}, \texttt{offline and then online}, or \texttt{fully online}.
+Finally, learning can be performed \textit{offline}, \texttt{online} or \textit{offine an then online} (\textit{mixed time}).
+% 
 The foster is the easiest scenario.
- Here indeed, we can leverage simulation to know the right collective value and then synthesis the right value. 
+ Here indeed, we can leverage simulation to know the right computational field.
+% 
 Supervised, Reinforcement Learning and even Evolutionary computing can be used here.
- offline and then online instead can be a mid-level complexity direction.
- Here, simulation can always be leveraged to reach a good behaviour, but then online in can be adapted in front of new environmental situations. 
- We know that it is quite a problematic scenario. 
+%
+Mixed-time learning can be a mid-level complexity direction instead.
+ Here, simulation can always be leveraged to reach a good behaviour, but at runtime, the behaviour can be adapted in front of new environmental situations. 
+%
 Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.
- Finally, we can imagine that learning will be performed fully online. 
+%
+Finally, we can imagine that learning can be performed online. 
  It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
@@
 \begin{enumerate}
-  \item Enhance building block through learning -- \textbf{offline}: in this first step, we need to understand what kind of learning modality is most suitable for us. 
+  \item Enhance building blocks through learning -- \textit{offline}: 
+  in this first step, we need to understand what kind of learning modality is most suitable for us. 
   To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
-  \item Enhance building blocks through learning -- \textbf{online}: here, the most challenging part in bringing the offline results could be in the reward function definition. 
+  \item Enhance building blocks through learning -- \textit{online}:
+  here, the most challenging part in bringing the offline results could be in the reward function definition.
   Indeed, at runtime, we cannot use a global view of the system. 
   A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
-  \item Create building block through learning: some building blocks currently lack a proper implementation -- 
+  \item Create building blocks through learning: some building blocks currently lack a proper implementation -- 
   such as the distributed leader election -- so learning can help us also in crafting them. 
-  Here, though, the research space increase drastically, so it is possible it cannot be easy to bring our result there.
-  \item Improve framework aspects through learning: when we successfully apply to learn in the first two steps, 
-  we can also consider moving learning at the framework level, to manage non-functional aspects. In this way,
-  developers can focus on the "high level" behaviour aspect.
+  Here, though, the research space increase drastically, 
+  so probably it might be not easy to bring our previous results there.
+  \item Improve framework aspects through learning: 
+  when we successfully apply to learn in the first two steps, 
+  we can also consider moving learning at the framework level, to manage non-functional aspects. 
+  In this way, developers can focus on the "high level" behaviour aspect.
 \end{enumerate}
@@
 \section{Aggregate Computing with Reinforcement}
-In this section, we describe our first result in Aggregate Computing and Machine learning combination,
- in particular, we briefly describe how nodes compute to follow the global specification,
+In this section, we describe our first result in Aggregate Computing and Machine learning combination.
+% 
+Hence we briefly describe how nodes compute to follow the global aggregate specification,
  how Reinforcement Learning can be injected in the computation loop,
@@
 \subsection{Computatational Model}
-An aggregate program execution consists in a continuously node-point execution of a \texttt{round}, that is
- the atomical part of aggregate computing.
+An aggregate program execution consists of a continuously node-point execution of a \round{}, 
+ that is the atomical part of aggregate computing.
 %
-Each node can interact only with-in its neighbourhoods sharing the \texttt{export}, namely the result produced by
- the evaluation of a \texttt{round}.
+Each node can interact only with-in its neighbourhoods sharing the \export{}, 
+ namely the result produced by the evaluation of a \round{}.
 
@@
   \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
-  This produces the \texttt{export}.
+  This produces the \export{}.
   \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
@@
 
-The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. This makes possible
- to scale easily with the node number size.
+The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
+ This makes it possible to scale easily with the node number size.
 
-This collective, proactive and periodical execution will eventually bring to the system-wide specification.
+This collective, proactive and periodical execution will eventually reach the system-wide specification.
 
@@
 $$
-f : CONTEXT => OUTPUT
+f : \textit{Context} \rightarrow \textit{Output}
 $$
@@
 \begin{itemize}
-  \item $S(name) => value$: a function from sensor name to the perceived value
-  \item $\phi(ID) => EXPORT$ a function from id to neighbourhoods data
+  \item $\sigma(name) \rightarrow \textit{data}$: a function from sensor name to the perceived value
+  \item $\phi(\textit{ID}) \rightarrow \textit{Export}$: a function from id to neighbourhoods export data
 \end{itemize}
@@
 
-Speaking by Reinforcement Learning settings, we need to define the state functions
-and the reward functions.
-
+Speaking by Reinforcement Learning settings, we need to define the state functions and the reward functions.
+%
 Here the execution needs to be distributed, so the state will be a function from the local node context:
 $$
-STATE := \Omega(CONTEXT)
+K := \Omega(\textit{Context})
 $$
 
-Reward function instead, might be depending on the learning modality. Indeed, if we perform learning offline, 
-it could be defined from the overall computational field:
+Reward function instead, might be depending on the learning technique, time and problem. 
+ Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
 $$
-REWARD_{global} := \psi({\theta})
+\textit{R}_{global} := \psi({\theta})
 $$
-If we use an online mode, it needs to be defined as a function to the local context:
+If we use an online mode, it needs to be defined as a function to the local node context:
 $$
-REWARD_{local} := \psi(\phi)
+\textit{R}_{local} := \psi(\textit{Context})
+$$
+Finally, actions influence the current aggregate program output, changing the program evaluation as:
+$$
+f_{rl} : \textit{Context} \rightarrow \lambda(\textit{Output}, \textit{Action})
+$$
+As remarked before, the policy $\pi$ need to be distributed, so each agent can choose the action that will be performed:
+$$
+\textit{Action} \leftarrow \pi(K)
 $$
 
-Finally, actions will be something that influences the current aggregate program output, changing the program evaluation as:
-$$
-f_{rl} = CONTEXT => \lambda(OUTPUT, ACTION)
-$$
+For our first tests, 
+ we decide to set up a Concurrent learning technique, by which each agent try to improve our current local policy.
+%
+In these settings, the environment became non-stationary making the learning process unstable.
+ But, as we observe in collective self-adaptive systems, practically these solutions led to good results.
 
-As remarked before, the policy $\pi$ need to be distributed so each agent to choose the action that will be performed:
-$$
-ACTION = \pi(STATE)
-$$
-
-For our first tests, we decide to test a concurrent learning technique by which each agent try to improve our current local policy.
-In these settings, the environment became non-stationary making the learning process quite unstable.
-But, as we observe in learning application in collective self-adaptive systems, practically these solutions led to good results.
-
-\section{Evaluation} %% optional section
+\section{Evaluation}
 In this section, we outline the results of the Reinforcement Learning application in one simple
@@
 
-A naive implementation of this kind of problem can be describe in Aggregate Computing as:
+A naive implementation of this kind of problem can be described in Aggregate Computing as:
 \begin{minted}{scala}
-rep(inf) { hopCounts => 
- mux(source) 
+def hopCount(): Int = {
+  rep(Double.PositiveInfinity) { 
+    hopCounts => mux(source) 
     { 0 } 
-  { minHood(hopCounts) + 1} 
+    { minHoodPlus(nbr{hopCounts}) + 1 }
+  }
 }
 \end{minted}
-This solution though suffers from the slow-rising problem. It consists of a slow convergence time when an equilibrium situation is broken. 
+This solution though suffers from the slow-rising problem. 
+ It consists of a slow convergence time when an equilibrium situation is broken. 
 \todos{expand with examples}
 
-So, our first intuition is to increase the speed of convergence through learning. In particular, the previous solution can be rewrite as:
+So, our first intuition is to increase the speed of convergence through learning. 
+ In particular, the previous solution can be rewrite as:
 
  \begin{minted}{scala}
-rep(inf) { hopCounts => 
-  val currentState = state(hopCounts)
+def hopCountRl(k: Int, w: Int): Int = {
+  rep(Double.PositiveInfinity) { 
+    hopCounts => 
+      val currentState = 
+        state(hopCounts, k, w)
       val action = localPolicy(state)
@@
         { 0 } 
-  { minHood(hopCounts) + 1 + action} 
+        { minHood(nbr{ hopCounts }) 
+          + 1 
+          + action
+        } 
+  }
 }
@@
 Where the actions are discrete values (0, 1 in the simplest scenario) that correct the local result following a policy refined through learning.
-
+%
 The state instead, is built upon previous output and of the neighbourhoods output.
-
  In particular, in this case, it is built from a temporal window in order to encode the hop count increasing speed:
 \begin{minted}{scala}
-  def speed(hopCount) {
-    val min = minHood(hopCounts)
+def speed(hopCount: Int, k: Int) {
+  val min = minHood(nbr{ hopCounts })
   val window = recentValues(min, k)
@@
  \begin{minted}{scala}
-  def space(hopCount) {
-    val currentSpeed = speed(hopCount)
+def space(hopCount: Int, k: Int, w: Int) {
+  val currentSpeed = speed(hopCount, k)
   recentValues(currentSpeed, w)
@@
 
-Where k and w are two parameters of this algorithm.
-
-The policy is improved locally, during simulation, using the Q-Learning algorithm. 
-The reward function is then evaluated locally and consist in: 
+Where \textbf{\texttt{k}} and \textbf{\texttt{w}} are two parameters of this algorithm.
+%
+The policy is learnt locally, during simulation, using the Q-Learning algorithm. 
+ The reward function consists in: 
 \[
@@
 This is a common reward function used to maintain the nodes in a stable situation or to
-fasten the time needed to reach a stable situation.
-
-So, the learnin loop, can be defined as:
-\begin{minted}{scala}
-rep(initialState) {
-  oldState =>
-  val output = rep(inf) { hopCounts => 
-   val currentState = state(hopCounts)
-   val action = localPolicy(state)
-   mux(source) 
+ fasten the time needed to reach a stable one.
+%
+So, the learning loop, can be defined as:
+\begin{minted}[obeytabs=true,tabsize=1]{scala}
+val inf = Double.PositiveInfinity
+rep((initState, initPolicy, inf)) {
+  (oldState, policy, hopCount) => 
+      val currentState = 
+        state(hopCount, k, w)
+      val action = policy(state)
+      val output = mux(source) 
         { 0 } 
-    { minHood(hopCounts) + 1 + action} 
+        { minHood(hopCount) 
+          + 1 
+          + action
         } 
-  val currentReward = reward(output)
+      }
+      val currentReward = 
+        reward(output)
+      val improvedPolicy = 
         qLearning(oldState, 
@@
           action, 
-   reward
+          reward,
+          policy
+      (currentState, policy, output)
     )
@@
 
-To verify if the aggregate can improve the convergence time using reinforcement learning,
+To verify if we can improve the convergence time using reinforcement learning,
  we simulate a situation where the rising problem emerges:
@@
  to eventually reach the distance towards $\alpha$.
-
-We use x node displacement into an irregular grid. Each simulation least y time unit.
- we run the simulation for 100 episodes.
- We use also a time-decay policy both for $\epsilon$ and $\alpha$ as:
+%
+We use 80 nodes displaced into an irregular grid 100x20 meters. 
+ Each node has a neighbourhoods range equal to 8 meters. 
+ So, on average, each node has 7 neighbours.
+ The time $t$ is fixed to 35.
+ Each simulation least 120-time unit.
+% 
+We run the simulation for 100 episodes.
+ The Q-learning parameters are:
 $$
-put a formula
+\epsilon = 0.01, 
+\alpha_t = 0.05,
+\gamma = 0.9
 $$
-
-Figure x shows the error for both naive implementation and Reinforcement Learning based implementation as the time changes.
-
+\Cref{fig:simulation} shows the error for both naive implementation and Reinforcement Learning based implementation as time changes.
+The error is evaluated as:
+$$
+error(i, t) \leftarrow  | \textit{rightValue}(i,t) - \textit{output}(i,t) |
+$$
+Where \textbf{i} is the node identifier and \textbf{t} is the time when the error is evaluated.
+%
 As you can see, at the time 100 using the greedy policy, the collective learning to recover to a stable situation
@@
       \includegraphics[width=\textwidth]{img/1}
-      \label{fig:y equals x}
   \end{subfigure}
@@
       \includegraphics[width=\textwidth]{img/50}
-      \label{fig:three sin x}
   \end{subfigure}
@@
       \includegraphics[width=\textwidth]{img/100}
-      \label{fig:five over x}
   \end{subfigure}
-     \caption{The error evolution compared with naive implementation.}
+  \caption{The error evolution compared with naive implementation. The \textit{black} line 
+  describes the Reinforcement Learning solution error. The \textit{yellow} line shows
+  the naive solution error.}
+  \label{fig:simulation}
 \end{figure}
@@
 \section{Conclusion and Discussion}
-In this article, we briefly overview how Aggregate Computing can be mixed with Machine Learning -- in particular Reinforcement 
-Learning. This leads us to compare Aggregate Computing problems with current Machine Learning solutions in MASs systems.
-We then define a possible roadmap towards full support of a so-called "Hybrid Aggregate Computing", where Machine Learning
-and Aggregate Computing collaborate.
+In this article, we briefly overview how Aggregate Computing can be mixed with Machine Learning -- 
+ in particular with Reinforcement Learning. 
+ This leads us to compare Aggregate Computing problems with current Machine Learning solutions in MASs systems.
+%
+We then define a possible roadmap towards full support of a so-called \hybridaggregate{}, 
+ where Machine Learning and Aggregate Computing collaborate.
+%
+To exemplify how this integration can be performed we build a preliminary use case where 
+ learning enhances successfully a standard Aggregate Computing program.
+
 \todos{discussion about the result}

### 2021-10-08 paper-2021-woa-ac-ml.tex
add code repository

@@
 To verify if we can improve the convergence time using reinforcement learning,
- we simulate a situation where the rising problem emerges:
+ we simulate a situation where the rising problem emerges (repository at \url{https://github.com/cric96/scafi-with-reinforcement-learning/tree/independent-q-learning-hop-count}):
  in the beginning, two source nodes exist in the network, ($\alpha$ and $\beta$).

### 2021-11-04 paper-2021-woa-ac-ml.tex
change style according to alpaca requirements

@@
-%% The first command in your LaTeX source must be the \documentclass command.
-%%
-%% Options:
-%% hf: enable header and footer.
-\documentclass[
-  twocolumn,
-  % hf
-]{ceurart}
+\documentclass[conference]{IEEEtran}
+\IEEEoverridecommandlockouts
+\usepackage{cite}
+
 \usepackage{tcolorbox}
@@
 \usepackage{xcolor}
+\usepackage{amsmath,amssymb,amsfonts}
+\usepackage{algorithmic}
+\usepackage{graphicx}
+\usepackage{textcomp}
+\usepackage{minted}
 \usepackage{subcaption}
-\usepackage{minted}
-\usepackage{listings}
-\usepackage{amsmath}
+\usepackage[hidelinks]{hyperref}
+\usepackage[inline]{enumitem}
+\usepackage[T1]{fontenc}
+\usepackage{cleveref}
 \usepackage{systeme}
-\usepackage{cleveref}
+\setminted[scala]{fontsize=\scriptsize,baselinestretch=1,bgcolor=white,frame=single}
 
-\setminted{frame=single}
+\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
+  T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 \sloppy
 
-\begin{document}
-\conference{ - X - }
-
-\title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
-
-\author[1]{Gianluca Aguzzi}[%
- orcid=todo,
- email=gianluca.aguzzi@unibo.it,
- url=https://todo/,
-]
-
-\author[2]{Roberto Casadei}[%
- orcid=todo,
- email=roby.casadei@unibo.it,
- url=todo,
-]
-
-\author[3]{Mirko Viroli}[%
- orcid=todo,
- email=mirko.viroli@unibo.it,
- url=todo,
-]
-\address[1]{Alma Mater Studiorum - Università di Bologna, Cesena, Italy}
+\newcommand{\unibo}{\textsc{Alma Mater Studiorum} -- \textit{Università di Bologna}}
+\newcommand{\disi}{\textit{Computer Science and Engineering - DISI}}
+\newcommand{\scafi}{{ScaFi}}
+\newcommand{\scalaloci}{{ScalaLoci}}
+\newcommand{\scafiloci}{{ScaFiLoci}}
+\newcommand{\scalainline}[1]{\mintinline[fontsize=\small]{scala}{#1}}
 %%%%% LEGENDA FOR NOTES %%%%%
@@
 \newcommand{\export}{\texttt{export}}
+
+\begin{document}
+
+\title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
+\author{\IEEEauthorblockN{
+Gianluca Aguzzi\IEEEauthorrefmark{1},
+Roberto Casadei\IEEEauthorrefmark{1},
+Mirko Viroli\IEEEauthorrefmark{1}
+}\\
+\IEEEauthorblockA{\IEEEauthorrefmark{1}
+\textsc{Alma Mater Studiorum}---Universit\`a di Bologna, Cesena, Italy\\
+Email: \{gianluca.aguzzi, roby.casadei, mirko.viroli\}@unibo.it
+}}
+
 %% Minted helpers
 \definecolor{mblue}{rgb}{0.27,0.33,0.53}
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+%\begin{keywords}
+%  Collective-Adaptive System \sep
+%  Aggregate Computing \sep
+%  Deep Learning \sep
+%  Reinforcement Learning \sep
+%\end{keywords}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+
+\maketitle
 \begin{abstract}
@@
 \end{abstract}
+\begin{IEEEkeywords}
+  Collective-Adaptive System,
+  Aggregate Computing,
+  Deep Learning,
+  Reinforcement Learning
+\end{IEEEkeywords}
 
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\begin{keywords}
-  Collective-Adaptive System \sep
-  Aggregate Computing \sep
-  Deep Learning \sep
-  Reinforcement Learning \sep
-\end{keywords}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-
-\maketitle
 \suggestions{Perhaps we could change the narration directly towards "reinforcement learning". I think that I cannot continue to explore multiple solutions..}
@@
 %% Define the bibliography file to be used
-\nocite{*}
+
+\bibliographystyle{IEEEtran}
 \bibliography{biblio}
-
 %%

### 2021-11-04 paper-2021-woa-ac-ml.tex
Remove generated stuff + fix affiliations
Beginning of intro

@@
 \newcommand{\scalainline}[1]{\mintinline[fontsize=\small]{scala}{#1}}
+
 %%%%% LEGENDA FOR NOTES %%%%%
@@
 \begin{abstract}
-\suggestions{}
 Artificial intelligence is here to stay. 
@@
 \suggestions{Perhaps we could change the narration directly towards "reinforcement learning". I think that I cannot continue to explore multiple solutions..}
+
 \section{Introduction}
-Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs)
- due to the increasing interconnected computational devices placed in ever-changing environments.
- Examples of those include robot swarm, smart cities and augmenting crowds.
 
-CASs are characterised by 
-\begin{itemize}
-\item a global order that emerges from local node interactions, 
-\item  decentralised control, and 
-\item adaptive behaviours towards environmental changes to pursuit a collective goal.
-\end{itemize}
-CAS are notoriously hard to design due as part of complex systems. 
+Recent trends like pervasive and autonomic computing 
+ promote a vision of artificially designed Collective Adaptive Systems (CASs),
+ namely collections of agents coordinating and adapting %as wholes 
+ to pursue joint goals in changing environments.
+% due to the increasing interconnected computational devices placed in ever-changing environments.
+ Examples of CASs include robot swarms, smart cities, and digitally-augmented crowds.
+%
+%CASs are characterised by 
+%\begin{itemize}
+%\item a global order that emerges from local node interactions, 
+%\item  decentralised control, and 
+%\item adaptive behaviours towards environmental changes to pursuit a collective goal.
+%\end{itemize}
+%
+CAS are notoriously hard to design and control 
+ %due to as part of complex systems
+ as globally desired behaviour
+ has to emerge from local behaviours and interactions. 
 %

### 2021-11-04 paper-2021-woa-ac-ml.tex
refactor the abstract, pass on intro

--- 2021-11-04 paper-2021-woa-ac-ml.tex
@@
 \begin{abstract}
-\suggestions{}
 Artificial intelligence is here to stay. 
- In particular, Deep Learning techniques in the last years outperform 
+ In particular, Machine Learning techniques in the last years outperform 
  hand-craft human solutions in different contexts, ranging from computer vision to videogames.
 %
-Currently, though, an all-encompassing Deep learning algorithm does not exist, 
+Currently, though, an all-encompassing Machine learning algorithm does not exist, 
  so researchers explore the possibility to apply these techniques in different fields.
@@
 A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing.
+ In particular, this paradigm enables collective program specification as a whole, 
+ as it is based on a distributed data structure manipulation called \textit{computational fields}.
+ It is successfully used in different scenarios ranging from Swarm Robotics to Smart cities.
+
 Even though it is promising for expressing high-level collective behaviour,
- it needs to build low-level blocks that tend to be tricky.
+ practically it is difficult to make programs work in highly dynamic networks, 
+ where nodes have to deal with continuous changes at the environment side.
+%
+Therefore, with this work, we stand the first step towards integration between Aggregate Computing and Machine Learning,
+ in order to improve adaptivity acting in the low-level specifications but maintaining the high-level program behaviours.
 
-On the opposite side, some research lines apply Deep learning (supervised, reinforcement, non-supervised and self-supervised)
- along with CAS. Currently, though, no one gain popularity and solution tend to be application-specific.
+In the article, we outline issues, opportunities and methodologies of these integrations, presenting a case study
+ where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
 %
-So, we outline a Machine Learning technique that could be applied in Aggregate Computing, 
- making the possibility to combine specification-based and automatic approaches that can easily scale in application complexity and to any system size.
+
 \end{abstract}
@@
 
-CASs are characterised by 
+CAS are notoriously hard~\cite{DBLP:conf/huc/Ferscha15} to design as they are characterised by 
 \begin{itemize}
@@
 \end{itemize} 
-CAS are notoriously hard to design due as part of complex systems. 
 %
@@
 %
-This kind of behaviour is called self-organisation, 
+These kinds of behaviours are accomplished with self-organisation, 
  namely a global order that emerges from continuous 
@@
 
-In this direction, Aggregate Computing is an innovative solution by which 
+In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15} is an innovative solution by which 
  developers can express collective self-organising behaviours through a functional program specification.
 %
-Indeed, an aggregate blueprint consists of the manipulation of \textit{computational fields}, a distributed
+Indeed, an aggregate program consists of the manipulation of \textit{computational fields}, a distributed
  data structure where each point in the spacetime is associated with the result of a computation
@@
 %
-These operations are defined in \textit{Field-calculus} --- the root of Aggregate Computing. 
+These operations are defined in \textit{Field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
  This calculus describes a set of minimal constructs to express any spatiotemporal computation.
@@
 %
-With those, different properties have been proven (such as \textit{eventual consistency}) and various
- applications have been built (such as swarm robotics, crowds engineering).
+With those, different properties have been proven (such as \textit{eventual consistency}~\cite{DBLP:journals/taas/BealVPD17}) and various
+ applications have been built (such as swarm robotics, crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
 
@@
 %
-This lead to a "hybrid" approach where a part of behaviour is still expressed using common 
- Aggregate Computing abstractions and another part is distilled through learning.
+This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
+ Aggregate Computing abstractions and another part is distilled through learning~\cite{research}.
 %
-In this paper, as the first steps towards \hybridaggregate{}, we investigate how Aggregate Computing can be
- extended with Reinforcement Learning (and in particular with Independent Reinforcement Learning) through a case
+In this paper, as the first steps towards a so-called \hybridaggregate{}, we investigate how Aggregate Computing can be
+ extended with Reinforcement Learning through a case
  study where we outline the performance improvement of the "hop count" program.
@@
 \suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be useful in the paper flow}
-\todos{Outline}
 
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
@@
   \label{fig:simulation}
-\end{figure}
-\todos{add blinking scenarios}
-\todos{add movement scenarios}
+\end{figure*}
+%\todos{add blinking scenarios}
+%\todos{add movement scenarios}
 
@@
 %% If your work has an appendix, this is the place to put it.
-\appendix
+%\appendix
 

### 2021-11-04 paper-2021-woa-ac-ml.tex
on background

@@
 \newcommand{\suggestions}[1]{\todo[inline, color=yellow]{#1}}
-\newcommand{\todos}[1]{\todo[inline, color=cyan]{#1}}
+\newcommand{\todos}[1]{\todo[inline, color=cyan]{\textbf{TODO}: #1}}
 \newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
@@
 
-\suggestions{Perhaps we could change the narration directly towards "reinforcement learning". I think that I cannot continue to explore multiple solutions..}
 \section{Introduction}
@@
  study where we outline the performance improvement of the "hop count" program.
-\suggestions{If we leave the part about "roadmap" and "machine learning similar setting", we need to add some description here.}
+\todos{Outline}
 \suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be useful in the paper flow}
@@
 \section{Background and Related Work}
+
+Computer Science is an incredibly evolving research area. 
+ In a few years, the environment had been filled with intercommunicating computational devices, such as smartphones, smart things, and personal computers.
+
+In particular, in literature, we speak of Collective Adaptive Systems
+ when the number is incredibly high, there is no central authority that coordinate nodes,
+ the overall behaviour is reached with local interactions, the systems expose self-* properties,
+ and the ensemble pursuit of collective goals
+ exhibiting swarm intelligence.
+
 \outline{Probably, for a "short" paper, this section can be avoided.}

### 2021-11-04 paper-2021-woa-ac-ml.tex
more on background: move computational model there and enrich the description

@@
 
-In the article, we outline issues, opportunities and methodologies of these integrations, presenting a case study
+In the article, we outline issues, opportunities and methodologies of this integration, presenting a case study
  where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
@@
 %
-Indeed, an aggregate program consists of the manipulation of \textit{computational fields}, a distributed
+Indeed, an aggregate program (i.e. a program written with Aggregate Computing) consists of the manipulation of \textit{computational fields}, a distributed
  data structure where each point in the spacetime is associated with the result of a computation
@@
 %
-These operations are defined in \textit{Field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
+These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
  This calculus describes a set of minimal constructs to express any spatiotemporal computation.
@@
 
+Engineering this kind of system is a thriving area, various approaches have been proposed.
+ First attempts leverage so-called \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
+ usually inspired by nature~\cite{DBLP:journals/swarm/BrambillaFBD13}. 
+%
+Though, a natural way to describe this kind of problem is by using an \textit{collective} view of the system.
+ And it is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, where the behaviours are designed \textit{top-down},
+ and then the local behaviour is generated automatically. 
+% 
+Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
+ a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)
+
+In our works, we are focused on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, global-to-local approach where
+ the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
+
+The latter could be seen as the extension of electromagnetic fields to computer science since it basically consists in a 
+ map from any device to its computational value (i.e. the result of a computation).
+
+The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
+ any for express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
+
+The novelty of Aggregate Computing consists in its composable (functional-like) program definition: using basic constructs
+ we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
+
+Crafting opportunistically these building-blocks is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} (i.e.
+the ensamble behaviour is independent from the underlying network details)
+ and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).
+
+Aggregate Computing is studied in various simulated scenario (thanks to Alchemist, a meta-simulator for pervasive systems), rangin from smart cities to crodws, thaks to programming language and toolchain that support
+ this paradigm, such as ScaFi, FCCP and Protelis.
+
+In the next sections, we show the field calculus operators that are essential to understand the case study presented and then
+ we underline how each node computes in order to follow the aggregate program speficiation 
+\subsection{Field Calculus operators}
+
+\subsection{Computational Model}
+An aggregate program execution consists of a continuously node-point execution of a \round{}, 
+ that is the atomical part of aggregate computing.
+%
+Each node can interact only with-in its neighbourhoods sharing the \export{}, 
+ namely the result produced by the evaluation of a \round{}.
+
+The main round steps consist in:
+\begin{enumerate}
+  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
+  program output
+  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
+  This produces the \export{}.
+  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
+  \item actuations: giving the export produced, nodes may performing some actuation (e.g. motor activations, steering, ...)
+\end{enumerate}
+
+The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
+ This makes it possible to scale easily with the node number size.
+
+This collective, proactive and periodical execution will eventually reach the system-wide specification.
+
 \outline{Probably, for a "short" paper, this section can be avoided.}
@@
 \outline{walkthrogh Aggregate Computing}
-\subsection{Field calcus}
-\todos{I think it can be helpful to introduce fields calculus through ScaFi...}
 \subsection{Machine Learning}
@@
 
-\todo[inline, color=green]{"application" pesprective: building block, framework, aggregate computing to guide learning}
-\todo[inline, color=green]{learning perspective: homogeneous team learning}
-\todo[inline, color=green]{learning modality perspective: offline through simulation, offline with local reward, offline then online, full-online}
-\todo[inline, color=green]{method perspective: Supervised, reinforcement, multi-agent reinforcement}
-
 \subsection{Roadmap}
@@
 % 
-Hence we briefly describe how nodes compute to follow the global aggregate specification,
- how Reinforcement Learning can be injected in the computation loop,
+Hence we briefly describe how Reinforcement Learning can be injected in the computation loop,
  and what kind of Reinforcement Learning can be used with Aggregate Computing.
-
-\subsection{Computatational Model}
-An aggregate program execution consists of a continuously node-point execution of a \round{}, 
- that is the atomical part of aggregate computing.
-%
-Each node can interact only with-in its neighbourhoods sharing the \export{}, 
- namely the result produced by the evaluation of a \round{}.
-
-The main round steps consist in:
-\begin{enumerate}
-  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
-  program output
-  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
-  This produces the \export{}.
-  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
-  \item actuations: giving the export produced, nodes may performing some actuation (e.g. motor activations, steering, ...)
-\end{enumerate}
-
-The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
- This makes it possible to scale easily with the node number size.
-
-This collective, proactive and periodical execution will eventually reach the system-wide specification.
 

### 2021-11-04 paper-2021-woa-ac-ml.tex
add references, remove old todos

@@
 \todos{Outline}
-\suggestions{I do not know if we can avoid it altogether, but also a brief description of machine learning approaches can be useful in the paper flow}
 
@@
 Aggregate Computing is studied in various simulated scenario (thanks to Alchemist, a meta-simulator for pervasive systems), rangin from smart cities to crodws, thaks to programming language and toolchain that support
- this paradigm, such as ScaFi, FCCP and Protelis.
+ this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
 
@@
 \subsection{Field Calculus operators}
+
+\subsection{ScaFi DSL}
 
@@
 This collective, proactive and periodical execution will eventually reach the system-wide specification.
-
-\outline{Probably, for a "short" paper, this section can be avoided.}
-\outline{walkthrogh traditional engineering way to design CAS}
-\outline{walkthrogh Aggregate Computing}
-\subsection{Machine Learning}
-\suggestions{I do not know, perhaps it is better to avoid this }
-\outline{walkthrogh Machine learning approach applied in MARL}
-\outline{current limitation (application-specific, not large scale, centralised in some cases,..)}
 
@@
  The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
-
-\todos{present approaches for learning in MAS, e.g., through prominent surveys}
 
@@
 
-\suggestions{Move the definition of team learning and concurrent learning in the previous section}
 \paragraph{Learning problem} 
@@
 \end{enumerate}
-\todo[inline, color=green]{AC + ML for building block creation -- online}
-\todo[inline, color=green]{AC + ML for framework level -- online}
 

### 2021-11-05 paper-2021-woa-ac-ml.tex
expand integration perspective, more on reinforcement learning integration

@@
 \usepackage{systeme}
-\setminted[scala]{fontsize=\scriptsize,baselinestretch=1,bgcolor=white,frame=single}
+\setminted[scala]{baselinestretch=1,bgcolor=white,frame=single}
 
@@
 %
-In this paper, as the first steps towards a so-called \hybridaggregate{}, we investigate how Aggregate Computing can be
- extended with Reinforcement Learning through a case
- study where we outline the performance improvement of the "hop count" program.
+In this paper, as the first steps towards a so-called \hybridaggregate{},
+ we investigate how Aggregate Computing can be extended with Reinforcement Learning through a case study,
+ where we outline the performance improvement of the "hop count" program.
 \todos{Outline}
@@
 
-Aggregate Computing is studied in various simulated scenario (thanks to Alchemist, a meta-simulator for pervasive systems), rangin from smart cities to crodws, thaks to programming language and toolchain that support
+Aggregate Computing is studied in various simulated scenarios (thanks to Alchemist, a meta-simulator for pervasive systems), ranging from smart cities to crowds, thanks to programming language and toolchain that support
  this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
 
-In the next sections, we show the field calculus operators that are essential to understand the case study presented and then
- we underline how each node computes in order to follow the aggregate program speficiation 
+In the next sections, we show the field calculus operators that are essential to understand the case study presented,
+ and then we underline how each node computes in order to follow the aggregate program specification.
 \subsection{Field Calculus operators}
-
-\subsection{ScaFi DSL}
-
+The main constructs that captures the essential aspects for programming self-organising systems are:
+\begin{itemize}
+  \item \textit{Stateful field evolution} expression \mintinline[escapeinside=||]{scala}{|{\color{teal}rep}|(|$e_1$|) {|($x$) => $e_2$|}} describes a field evolving in time. 
+  $e_1$ is the initial field value and the function $(x) => e_2$ define how the field changes at each execution.
+  \item \textit{Neighbour interaction} expression \mintinline[escapeinside=||]{scala}{|\color{teal}nbr|{|$e$|}} a view
+  of the field values in the surroundings of each device where neighbours are
+  mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
+  For instance, $minHood$ return the minimum value 
+  from the neighbourhoods.
+  \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
+  field into two non-communicating zones hosting isolated subcomputations:
+  $e_1$ where $e_0$ is true, and $e_2$ where $e_0$ is false.
+\end{itemize}
 \subsection{Computational Model}
@@
   \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
-  \item actuations: giving the export produced, nodes may performing some actuation (e.g. motor activations, steering, ...)
+  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...)
 \end{enumerate}
 
-The rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
+Rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
  This makes it possible to scale easily with the node number size.
@@
 
-\paragraph{Multi- and many-agent system}
+\paragraph{Multi-and many-agent system}
 %
@@
 % to expand and references
-In general, there is not one technical solution in the multi-agent context.
- There are several (even orthogonal) viewpoints on which researchers have been focused.
- One line of research (that of game theory) has tried to find techniques that achieve equilibrium such as Nash-Q learning~\cite{nash-q}.
- Others exploiting coordination between agents. 
- Others have applied single RL (such as Hysteretic Q-learning~\cite{hysteretic-q}) to each agent and achieved good results (even if no formal proof exist).
+This field is a conjunction of Game theory and Reinforcement Learning, 
+ and there are several (even orthogonal) viewpoints on which researchers have been focused.
+
+First attempts from Reinforcement Learning viewpoint, 
+ goes towards a so-called \textit{conccurent/independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
+ where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}. 
+Here various approaches have been proposed,
+ such as Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
+ Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value.
+The pro of these approaches is that their complexity does not scale up with the number of agents, 
+ but, unfortunately, the learning process is extremely non-stationary and non-stable.
+ Furthermore, they are heuristic and does not exist any convergence proof to a stationary state (even if practically often convergence to a stable policy).
+
+%
+The game theory viewpoint instead has tried to find techniques that achieve equilibrium such as Nash-equilibrium of Pareto optimality.
+ The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q.
+The problem there is that is used with few agents and scale hardly with the application complexity.
+%
+So, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
+ agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
+ Most promising approaches in this direction are MAPPO~\cite{DBLP:journals/corr/abs-2103-01955}, COMA~\cite{DBLP:journals/corr/FoersterFANW17} and MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17}.
+
 %
@@
 In other words, the system state is only \emph{partially observable}.
+
+In practice, CTDE technique aims to solve these problems, using a \emph{global} view at training time but
+ a \emph{partial view} at execution time.
+
+Furthermore, when the system node size is very large, novel works of Mean-Field Reinforcement Learning~\cite{DBLP:conf/ijcnn/ZhouZX21} are able to
+ handle learning efficiently, reasoning about agent and agent neighbourhoods.
 %
-\suggestions{What kind of ML algorithms can be used here?}
 \paragraph{Learning goal: collective behaviour}
@@
 
-\paragraph{Self-organisation, swarm robotics}
+% \paragraph{Self-organisation, swarm robotics} it is condensed it other perspective, so I think it is better to avoid this paragraph.
 %
-\suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
+% \suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
 
@@
 %
+
+Typically, in CAS, single-agent RL brings to emergent behaviour and is used in various applications~\cite{DBLP:conf/icse/DAngeloGGGNPT19}.
 
@@
 
+A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14}.
+It is used to capture agents contributions to the system's global performance.
 \paragraph{Transient phase}
@@
 %
-If we admit \emph{mobility} and \emph{failure},
+If we admit \emph{mobility} and \emph{failure} (rarely considered in related works, where node population is fixed),
  then we must also consider
@@
 Here we outline the different perspectives about \hybridaggregate{} with the aim
- of clarifying the current research gap and to help us in the definition of a research roadmap.
+ of clarifying the current research.
 %
@@
  mainly aim at "non-functional" aspects. 
- Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that currently need to be handle by hand.
+ Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could need to be handled by hand.
 
 \paragraph{Learning problem} 
-Multi-agent learning can be divided in \textit{Concurrent} and \textit{Team} learning. 
+The learning problem in multi-agent settings can be pose as \textit{Independent} and \textit{Team} learning. 
 In the foster, each agent perform the learning process and in the latter,
@@
 
-On the other hand, Concurrent Learning, in \cite{csas-and-marl} seems to be usually applied in Collective
- (Self-)Adaptive System, so again a field near to Aggregate Computing. 
- In particular, Indipendedet Reinforcement Learning seems to reach good results in this kind of system, 
+On the other hand, Independent Learning is usually applied in Collective
+ (Self-)Adaptive System \cite{csas-and-marl}, so again a field near to Aggregate Computing.
+ In particular, Independent Reinforcement Learning seems to reach good results in this kind of system, 
  even if no theorems exist as in the single-agent context.
@@
 With Concurrent learning though is still possible to reach a global collective behaviour convergence. 
- Indeed using a common reward signal, the collective may reach the same behaviour as in the team learning settings as done here~\cite{iima2008swarm}.
+ Indeed using a common reward signal, the collective may reach the same behaviour \cite{iima2008swarm, nguyen2018swarm}.
 
@@
 Reinforcement Learning is predominantly used in CAS, 
- due to its flexibility and thanks to the plethora of approaches used in this scenario.
+ due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
 % 
@@
 The foster is the easiest scenario.
- Here indeed, we can leverage simulation to know the right computational field.
+ Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
 % 
@@
 %
-Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.
-%
 Finally, we can imagine that learning can be performed online. 
  It is a common situation where we do not know the environment and also simulation can be very hard to accomplish.
- Here, again, Reinforcement learning seems to be the most suitable approach.
 
-\subsection{Roadmap}
-The path towards full integration of Machine Learning and Aggregate Computing is long and probably we do not know the implications that it can bring. 
-Currently, we program to:
-\begin{enumerate}
-  \item Enhance building blocks through learning -- \textit{offline}: 
-  in this first step, we need to understand what kind of learning modality is most suitable for us. 
-  To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
-  \item Enhance building blocks through learning -- \textit{online}:
-  here, the most challenging part in bringing the offline results could be in the reward function definition.
-  Indeed, at runtime, we cannot use a global view of the system. 
-  A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
-  \item Create building blocks through learning: some building blocks currently lack a proper implementation -- 
-  such as the distributed leader election -- so learning can help us also in crafting them. 
-  Here, though, the research space increase drastically, 
-  so probably it might be not easy to bring our previous results there.
-  \item Improve framework aspects through learning: 
-  when we successfully apply to learn in the first two steps, 
-  we can also consider moving learning at the framework level, to manage non-functional aspects. 
-  In this way, developers can focus on the "high level" behaviour aspect.
-\end{enumerate}
+In mixed and online cases, Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.
 
@@
 % 
-Hence we briefly describe how Reinforcement Learning can be injected in the computation loop,
+Hence we briefly describe how Reinforcement Learning can be injected into the computation loop,
  and what kind of Reinforcement Learning can be used with Aggregate Computing.
@@
 For our first tests, 
- we decide to set up a Concurrent learning technique, by which each agent try to improve our current local policy.
+ we decide to set up an Independent learning technique, by which each agent try to improve our current local policy.
+ Our approach is partially inspired by Swarm Q-Learning~\cite{nguyen2018swarm} and leverage Hysteretic Q-Learning~\cite{hysteretic-q} to 
+ handle
+ the non-stationarity in the environment.
 %
-In these settings, the environment became non-stationary making the learning process unstable.
- But, as we observe in collective self-adaptive systems, practically these solutions led to good results.
+In particular, each agent maintains a local Q table ($Q_i(s, a)$, where i is the agent id) following the standard Hysteretic update:
+$$
+Q(s, a) = 
+$$
+At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, we combine the N q table
+ selecting the maximum value for each state-action pair:
+$$
+Q_{max}(s, a) = max_i(Q_i(s, a))
+$$
+Then, in the new episode, each agent starts with the $Q_max$ and try to refine it. 
+ This max combination works because we assume a homogeneous behaviour, where each agent perform
+ in the same way when reaching the same state.
 
@@
 
+\subsection{Roadmap}
+The path towards full integration of Machine Learning and Aggregate Computing is long and probably we do not know the implications that it can bring. 
+Currently, we program to:
+\begin{enumerate}
+  \item Enhance building blocks through learning -- \textit{offline}: 
+  in this first step, we need to understand what kind of learning modality is most suitable for us. 
+  To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
+  \item Enhance building blocks through learning -- \textit{online}:
+  here, the most challenging part in bringing the offline results could be in the reward function definition.
+  Indeed, at runtime, we cannot use a global view of the system. 
+  A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
+  \item Create building blocks through learning: some building blocks currently lack a proper implementation -- 
+  such as the distributed leader election -- so learning can help us also in crafting them. 
+  Here, though, the research space increase drastically, 
+  so probably it might be not easy to bring our previous results there.
+  \item Improve framework aspects through learning: 
+  when we successfully apply to learn in the first two steps, 
+  we can also consider moving learning at the framework level, to manage non-functional aspects. 
+  In this way, developers can focus on the "high level" behaviour aspect.
+\end{enumerate}
+
 \todos{discussion about the result}

### 2021-11-05 paper-2021-woa-ac-ml.tex
add hysteretic equation

@@
 $$
-Q(s, a) = 
+\delta \leftarrow r + \gamma * max_a Q_i(s', a) - Q_i(s_t, a_t)
 $$
+$$
+Q(s_t, a_t) =  \begin{cases} 
+  Q(s_t, a_t) + \alpha * \delta & \mbox{if } delta >= 0 \\ 
+  Q(s_t, a_t) + \beta * \delta & \mbox{otherwise }
+\end{cases}
+$$
+
 At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, we combine the N q table

### 2021-11-06 paper-2021-woa-ac-ml.tex
add figure of aggregate program learning loop, hop count and the overall learning program

@@
 \section{Aggregate Computing with Reinforcement}
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
+  \caption{This figure shows the main steps of reinforcement learning in aggregate computing.
+  Each agent starts with his copy of a q-table. 
+  Then, during the simulation, they refine them q-tables changing q-values according to their experience (different colours mean different values). 
+  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
+  Then, a new episode starts using this $Q_{max}$ as a starting point.
+  }
+  \label{fig:aggregate-q-learning}
+\end{figure}
 In this section, we describe our first result in Aggregate Computing and Machine learning combination.
@@
 
-\subsection{Reinforcement Learning integration}
 From a local view, program execution can be seen as:
@@
 For our first tests, 
- we decide to set up an Independent learning technique, by which each agent try to improve our current local policy.
+ we decide to set up an Independent learning technique, by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}).
  Our approach is partially inspired by Swarm Q-Learning~\cite{nguyen2018swarm} and leverage Hysteretic Q-Learning~\cite{hysteretic-q} to 
@@
 $$
-Then, in the new episode, each agent starts with the $Q_max$ and try to refine it. 
- This max combination works because we assume a homogeneous behaviour, where each agent perform
- in the same way when reaching the same state.
+Then, in the next episode, each agent starts with the $Q_{max}$ and try to refine it.
+ This max combination works because we assume a homogeneous behaviour, where each agent perform in the same way when reaching the same state.
 
 \section{Evaluation}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\linewidth]{img/hop-count.png}
+  \caption{Graphical result of hop count evaluation. The red node is the target zone. The label 
+  associated with each node is the result of the aggregate program evaluated in that zone.}
+  \label{fig:hop-count}
+\end{figure}
 In this section, we outline the results of the Reinforcement Learning application in one simple
- aggregate program: the hop count.
+ aggregate program: the hop count (\Cref{fig:aggregate-q-learning}).
 
@@
  \begin{minted}{scala}
-def space(hopCount: Int, k: Int, w: Int) {
+def state(hopCount: Int, k: Int, w: Int) {
   val currentSpeed = speed(hopCount, k)
@@
 %
-So, the learning loop, can be defined as:
-\begin{minted}[obeytabs=true,tabsize=1]{scala}
-val inf = Double.PositiveInfinity
-rep((initState, initPolicy, inf)) {
-  (oldState, policy, hopCount) => 
-      val currentState = 
-        state(hopCount, k, w)
-      val action = policy(state)
-      val output = mux(source) 
-        { 0 } 
-        { minHood(hopCount) 
-          + 1 
-          + action
-        } 
-      }
-      val currentReward = 
-        reward(output)
-      val improvedPolicy = 
-        qLearning(oldState, 
-          state, 
-          action, 
-          reward,
-          policy
-      (currentState, policy, output)
-    )
-}
-\end{minted}
+The \Cref{fig:aggregate-learning-loop} shows the learning loop in details.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{img/aggregate-program-evaluation.pdf}
+  \caption{todo}
+  \label{fig:aggregate-learning-loop}
+\end{figure}
+%\begin{minted}[obeytabs=true,tabsize=1]{scala}
+%val inf = Double.PositiveInfinity
+%rep((initState, initPolicy, inf)) {
+%  (oldState, policy, hopCount) => 
+%      val currentState = 
+%        state(hopCount, k, w)
+%      val action = policy(state)
+%      val output = mux(source) 
+%        { 0 } 
+%        { minHood(hopCount) 
+%          + 1 
+%          + action
+%        } 
+%      }
+%      val currentReward = 
+%        reward(output)
+%      val improvedPolicy = 
+%        qLearning(oldState, 
+%          state, 
+%          action, 
+%          reward,
+%          policy
+%      (currentState, policy, output)
+%    )
+%}
+%\end{minted}
 

### 2021-11-07 paper-2021-woa-ac-ml.tex
change images, pass on abstract and intro

@@
 
-One of the most challenging is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
+One of the most challenging fields for Machine Learning application is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
  Here, engineers have to handle large scale, global-to-local behaviour mapping,
@@
 
-Even though it is promising for expressing high-level collective behaviour,
+Even though it is promising for expressing high-level collective behaviour system-size independent,
  practically it is difficult to make programs work in highly dynamic networks, 
@@
  due to the increasing interconnected computational devices placed in ever-changing environments.
- Examples of those include robot swarm, smart cities and augmenting crowds.
+ Examples of those include robot swarm, smart cities and augmented crowds with smart devices.
 
@@
 Here, a sort of hive mind arises -- 
- something referred to as swarm intelligence -- 
+ referred also as swarm intelligence -- 
 born from local interactions between agents. 
@@
 Initially, researchers try to bring those properties into the computer science world 
- by miming the collective behaviour bottom-up, 
+ by miming the collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
  namely trying to find the right single node behaviour 
@@
  abstracting over non-functional aspects such as the network topology, environment uncertainties and 
- so on.
+ system size.
 
@@
 These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
- This calculus describes a set of minimal constructs to express any spatiotemporal computation.
-%
-Thanks to the ``composable'' nature of Aggregate Computing, we find common abstractions 
- in collective behaviour specification, and then we encapsulate them in so-called ``building blocks''.
+% This calculus describes a set of minimal constructs to express any spatiotemporal computation.
+
+In Aggregate Computing there are common abstractions in collective behaviour specification. 
+ These are encapsulated in so-called ``building blocks''.
 %
@@
 
-Anyway, building block synthesis concerns different stuff related not only to the behaviour itself
+Though, building block synthesis concerns different stuff related not only to the behaviour itself
  but also to the "ensemble" specifications. 
@@
  that easily adapt w.r.t the node movements rapidity.
+
+Ideally, we would continue to build our application using these building blocks -- so the high-level specification remains the same -- using some
+ under-the-hood mechanisms that adapt the behaviour according to different environmental changes.
 
@@
  Aggregate Computing abstractions and another part is distilled through learning~\cite{research}.
-%
-In this paper, as the first steps towards a so-called \hybridaggregate{},
- we investigate how Aggregate Computing can be extended with Reinforcement Learning through a case study,
+
+In particular, In this paper,
+ we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
  where we outline the performance improvement of the "hop count" program. 
+ Furthermore, we show that the learnt policy scales up to different system sizes.
 \todos{Outline}
@@
 \subsection{Field Calculus operators}
-The main constructs that captures the essential aspects for programming self-organising systems are:
+The main constructs that capture the essential aspects for programming self-organising systems are:
 \begin{itemize}
@@
 \end{itemize}
+
+To a more in depth explanation of these opearatos, please refear to~\cite{DBLP:journals/computer/BealPV15}.
 \subsection{Computational Model}
@@
 
-First attempts from Reinforcement Learning viewpoint, 
+First attempts from RL viewpoint, 
  goes towards a so-called \textit{conccurent/independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
@@
 \subsection{Integration perspectives}
-Here we outline the different perspectives about \hybridaggregate{} with the aim
- of clarifying the current research.
+Here we outline the different perspectives about Aggregate Computing with learning, for
+ of clarifying the current and the future research path.
 %
 \paragraph{Application} Learning can be useful at different levels in our framework.
- Surely, what for us is most important is to improve our building blocks definition.
+ First of all, learning can be applied to the building blocks definition.
 % 
@@
  mainly aim at "non-functional" aspects. 
- Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could need to be handled by hand.
+ Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that currently have been defined by hand.
 
@@
  Here, a common reward signal exists and evaluate the overall collective behaviour.
- Aggregate Computing is quite near to this kind of setting.
 % 
-Indeed, each agent follows the \textit{same} aggregate program shared within the entire system.
+In Aggregate Computing, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
+ be applied seamlessly.
 
 On the other hand, Independent Learning is usually applied in Collective
- (Self-)Adaptive System \cite{csas-and-marl}, so again a field near to Aggregate Computing.
- In particular, Independent Reinforcement Learning seems to reach good results in this kind of system, 
+ (Self-)Adaptive System \cite{csas-and-marl}, a field near to Aggregate Computing.
+ In particular, Independent Reinforcement Learning reach good results in this kind of system, 
  even if no theorems exist as in the single-agent context.
 %
-With Concurrent learning though is still possible to reach a global collective behaviour convergence. 
- Indeed using a common reward signal, the collective may reach the same behaviour \cite{iima2008swarm, nguyen2018swarm}.
+With Independent learning though is still possible to reach a global collective behaviour convergence. 
+ Indeed using a common reward signal, the collective may reach the same behaviour through emergence \cite{iima2008swarm, nguyen2018swarm}.
 
@@
 %
-Supervised learning is quite rare in CSAS -- or in the ``swarm" like system in general.
+Supervised learning is rare in CSAS -- or in the ``swarm" like system in general.
 %
-Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, supposing to know the right behaviour (i.e.
-produced via global vision simulation).
+Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
+ supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).
 %
-In Aggregate Computing, Supervised Learning makes sense if we want to craft new building blocks by zero. 
+In Aggregate Computing, Supervised Learning can be applied when we want to craft new building blocks by zero. 
 % 
@@
 
-Reinforcement Learning is predominantly used in CAS, 
+RL is predominantly used in CAS, 
  due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
 % 
-In our case, Reinforcement Learning can be used both at the building block level and the framework level.
+In our case, RL can be used both at the building block level and the framework level.
 %
@@
 % 
-The foster is the easiest scenario.
- Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
-% 
-Supervised, Reinforcement Learning and even Evolutionary computing can be used here.
-%
+The foster is the common scenario in multi-agent settings.
+ Here indeed, we can leverage simulation to know the right computational field and then use one of the novel CTDE techniques.
+
 Mixed-time learning can be a mid-level complexity direction instead.
  Here, simulation can always be leveraged to reach a good behaviour, but at runtime, the behaviour can be adapted in front of new environmental situations.
-%
+
 Finally, we can imagine that learning can be performed online. 
@@
 
-In mixed and online cases, Reinforcement Learning is the most suitable mode here --- also the "meta" learning can help.
+In mixed and online cases, Reinforcement Learning is the most suitable model, because agents \textit{learn by doing}.
 
@@
   \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
-  \caption{This figure shows the main steps of reinforcement learning in aggregate computing.
+  \caption{This figure shows the main steps of RL in Aggregate Computing.
   Each agent starts with his copy of a q-table. 
@@
 \end{figure}
-In this section, we describe our first result in Aggregate Computing and Machine learning combination.
+As first experiments, we leverage RL to improve building blocks. 
 % 
-Hence we briefly describe how Reinforcement Learning can be injected into the computation loop,
- and what kind of Reinforcement Learning can be used with Aggregate Computing.
+In this section, we briefly describe how Reinforcement Learning can be injected into the computation loop,
+ and what Reinforcement Learning we used with Aggregate Computing.
 
@@
 
-Speaking by Reinforcement Learning settings, we need to define the state functions and the reward functions.
+Speaking by Reinforcement Learning settings, we need to define the state function and the reward function.
 %
-Here the execution needs to be distributed, so the state will be a function from the local node context:
+The execution needs to be distributed, so the state will be a function from the local node context:
 $$
@@
 
-For our first tests, 
- we decide to set up an Independent learning technique, by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}).
+Towards a complete integration between Aggregate Computing and RL, we decide to set up an Independent learning technique, 
+ by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}) inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.
+
 Our approach is partially inspired by Swarm Q-Learning~\cite{nguyen2018swarm} and leverage Hysteretic Q-Learning~\cite{hysteretic-q} to 
- handle
- the non-stationarity in the environment.
+ handle the non-stationarity in the environment.
 %
-In particular, each agent maintains a local Q table ($Q_i(s, a)$, where i is the agent id) following the standard Hysteretic update:
+In particular, each agent maintains a local Q table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
 $$
@@
 
-At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, we combine the N q table
- selecting the maximum value for each state-action pair:
+At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, 
+ we combine the N Q table selecting the maximum value for each state-action pair since we assume a homogeneous behaviour, 
+ where each agent perform in the same way when reaching the same state.:
 $$
@@
 Then, in the next episode, each agent starts with the $Q_{max}$ and try to refine it.
- This max combination works because we assume a homogeneous behaviour, where each agent perform in the same way when reaching the same state.
-
 \section{Evaluation}
-\begin{figure}
+\begin{figure*}[h]
   \centering
-  \includegraphics[width=0.9\linewidth]{img/hop-count.png}
-  \caption{Graphical result of hop count evaluation. The red node is the target zone. The label 
+  \begin{subfigure}[b]{0.32\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/hop-count-1.png}
+      \caption{Stable situation: Two source nodes exists}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.32\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/hop-count-2.png}
+      \caption{Unstable situation: One source disappears}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.32\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/hop-count-3.png}
+      \caption{Stable state reached}
+  \end{subfigure}
+  \caption{Graphical result of hop count evaluation. The node with a red cross is the target zone. The label 
   associated with each node is the result of the aggregate program evaluated in that zone.}
   \label{fig:hop-count}
-\end{figure}
+\end{figure*}
 In this section, we outline the results of the Reinforcement Learning application in one simple
- aggregate program: the hop count (\Cref{fig:aggregate-q-learning}).
+ aggregate program: the hop count.
 
-Here, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes.
+Here, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes ((\Cref{fig:hop-count})).
 
@@
  It consists of a slow convergence time when an equilibrium situation is broken.
-\todos{expand with examples}
-
-So, our first intuition is to increase the speed of convergence through learning. 
+So, our first intuition is to accelerate -- already done algorithmically -- of convergence through learning.
  In particular, the previous solution can be rewrite as:
@@
  \begin{minted}{scala}
-def hopCountRl(k: Int, w: Int): Int = {
+def hopCountRl(k: Int, w: Int): Int = 
+{
   rep(Double.PositiveInfinity) { 
@@
 
-Where the actions are discrete values (0, 1 in the simplest scenario) that correct the local result following a policy refined through learning.
+Where the actions are discrete values (0, 1, 2 in our experiments) that correct the local result following a policy refined through learning.
 %
 The state instead, is built upon previous output and of the neighbourhoods output.
- In particular, in this case, it is built from a temporal window in order to encode the hop count increasing speed:
+ In particular, in this case, it is built from a temporal window in order to encode the hop count velocity:
 \begin{minted}{scala}
-def speed(hopCount: Int, k: Int) {
+def speed(hopCount: Int, k: Int) =
+{
   val min = minHood(nbr{ hopCounts })
@@
 Then, due to the partially observability of the environment, we mantain a 
- list of speed, to understand the hop count direction:
+ list of velocity, to understand the hop count direction:
 
  \begin{minted}{scala}
-def state(hopCount: Int, k: Int, w: Int) {
+def state(hopCount: Int, k: Int, w: Int) =
+{
   val currentSpeed = speed(hopCount, k)
@@
 %
-The policy is learnt locally, during simulation, using the Q-Learning algorithm. 
+The policy is learnt locally, during simulation, using our version of Q-Learning algorithm. 
  The reward function consists in: 
 \[
-\systeme*{0 : \textit{hop count output is right}, -1 : \textit{otherwise}}
+\systeme*{0 : & \mbox{\textit{hop count output is right}}, -1 : & \mbox{\textit{otherwise}}}
 \]
-This is a common reward function used to maintain the nodes in a stable situation or to
+The idea is to maintain the nodes in a stable situation or to
  fasten the time needed to reach a stable one.
 %
-The \Cref{fig:aggregate-learning-loop} shows the learning loop in details.
+\Cref{fig:aggregate-learning-loop} shows the learning loop in details.
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{img/aggregate-program-evaluation.pdf}
-  \caption{todo}
+  \caption{Each agent, starting from their context, 
+  performs periodically the standard hop count. 
+  Using the output produced and the history of state and output, 
+  each agent computes its current state. 
+  Using it, they evaluate the action to apply to speed up the hop count convergence following a policy. 
+  This new refined output is then used to evaluate the reward and to refine the q table (dotted line path)}
   \label{fig:aggregate-learning-loop}
@@
 
-To verify if we can improve the convergence time using reinforcement learning,
- we simulate a situation where the rising problem emerges (repository at \url{https://github.com/cric96/scafi-with-reinforcement-learning/tree/independent-q-learning-hop-count}):
- in the beginning, two source nodes exist in the network, ($\alpha$ and $\beta$).
- At the time $t$, $\beta$ disappears. So the node near to $\beta$ start a slow process
- to eventually reach the distance towards $\alpha$.
+\subsection{Simulation setup}
+We verify our learning algorithm through simulation~\footnote{repository at: \url{https://github.com/cric96/alpaca-2022-ac-rl}}.
+ To check if the aggregate eventually speeds up the hop count convergence time, we create a situation where the rising problem emerges (\Cref{fig:hop-count}):
+ in the beginning, two source nodes exist in the network, ($A$ and $B$).
+ At the time $t$, $B$ disappears. So the nodes near $B$ start a slow process
+ to eventually reach the distance towards $A$.
 %
-We use 80 nodes displaced into an irregular grid 100x20 meters. 
- Each node has a neighbourhoods range equal to 8 meters. 
- So, on average, each node has 7 neighbours.
- The time $t$ is fixed to 35.
- Each simulation least 120-time unit.
+We use 80 ($N$) nodes displaced into an irregular grid with a width of 100 meters and a height of 20 meters. 
+ Each node has a neighbourhoods range equal to 8 meters --- on average, each node has 7 neighbours. 
+ The time $t$ is fixed to 20.
+ Each simulation episode least 120-time unit.
 % 
-We run the simulation for 100 episodes.
+We training the aggregate for 500 episodes.
  The Q-learning parameters are:
 $$ 
-\epsilon = 0.01, 
-\alpha_t = 0.05,
+\alpha = 0.5,
+\beta = 0.05,
 \gamma = 0.9
 $$
-\Cref{fig:simulation} shows the error for both naive implementation and Reinforcement Learning based implementation as time changes.
+Furthermore, we use a $\epsilon$ decay policy for the $\epsilon$-greedy policy as:
+\begin{equation*}
+  \begin{array}{l}
+    \epsilon_t = \epsilon_{start} * e^{-t/\theta}\\ 
+    \epsilon_{start }= 0.99, \theta=4000
+  \end{array}
+\end{equation*}
+In doing so, we what that agent to explore as more as possible the state-action space.
+
+\Cref{fig:simulation}-a shows the error for both naive implementation and Reinforcement Learning based implementation as time changes.
 The error is evaluated as:
-$$
-error(i, t) \leftarrow  | \textit{rightValue}(i,t) - \textit{output}(i,t) |
-$$
+\begin{equation*}
+  \begin{array}{l}
+    error(i, t) \leftarrow  | \textit{rightValue}(i,t) - \textit{output}(i,t) | \\
+    total_{error} = \frac{1}{T} * \sum_{i = 0}^T \frac{1}{N} * \sum_{j = 0}^N = error(i, j)
+  \end{array}
+\end{equation*}
+
 Where \textbf{i} is the node identifier and \textbf{t} is the time when the error is evaluated.
 %
-As you can see, at the time 100 using the greedy policy, the collective learning to recover to a stable situation
- faster than the standard naive implementation.
+
+Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b,c,d). To do that, we share the Q-table learn to the 
+entire aggregate and we test the program changing the system size (80, 360 and 760) nodes.
+For each population size, we run 50 episodes and we average the error at each time step. As results, 
+ the aggregate speed-up convergence in each new situation.
 
 \begin{figure*}
+  \begin{subfigure}[b]{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{img/mean-error}
+    \caption{The \textit{orange} line 
+    describes the Reinforcement Learning solution error. The \textit{blue} line shows
+    the naive solution error.}
+    \hspace{5mm}
+  \end{subfigure}
+
   \centering
@@
       \centering
-      \includegraphics[width=\textwidth]{img/1}
+      \includegraphics[width=\textwidth]{img/80}
+      \caption{}
+      \label{fig:simulation-b}
   \end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{img/50}
+      \includegraphics[width=\textwidth]{img/360}
+      \caption{}
+      \label{fig:simulation-c}
   \end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{img/100}
+      \includegraphics[width=\textwidth]{img/760}
+      \caption{}
+      \label{fig:simulation-d}
   \end{subfigure}
-  \caption{The error evolution compared with naive implementation. The \textit{black} line 
-  describes the Reinforcement Learning solution error. The \textit{yellow} line shows
-  the naive solution error.}
+  \caption{ (a) shows the $total_{error}$ evolution during the learning process. (b), (c), and (d) shows the average error in three different simulations with 80, 360 and 760 nodes. 
+  The black line is the error done by RL based solution, 
+  the green line is the error produced by standard hop count implementation}
   \label{fig:simulation}

### 2021-11-07 paper-2021-woa-ac-ml.tex
refinements to whole document: grammar and content

@@
   T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\sloppy
 
@@
 \newcommand{\export}{\texttt{export}}
-
+\sloppy
 \begin{document}
@@
 One of the most challenging fields for Machine Learning application is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
- Here, engineers have to handle large scale, global-to-local behaviour mapping,
+ Here, engineers have to handle a massive number of nodes, global-to-local behaviour mapping, and
  highly environment stochasticity.
-%
+
 Historically, researchers favour handling them with self-organisation,
  by which a collective behaviour emerges from continuously local interaction between simple components.
-%
+
 A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing.
@@
  where nodes have to deal with continuous changes at the environment side.
-%
+
 Therefore, with this work, we stand the first step towards integration between Aggregate Computing and Machine Learning,
@@
  due to the increasing interconnected computational devices placed in ever-changing environments.
- Examples of those include robot swarm, smart cities and augmented crowds with smart devices.
+ Examples of those include robot swarms, smart cities and augmented crowds with smart devices.
 
@@
 %
-So, over the years, in literature, various approaches aim at handling this complexity. 
+So, over the years, in literature, various approaches have tried to handle this complexity. 
 %
 Some of them take inspiration from the natural system
- composed of social animals --- like bees, fish and ants. 
+ composed of social animals --- like bees, fish, and ants. 
 %
 Here, a sort of hive mind arises -- 
- referred also as swarm intelligence -- 
+ also referred to as swarm intelligence --  
  born from local interactions between agents. 
@@
 
-Initially, researchers try to bring those properties into the computer science world 
- by miming the collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
- namely trying to find the right single node behaviour 
+Initially, researchers had tried to migrate those properties into the computer science world 
+ by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
+ namely trying to find the right single-node-behaviour 
  to reach a global behaviour, e.g. flocking, foraging, etc.
@@
 Though, this mapping isn't easily definable ---
- since we handle complex systems.
+ since we are coping with complex systems.
 %
-Ideally, we would like to express collective behaviours focusing only at the ensemble level,
+Ideally, we would like to express collective behaviours at the ensemble level,
  abstracting over non-functional aspects such as the network topology, environment uncertainties and 
@@
 %
-These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
+%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
 % This calculus describes a set of minimal constructs to express any spatiotemporal computation.
 
-In Aggregate Computing there are common abstractions in collective behaviour specification. 
- These are encapsulated in so-called ``building blocks''.
-%
-With those, different properties have been proven (such as \textit{eventual consistency}~\cite{DBLP:journals/taas/BealVPD17}) and various
- applications have been built (such as swarm robotics, crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
+Aggregate Computing captures common abstractions for the collective behaviour specification. 
+ These abstractions are encapsulated in the so-called ``building blocks'' through which those various
+ applications have been built 
+ (such as swarm robotics, crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
 
-Though, building block synthesis concerns different stuff related not only to the behaviour itself
+Though, building block synthesis concerns different stuff not only related to the behaviour itself
  but also to the "ensemble" specifications. 
@@
 
-Ideally, we would continue to build our application using these building blocks -- so the high-level specification remains the same -- using some
+Ideally, we would like to continue building our application using these building blocks -- so the high-level specification remains the same -- but using some
  under-the-hood mechanisms that adapt the behaviour according to different environmental changes.
@@
  we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
- where we outline the performance improvement of the "hop count" program. 
+ where we outline the performance improvement of the ``hop count" program --- 
+ i.e. an aggregate program that produces a field that stores the distance from a source zone. 
  Furthermore, we show that the learnt policy scales up to different system sizes.
@@
 
-In particular, in literature, we speak of Collective Adaptive Systems
- when the number is incredibly high, there is no central authority that coordinate nodes,
- the overall behaviour is reached with local interactions, the systems expose self-* properties,
- and the ensemble pursuit of collective goals
- exhibiting swarm intelligence.
+In particular, in literature, we speak about Collective Adaptive Systems~\cite{DBLP:conf/huc/Ferscha15}
+ when the number of nodes participating is incredibly high, there is no central authority that coordinate nodes,
+ the overall behaviour is reached with local interactions, and the systems expose self-* properties.
 
-Engineering this kind of system is a thriving area, various approaches have been proposed.
- First attempts leverage so-called \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
+Engineering this kind of system is a thriving area, and so various approaches have been proposed.
+ First attempts leverage \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
  usually inspired by nature~\cite{DBLP:journals/swarm/BrambillaFBD13}. 
 %
-Though, a natural way to describe this kind of problem is by using an \textit{collective} view of the system.
- And it is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, where the behaviours are designed \textit{top-down},
- and then the local behaviour is generated automatically. 
+Though, a natural way to describe this kind of problem is by using a \textit{collective} view of the system.
+ This is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
+ where the behaviours are designed \textit{top-down}, and then the local behaviour is generated automatically from this ``system-wide'' specification. 
 % 
@@
 
-In our works, we are focused on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, global-to-local approach where
+In our works, we are focused on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
  the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
 
-The latter could be seen as the extension of electromagnetic fields to computer science since it basically consists in a 
- map from any device to its computational value (i.e. the result of a computation).
+The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
+ map from any devices to their computational value (i.e. the result of a computation).
 
 The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
- any for express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
+ to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
 
-The novelty of Aggregate Computing consists in its composable (functional-like) program definition: using basic constructs
- we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
+The novelty of Aggregate Computing consists in its composable (functional-like) system-size independent program definition: 
+using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
 
-Crafting opportunistically these building-blocks is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} (i.e.
-the ensamble behaviour is independent from the underlying network details)
+By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} 
+ (i.e. the ensemble behaviour is independent of the underlying network details)
  and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).
 
-Aggregate Computing is studied in various simulated scenarios (thanks to Alchemist, a meta-simulator for pervasive systems), ranging from smart cities to crowds, thanks to programming language and toolchain that support
+Aggregate Computing is studied in various simulated scenarios ranging from smart cities to crowds, 
+ thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
  this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
@@
   mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
-  For instance, $minHood$ return the minimum value 
-  from the neighbourhoods.
+  For instance, \mintinline{scala}{minHood} return the minimum value 
+  within the neighbourhood.
   \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
@@
 
-To a more in depth explanation of these opearatos, please refear to~\cite{DBLP:journals/computer/BealPV15}.
+For a more in-depth explanation of these operators, please refer to~\cite{DBLP:journals/computer/BealPV15}.
 \subsection{Computational Model}
@@
 
-The main round steps consist in:
+The main round steps consist of:
 \begin{enumerate}
-  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
-  program output
+  \item execution context creation: nodes create the execution context collecting information from the sensors, 
+  the neighbourhood messages, and the old program output;
   \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
-  This produces the \export{}.
-  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
-  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...)
+  This produces the \export{};
+  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes;
+  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...).
 \end{enumerate}
@@
  map these to relevant machine learning contributions in literature,
- pointing out different learning perspectives about the integration of Aggregate Computing with Machine Learning,
- and summarise the analysis into a roadmap.
+ and pointing out different learning perspectives about the integration of Aggregate Computing with Machine Learning
 
@@
  of autonomous entities are programmed to achieve 
- some collective behaviour by means of \emph{repeated} 
+ some collective behaviour through \emph{repeated} 
  sensing, computation, communication, and actuation steps.
@@
  program the optimal behaviour for all agents in advance.
- This results in the need to create intelligent agents
+ This uncertainty results in the need of creating intelligent agents
  so that they can learn the optimal behaviour and adapt to environmental changes.
@@
 In recent decades there has been an emerging trend in the use of Reinforcement Learning 
- in multi-agent settings, as a powerful, robust and adaptive learning paradigm.
+ in multi-agent settings (called Multi-Agent Reinforcement Learning), as a powerful, robust and adaptive learning paradigm.
  Progress has been considerable and a wide range of algorithms are now available.
-% to expand and references
-This field is a conjunction of Game theory and Reinforcement Learning, 
+
+Multi-Agent Reinforcement Learning is a conjunction of Game theory and Reinforcement Learning, 
  and there are several (even orthogonal) viewpoints on which researchers have been focused.
@@
 First attempts from RL viewpoint, 
- goes towards a so-called \textit{conccurent/independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
+ goes towards a so-called \textit{independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
  where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}.
-Here various approaches have been proposed,
- such as Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
- Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value.
+
+Examples of these are Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
+ Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value~\cite{DBLP:conf/atal/KaisersT10}.
+%
 The pro of these approaches is that their complexity does not scale up with the number of agents, 
- but, unfortunately, the learning process is extremely non-stationary and non-stable.
- Furthermore, they are heuristic and does not exist any convergence proof to a stationary state (even if practically often convergence to a stable policy).
+ but, unfortunately, the learning process is extremely non-stationary and unstable.
+% 
+Furthermore, they are heuristic and do not exist any convergence proof (even if, in practice, they often reach a good policy).
 
+The game theory viewpoint instead has tried to find techniques that achieve equilibrium conditions such as Nash-equilibrium of Pareto optimality.
+ The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q~\cite{DBLP:conf/icml/Littman94}.
+The problem is that it is used with few agents and scale hardly with the application complexity.
 %
-The game theory viewpoint instead has tried to find techniques that achieve equilibrium such as Nash-equilibrium of Pareto optimality.
- The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q.
-The problem there is that is used with few agents and scale hardly with the application complexity.
-%
-So, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
+Finally, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
  agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
@@
 
-%
 The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
+ In Aggregate Computing instead, we need to find a technique that does not depend on the system node count.
 
@@
 %
-In aggregate systems, a device can only directly interact with neighbours.
+In aggregate systems, a device can directly interact only with neighbours.
 %
@@
 %
-The environment can also be used, via stigmergy,
+The environment can also be used, via stigmergy~\cite{DBLP:journals/scp/NicolaSI20},
  for indirect communication;
@@
  we often focus on the stabilised outcomes,
- tolerating outcomes that are being generated
+ tolerating outcomes that are generated
  during the transient phase.
@@
 \subsection{Integration perspectives}
-Here we outline the different perspectives about Aggregate Computing with learning, for
- of clarifying the current and the future research path.
+Here we outline the different perspectives about Aggregate Computing with learning, 
+ to clarify the current and the future research path.
 %
@@
 %
-The building block improvement may consist either of learning a correction factor or by learning the behaviour entirely.
+The building block improvement may consist either of learning a correction factor or by entirely learning the behaviour.
 
 On the other side, learning can be useful also for framework related stuff,
- mainly aim at "non-functional" aspects. 
- Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that currently have been defined by hand.
+ mainly aiming at "non-functional" aspects. 
+ Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could take advantage of using Machine Learning algorithms.
 
 \paragraph{Learning problem} 
-The learning problem in multi-agent settings can be pose as \textit{Independent} and \textit{Team} learning. 
-In the foster, each agent perform the learning process and in the latter,
- the learning process is performed only by one agent and then shared with the collective.
+The learning problem in multi-agent settings can be posed as \textit{Independent} and \textit{Team} learning. 
+In the foster, each agent perform the learning process and while in the latter,
+ the learning process is performed only by one agent and then shared with the collective (in a typical CDTE setting).
 % 
@@
  (Self-)Adaptive System \cite{csas-and-marl}, a field near to Aggregate Computing.
+
 In particular, Independent Reinforcement Learning reach good results in this kind of system, 
- even if no theorems exist as in the single-agent context.
+ even if no theorem exists as in the single-agent context.
 %
@@
  supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).
+
+In Aggregate Computing, Supervised Learning can be applied when we want to build up new building blocks from zero. 
 % 
-In Aggregate Computing, Supervised Learning can be applied when we want to craft new building blocks by zero. 
-% 
-Indeed, we know eventually the right state of the system, but we do not know how the system can reach it.
+Indeed, we eventually know the right state of the system, but we do not know how the system can reach it.
 
@@
  due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
-% 
+
 In our case, RL can be used both at the building block level and the framework level.
@@
 \paragraph{Learning time}
-Finally, learning can be performed \textit{offline}, \texttt{online} or \textit{offine an then online} (\textit{mixed time}).
+Finally, learning can be performed \textit{offline}, \textit{online} or \textit{offline an then online} (\textit{mixed time}).
 % 
@@
 Mixed-time learning can be a mid-level complexity direction instead.
- Here, simulation can always be leveraged to reach a good behaviour, but at runtime, the behaviour can be adapted in front of new environmental situations.
+ Here, simulation can always be exploited to reach a good behaviour, but at runtime, 
+ the behaviour can be adapted based on the new environmental situations.
 
@@
   \caption{This figure shows the main steps of RL in Aggregate Computing.
-  Each agent starts with his copy of a q-table. 
-  Then, during the simulation, they refine them q-tables changing q-values according to their experience (different colours mean different values). 
+  Each agent starts with his copy of a Q table. 
+  Then, during the simulation, they refine them Q tables changing q values according to their experience (different colours mean different values). 
   In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
-  Then, a new episode starts using this $Q_{max}$ as a starting point.
+  Finally, a new episode starts using this $Q_{max}$ as a starting Q table.
   }
@@
 $$
-Where the context contains:
+Where $\textit{Context}$ contains:
 \begin{itemize}
-  \item $\sigma(name) \rightarrow \textit{data}$: a function from sensor name to the perceived value
-  \item $\phi(\textit{ID}) \rightarrow \textit{Export}$: a function from id to neighbourhoods export data
+  \item $\sigma(name) \rightarrow \textit{data}$: a function from sensor name to the perceived value;
+  \item $\phi(\textit{ID}) \rightarrow \textit{Export}$: a function from id to neighbourhoods export data.
 \end{itemize}
@@
 
-Speaking by Reinforcement Learning settings, we need to define the state function and the reward function.
+Speaking of Reinforcement Learning settings, we need to define the state function and the reward function.
 %
@@
 
-Towards a complete integration between Aggregate Computing and RL, we decide to set up an Independent learning technique, 
- by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}) inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.
+As a first step towards a complete integration between Aggregate Computing and RL, we decide to set up an Independent learning technique, 
+ by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}).
+
+We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.
 
@@
 At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, 
- we combine the N Q table selecting the maximum value for each state-action pair since we assume a homogeneous behaviour, 
- where each agent perform in the same way when reaching the same state.:
+ we combine the N (the number of nodes) Q table selecting the maximum value for each state-action pair since we assume a homogeneous behaviour:
 $$
@@
 A naive implementation of this kind of problem can be described in Aggregate Computing as:
-\begin{minted}{scala}
+\begin{minted}[fontsize=\scriptsize]{scala}
 def hopCount(): Int = {
@@
  It consists of a slow convergence time when an equilibrium situation is broken.
-So, our first intuition is to accelerate -- already done algorithmically -- of convergence through learning.
- In particular, the previous solution can be rewrite as:
+So, our first intuition is to accelerate convergence through learning.
+ In particular, the previous solution can be rewritten as:
 
- \begin{minted}{scala}
+ \begin{minted}[fontsize=\scriptsize]{scala}
 def hopCountRl(k: Int, w: Int): Int = 
@@
       val action = localPolicy(state)
-      mux(source) 
-        { 0 } 
-        { minHood(nbr{ hopCounts }) 
-          + 1 
-          + action
-        } 
+      hopCout() + action
   }
@@
  In particular, in this case, it is built from a temporal window in order to encode the hop count velocity:
-\begin{minted}{scala}
+\begin{minted}[fontsize=\scriptsize]{scala}
 def speed(hopCount: Int, k: Int) =
@@
 \end{minted}
-Then, due to the partially observability of the environment, we mantain a 
+\mintinline{scala}{recentValues(data, k)} is an aggregate operator that returns a list containing the last k data perceived by the agent.
+
+Then, due to the partial observability of the environment, we maintain a 
  list of velocity, to understand the hop count direction:
 
- \begin{minted}{scala}
+ \begin{minted}[fontsize=\scriptsize]{scala}
 def state(hopCount: Int, k: Int, w: Int) =
@@
   \caption{Each agent, starting from their context, 
-  performs periodically the standard hop count. 
+  periodically performs the standard hop count. 
   Using the output produced and the history of state and output, 
   each agent computes its current state. 
-  Using it, they evaluate the action to apply to speed up the hop count convergence following a policy. 
-  This new refined output is then used to evaluate the reward and to refine the q table (dotted line path)}
+  Using it, they evaluate the action to speed up the hop count convergence following a policy. 
+  This new refined output is then used to evaluate the reward and to refine the Q table (dotted line path)}
   \label{fig:aggregate-learning-loop}
@@
 \subsection{Simulation setup}
-We verify our learning algorithm through simulation~\footnote{repository at: \url{https://github.com/cric96/alpaca-2022-ac-rl}}.
+We verify our learning algorithm through simulation~\footnote{repository at \url{https://github.com/cric96/alpaca-2022-ac-rl}} leveraging Alchemist and ScaFi.
  To check if the aggregate eventually speeds up the hop count convergence time, we create a situation where the rising problem emerges (\Cref{fig:hop-count}):
@@
  The time $t$ is fixed to 20.
- Each simulation episode least 120-time unit.
+ Each simulation episode lasts 120-time unit. Finally, we set $k = 3$ and $w = 2$.
 % 
@@
 
-Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b,c,d). To do that, we share the Q-table learn to the 
-entire aggregate and we test the program changing the system size (80, 360 and 760) nodes.
-For each population size, we run 50 episodes and we average the error at each time step. As results, 
- the aggregate speed-up convergence in each new situation.
+Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b,c,d). To do that, we share the learnt Q-table to the 
+entire aggregate and we test the program changing the system size (80, 360 and 760 nodes).
+For each population size, we run 50 episodes and then we average the error at each time step. 
+As a result, the aggregate speed-up convergence in each new situation.
 
@@
 %
-We then define a possible roadmap towards full support of a so-called \hybridaggregate{}, 
- where Machine Learning and Aggregate Computing collaborate.
+We then define a possible roadmap towards full support of Learning in our applications, 
+ bringing seamlessly Machine Learning and Aggregate Computing collaboration.
 %
 To exemplify how this integration can be performed we build a preliminary use case where 
- learning enhances successfully a standard Aggregate Computing program.
+ learning successfully enhances a standard Aggregate Computing program.
+Furthermore, the learnt program can easily scale up to system size.
 
-\subsection{Roadmap}
-The path towards full integration of Machine Learning and Aggregate Computing is long and probably we do not know the implications that it can bring. 
-Currently, we program to:
-\begin{enumerate}
-  \item Enhance building blocks through learning -- \textit{offline}: 
-  in this first step, we need to understand what kind of learning modality is most suitable for us. 
-  To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
-  \item Enhance building blocks through learning -- \textit{online}:
-  here, the most challenging part in bringing the offline results could be in the reward function definition.
-  Indeed, at runtime, we cannot use a global view of the system. 
-  A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
-  \item Create building blocks through learning: some building blocks currently lack a proper implementation -- 
-  such as the distributed leader election -- so learning can help us also in crafting them. 
-  Here, though, the research space increase drastically, 
-  so probably it might be not easy to bring our previous results there.
-  \item Improve framework aspects through learning: 
-  when we successfully apply to learn in the first two steps, 
-  we can also consider moving learning at the framework level, to manage non-functional aspects. 
-  In this way, developers can focus on the "high level" behaviour aspect.
-\end{enumerate}
+%\subsection{Roadmap}
+%The path towards full integration of Machine Learning and Aggregate Computing is long and probably we do not know the implications that it can bring. 
+%Currently, we program to:
+%\begin{enumerate}
+%  \item Enhance building blocks through learning -- \textit{offline}: 
+%  in this first step, we need to understand what kind of learning modality is most suitable for us. 
+%  To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
+%  \item Enhance building blocks through learning -- \textit{online}:
+%  here, the most challenging part in bringing the offline results could be in the reward function definition.
+%  Indeed, at runtime, we cannot use a global view of the system. 
+%  A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
+%  \item Create building blocks through learning: some building blocks currently lack a proper implementation -- 
+%  such as the distributed leader election -- so learning can help us also in crafting them. 
+%  Here, though, the research space increase drastically, 
+%  so probably it might be not easy to bring our previous results there.
+%  \item Improve framework aspects through learning: 
+%  when we successfully apply to learn in the first two steps, 
+%  we can also consider moving learning at the framework level, to manage non-functional aspects. 
+%  In this way, developers can focus on the "high level" behaviour aspect.
+%\end{enumerate}
 
@@
 \todos{future work: integrate deep learning solution, integrate actor-critic method (team learning)}
+\todos{future work: move learning at the framework level}
+
 %%

### 2021-11-07 paper-2021-woa-ac-ml.tex
put intro changes made by metaphori

@@
 \section{Introduction}
-Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs)
- due to the increasing interconnected computational devices placed in ever-changing environments.
- Examples of those include robot swarms, smart cities and augmented crowds with smart devices.
-
-CAS are notoriously hard~\cite{DBLP:conf/huc/Ferscha15} to design as they are characterised by 
-\begin{itemize}
-\item a global order that emerges from local node interactions, 
-\item decentralised control, and 
-\item adaptive behaviours towards environmental changes to pursuit a collective goal.
-\end{itemize} 
+Recent trends like pervasive and autonomic computing 
+ promote a vision of artificially designed Collective Adaptive Systems (CASs),
+ namely collections of agents coordinating and adapting %as wholes 
+ to pursue joint goals in changing environments.
+% due to the increasing interconnected computational devices placed in ever-changing environments.
+ Examples of CASs include robot swarms, smart cities, and digitally-augmented crowds.
 %
+%CASs are characterised by 
+%\begin{itemize}
+%\item a global order that emerges from local node interactions, 
+%\item  decentralised control, and 
+%\item adaptive behaviours towards environmental changes to pursuit a collective goal.
+%\end{itemize}
+%
+CAS are notoriously hard to design and control 
+ %due to as part of complex systems
+ as globally desired behaviour
+ has to emerge from local behaviours and interactions. 
 So, over the years, in literature, various approaches have tried to handle this complexity. 

### 2021-11-08 paper-2021-woa-ac-ml.tex
put right episode length
add prof. Musolesi as author

@@
 Roberto Casadei\IEEEauthorrefmark{1},
+Mirco Musolesi\IEEEauthorrefmark{1},
 Mirko Viroli\IEEEauthorrefmark{1}
@@
 \textsc{Alma Mater Studiorum}---Universit\`a di Bologna, Cesena, Italy\\
-Email: \{gianluca.aguzzi, roby.casadei, mirko.viroli\}@unibo.it
+Email: \{gianluca.aguzzi, roby.casadei, mirco.musolesi, mirko.viroli\}@unibo.it
 }}
@@
  The time $t$ is fixed to 20.
- Each simulation episode lasts 120-time unit. Finally, we set $k = 3$ and $w = 2$.
+ Each simulation episode lasts 60-time unit. Finally, we set $k = 3$ and $w = 2$.
 % 

### 2021-11-24 paper-2021-woa-ac-ml.tex
add COMA reference
first pass (Musolesi)

@@
 \title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
-\author{\IEEEauthorblockN{
+\author{
+\IEEEauthorblockN{
 Gianluca Aguzzi\IEEEauthorrefmark{1},
 Roberto Casadei\IEEEauthorrefmark{1},
-Mirco Musolesi\IEEEauthorrefmark{1},
+Mirco Musolesi\IEEEauthorrefmark{1}\IEEEauthorrefmark{2},
 Mirko Viroli\IEEEauthorrefmark{1}
-}\\
+}
 \IEEEauthorblockA{\IEEEauthorrefmark{1}
-\textsc{Alma Mater Studiorum}---Universit\`a di Bologna, Cesena, Italy\\
+\textsc{Alma Mater Studiorum}---Universit\`a di Bologna, Cesena, Italy}
+\IEEEauthorblockA{\IEEEauthorrefmark{2}
+University College London, London, United Kingdom\\
 Email: \{gianluca.aguzzi, roby.casadei, mirco.musolesi, mirko.viroli\}@unibo.it
@@
 Artificial intelligence is here to stay. 
- In particular, Machine Learning techniques in the last years outperform 
+ In particular, in the recent years, Machine Learning techniques have outperformed 
  hand-craft human solutions in different contexts, ranging from computer vision to videogames.
 %
-Currently, though, an all-encompassing Machine learning algorithm does not exist, 
- so researchers explore the possibility to apply these techniques in different fields.
+%Currently, though, an all-encompassing Machine learning algorithm does not exist, 
+% so researchers explore the possibility to apply these techniques in different fields.
 
-One of the most challenging fields for Machine Learning application is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
- Here, engineers have to handle a massive number of nodes, global-to-local behaviour mapping, and
- highly environment stochasticity.
-
-Historically, researchers favour handling them with self-organisation,
- by which a collective behaviour emerges from continuously local interaction between simple components.
-
+An interesting field for Machine Learning applications is the design of algorithms for Collective-Adaptive Systems (CASs). % Or do we want to use Cyber-Physical Swarm??
+Researchers have to deal with a series of challenges including dealing with massive number of nodes, global-to-local behaviour mapping, and highly environment stochasticity.
+Historically, researchers have favoured the use of self-organisation, by which a collective behaviour emerges from continuously local interaction between simple components.
 A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing.
  In particular, this paradigm enables collective program specification as a whole, 
- as it is based on a distributed data structure manipulation called \textit{computational fields}.
+ as it is based on a distributed data structure manipulation called \textit{computational field}.
  It is successfully used in different scenarios ranging from Swarm Robotics to Smart cities.
+Even though it is promising for expressing high-level collective behaviour system-size independent,
+ practically it is difficult to write programs that work in highly dynamic networks, 
+ where nodes have to deal with continuous changes in terms of environment.
 
-Even though it is promising for expressing high-level collective behaviour system-size independent,
- practically it is difficult to make programs work in highly dynamic networks, 
- where nodes have to deal with continuous changes at the environment side.
-
-Therefore, with this work, we stand the first step towards integration between Aggregate Computing and Machine Learning,
- in order to improve adaptivity acting in the low-level specifications but maintaining the high-level program behaviours.
-
-In the article, we outline issues, opportunities and methodologies of this integration, presenting a case study
- where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
+Therefore, with this work, we propose a solution that can be seen as the first step towards the integration between Aggregate Computing and Machine Learning, and, in particular Reinforcement Learning.
+The goal is to improve adaptivity through low-level specifications but maintaining high-level program behaviours.
+In this work, we outline issues, opportunities and methodologies of this integration, presenting a case study where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
 %
@@
  has to emerge from local behaviours and interactions. 
-So, over the years, in literature, various approaches have tried to handle this complexity. 
+For this reason, over the years, various approaches have tried to handle this complexity.
+Some of them take inspiration from the natural system composed of social animals --- like bees, fish, and ants.
 %
-Some of them take inspiration from the natural system
- composed of social animals --- like bees, fish, and ants. 
-%
-Here, a sort of hive mind arises -- 
+In this case, a sort of hive mind arises -- 
  also referred to as swarm intelligence --  
@@
 
-Initially, researchers had tried to migrate those properties into the computer science world 
- by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
+Initially, researchers had tried to translate these properties to computing systems by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
  namely trying to find the right single-node-behaviour 
@@
 %
-Though, this mapping isn't easily definable ---
- since we are coping with complex systems.
+However, this mapping cannot be easily defined, since we are dealing with complex systems.
 %
@@
 
-In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15} is an innovative solution by which 
+In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15} is an innovative paradigm, according to which
  developers can express collective self-organising behaviours through a functional program specification.
@@
 
-Aggregate Computing captures common abstractions for the collective behaviour specification. 
- These abstractions are encapsulated in the so-called ``building blocks'' through which those various
+Aggregate Computing captures common abstractions for collective behaviour specification. 
+ These abstractions are encapsulated in so-called ``building blocks'' through which various
  applications have been built 
- (such as swarm robotics, crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
-
-Though, building block synthesis concerns different stuff not only related to the behaviour itself
+ (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
+However, building block synthesis concerns different stuff not only related to the behaviour itself
  but also to the "ensemble" specifications. 
@@
  under-the-hood mechanisms that adapt the behaviour according to different environmental changes.
-
-So, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
+For this reason we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
  to developers of defining the right collective specifications.
-%
 This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
@@
 
-In particular, In this paper,
+In particular, in this paper,
  we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
@@
 
-Computer Science is an incredibly evolving research area. 
- In a few years, the environment had been filled with intercommunicating computational devices, such as smartphones, smart things, and personal computers.
+%Computer Science is an incredibly evolving research area. 
+% In a few years, the environment had been filled with intercommunicating computational devices, such as smartphones, smart things, and personal computers.
 
-In particular, in literature, we speak about Collective Adaptive Systems~\cite{DBLP:conf/huc/Ferscha15}
- when the number of nodes participating is incredibly high, there is no central authority that coordinate nodes,
- the overall behaviour is reached with local interactions, and the systems expose self-* properties.
+Collective Adaptive Systems~\cite{DBLP:conf/huc/Ferscha15} are characterized by a very high number of nodes without a central coordinating authority that coordinate nodes.
+The overall behaviour is a consequence of local interactions.
+The systems expose self-* properties.
 
@@
 %
-Though, a natural way to describe this kind of problem is by using a \textit{collective} view of the system.
+ A natural way to describe this kind of problem is through a \textit{collective} view of the system.
  This is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
@@
 % 
-Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
+Also, ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
  a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)
 
-In our works, we are focused on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
+In this work, we focus on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
  the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
-
 The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
  map from any devices to their computational value (i.e. the result of a computation).
-
 The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
  to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
-
 The novelty of Aggregate Computing consists in its composable (functional-like) system-size independent program definition: 
 using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
-
 By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} 
@@
 \end{itemize}
+For a more in-depth explanation of these operators, please refer to~\cite{DBLP:journals/computer/BealPV15}.
 
-For a more in-depth explanation of these operators, please refer to~\cite{DBLP:journals/computer/BealPV15}.
 \subsection{Computational Model}
 An aggregate program execution consists of a continuously node-point execution of a \round{}, 
- that is the atomical part of aggregate computing.
-%
-Each node can interact only with-in its neighbourhoods sharing the \export{}, 
- namely the result produced by the evaluation of a \round{}.
+ that is the atomical part of aggregate computing. Each node can interact only with-in its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.
 
@@
 Rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
- This makes it possible to scale easily with the node number size.
+ This makes it possible to scale easily with the node number size. This collective, proactive and periodical execution will eventually lead to the satisfaction of the system-wide specification.
 
-This collective, proactive and periodical execution will eventually reach the system-wide specification.
+\section{Aggregate Computing and Machine Learning}
 
-\section{Aggregate computing and Machine learning}
-
-In this section,
- we analyse the characteristics of the aggregate computing paradigm,
- map these to relevant machine learning contributions in literature,
- and pointing out different learning perspectives about the integration of Aggregate Computing with Machine Learning
+In this section, we analyse the characteristics of the aggregate computing paradigm,
+ map these to relevant machine learning contributions in the literature,
+ and point out different learning perspectives about the integration of Aggregate Computing with Machine Learning
 
@@
 %
-In recent decades there has been an emerging trend in the use of Reinforcement Learning 
+In recent decades there has been an emerging trend in the use of Reinforcement Learning (RL) 
  in multi-agent settings (called Multi-Agent Reinforcement Learning), as a powerful, robust and adaptive learning paradigm.
@@
 
-The game theory viewpoint instead has tried to find techniques that achieve equilibrium conditions such as Nash-equilibrium of Pareto optimality.
+Other efforts have focussed on achieving equilibria such as Nash-equilibrium of Pareto optimality.
  The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q~\cite{DBLP:conf/icml/Littman94}.
@@
 %
-In aggregate systems, a device can directly interact only with neighbours.
+In aggregate systems, a device can directly interact only with its neighbours.
 %
@@
 In other words, the system state is only \emph{partially observable}.
-
 In practice, CTDE technique aims to solve these problems, using a \emph{global} view at training time but
@@
 
-A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14}.
+A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14, DBLP:conf/aaai/FoersterFANW18}.
 It is used to capture agents contributions to the system's global performance.
@@
 When an aggregate behaviour is carried out,
- we often focus on the stabilised outcomes,
- tolerating outcomes that are generated
- during the transient phase.
+ we often focus on the final stable outcomes after the transitory period.
 %
-However, intermediate results are often important,
+However, intermediate results are also important,
  and invariants should hold throughout the entire computation.
@@
 
-\subsection{Integration perspectives}
-Here we outline the different perspectives about Aggregate Computing with learning, 
+\subsection{Integrating Learning Mechanisms into Aggregate Computing}
+We now outline the different perspectives about Aggregate Computing with learning, 
  to clarify the current and the future research path.
@@
 %
-With Independent learning though is still possible to reach a global collective behaviour convergence. 
+With Independent learning is still possible to reach a global collective behaviour convergence. 
  Indeed using a common reward signal, the collective may reach the same behaviour through emergence \cite{iima2008swarm, nguyen2018swarm}.
@@
  Here indeed, we can leverage simulation to know the right computational field and then use one of the novel CTDE techniques. 
-
 Mixed-time learning can be a mid-level complexity direction instead.
@@
  the behaviour can be adapted based on the new environmental situations. 
-
 Finally, we can imagine that learning can be performed online. 
  It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
-
  In mixed and online cases, Reinforcement Learning is the most suitable model, because agents \textit{learn by doing}.
@@
  by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}). 
-
+% 
 We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.
@@
 \end{figure*}
-In this section, we outline the results of the Reinforcement Learning application in one simple
+In this section, we outline the results of the application of RL for one simple
  aggregate program: the hop count.
-
-Here, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes ((\Cref{fig:hop-count})).
+In this case, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes ((\Cref{fig:hop-count})).
 
@@
 \end{minted}
-This solution though suffers from the slow-rising problem. 
+However, this solution suffers from the slow-rising problem. 
  It consists of a slow convergence time when an equilibrium situation is broken.
@@
 \mintinline{scala}{recentValues(data, k)} is an aggregate operator that returns a list containing the last k data perceived by the agent.
-
 Then, due to the partial observability of the environment, we maintain a 
@@
 
-\subsection{Simulation setup}
+\subsection{Simulation Setup}
 We verify our learning algorithm through simulation~\footnote{repository at \url{https://github.com/cric96/alpaca-2022-ac-rl}} leveraging Alchemist and ScaFi.
@@
  to eventually reach the distance towards $A$.
-%
 We use 80 ($N$) nodes displaced into an irregular grid with a width of 100 meters and a height of 20 meters. 
@@
 % 
-We training the aggregate for 500 episodes.
+We train the aggregate for 500 episodes.
  The Q-learning parameters are:
@@
 \end{equation*}
-In doing so, we what that agent to explore as more as possible the state-action space.
+In doing so, the agent will explore the state-action space as much as possible.
 
-\Cref{fig:simulation}-a shows the error for both naive implementation and Reinforcement Learning based implementation as time changes.
+\Cref{fig:simulation}-a shows the error for both naive implementation and RL-based implementation as time changes.
 The error is evaluated as:
@@
 
-Where \textbf{i} is the node identifier and \textbf{t} is the time when the error is evaluated.
+where \textbf{i} is the node identifier and \textbf{t} is the time when the error is evaluated.
 %
@@
     \caption{The \textit{orange} line 
-    describes the Reinforcement Learning solution error. The \textit{blue} line shows
+    describes the RL solution error. The \textit{blue} line shows
     the naive solution error.}
@@
   \end{subfigure}
-  \caption{ (a) shows the $total_{error}$ evolution during the learning process. (b), (c), and (d) shows the average error in three different simulations with 80, 360 and 760 nodes. 
-  The black line is the error done by RL based solution, 
-  the green line is the error produced by standard hop count implementation}
+  \caption{Analysis of the learning process.
+(a) shows the $total_{error}$ evolution during the learning process.
+(b), (c), and (d) shows the average error in three different simulations with 80, 360 and 760 nodes.
+  The black line is the error of the RL-based solution, 
+  the green line is the error produced by the standard hop count implementation.}
   \label{fig:simulation}
@@
 
-\section{Conclusion and Discussion}
-In this article, we briefly overview how Aggregate Computing can be mixed with Machine Learning -- 
- in particular with Reinforcement Learning. 
- This leads us to compare Aggregate Computing problems with current Machine Learning solutions in MASs systems.
+\section{Discussion and Concluding Remarks}
+In this article, we have discussed a preliminary example of how Aggregate Computing can be integrated with Reinforcement Learning. 
+We have then compared Aggregate Computing problems with current Machine Learning solutions in MASs systems.
 %
-We then define a possible roadmap towards full support of Learning in our applications, 
- bringing seamlessly Machine Learning and Aggregate Computing collaboration.
-%
-To exemplify how this integration can be performed we build a preliminary use case where 
+We have defined a possible roadmap towards full support of learning mechanisms in Aggregate Computing. %
+As a proof of concept, we have discussed preliminary use case where 
  learning successfully enhances a standard Aggregate Computing program.
-Furthermore, the learnt program can easily scale up to system size.
+Furthermore, we have shown that the learnt program can easily scale up to system size.
 

### 2021-11-25 paper-2021-woa-ac-ml.tex
on intro outline

@@
 Artificial intelligence is here to stay. 
- In particular, in the recent years, Machine Learning techniques have outperformed 
+ In particular, in recent years, Machine Learning techniques have outperformed 
  hand-craft human solutions in different contexts, ranging from computer vision to videogames.
@@
  The goal is to improve adaptivity through low-level specifications but maintaining high-level program behaviours.
+
 In this work, we outline issues, opportunities and methodologies of this integration, presenting a case study where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
@@
 
-In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15} is an innovative paradigm, according to which
+In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}(AC) is an innovative paradigm, according to which
  developers can express collective self-organising behaviours through a functional program specification.
@@
 
-Aggregate Computing captures common abstractions for collective behaviour specification. 
+AC captures common abstractions for collective behaviour specification. 
  These abstractions are encapsulated in so-called ``building blocks'' through which various
@@
  under-the-hood mechanisms that adapt the behaviour according to different environmental changes.
-For this reason we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
+For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
  to developers of defining the right collective specifications.
 This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
- Aggregate Computing abstractions and another part is distilled through learning~\cite{research}.
+ AC abstractions and another part is distilled through learning~\cite{research}.
 
@@
  Furthermore, we show that the learnt policy scales up to different system sizes.
-\todos{Outline}
 
-\section{Background and Related Work}
+The paper is divided as follow, in \Cref{background} we describe
+ the AC framework, showing the context in which
+ is typically applied and the overall idea of how it works, in \Cref{aggregate-and-machine-learning} we detail the relevant properties of Aggregate
+ Computing in terms of machine learning problems, in \Cref{aggregate-and-rl} we describe a combination of AC with RL and then in
+ \Cref{evaluation} we practically applied it in a paradigmatic AC case study,
+ finally, \Cref{conclusion} we discuss the result showed and we underline a path toward an in-depth AC and RL integration.
+
+ \section{Background and Related Work}\label{background}
 
@@
 
-\section{Aggregate Computing and Machine Learning}
+\section{Aggregate Computing and Machine Learning}\label{aggregate-and-machine-learning}
 
@@
 The foster is the common scenario in multi-agent settings.
- Here indeed, we can leverage simulation to know the right computational field and then use one of the novel CTDE techniques. 
+ Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
 Mixed-time learning can be a mid-level complexity direction instead.
@@
 
-\section{Aggregate Computing with Reinforcement}
+\section{Aggregate Computing with Reinforcement}\label{aggregate-and-rl}
 \begin{figure}
@@
 Then, in the next episode, each agent starts with the $Q_{max}$ and try to refine it.
-\section{Evaluation}
+\section{Evaluation}\label{evaluation}
 \begin{figure*}[h]
@@
 
-\section{Discussion and Concluding Remarks}
+\section{Discussion and Concluding Remarks}\label{conclusion}
 In this article, we have discussed a preliminary example of how Aggregate Computing can be integrated with Reinforcement Learning. 

### 2021-11-25 paper-2021-woa-ac-ml.tex
fix acronyms (Aggregate Computing => AC, Multi-Agent Reinforcement Learning => MARL, Reinforcement Learning => RL, Supervised Learning => SL

@@
 %% Utility commands
-\newcommand{\hybridaggregate}{\textit{Hybrid Aggregate Computing}}
 \newcommand{\round}{\texttt{round}}
@@
 %
-Indeed, an aggregate program (i.e. a program written with Aggregate Computing) consists of the manipulation of \textit{computational fields}, a distributed
+Indeed, an aggregate program (i.e. a program written with AC) consists of the manipulation of \textit{computational fields}, a distributed
  data structure where each point in the spacetime is associated with the result of a computation
@@
 In particular, in this paper,
- we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
+ we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
  where we outline the performance improvement of the ``hop count" program --- 
@@
  to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
-The novelty of Aggregate Computing consists in its composable (functional-like) system-size independent program definition: 
+The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
 using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
@@
 
-Aggregate Computing is studied in various simulated scenarios ranging from smart cities to crowds, 
+AC is studied in various simulated scenarios ranging from smart cities to crowds, 
  thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
@@
 An aggregate program execution consists of a continuously node-point execution of a \round{}, 
- that is the atomical part of aggregate computing. Each node can interact only with-in its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.
+ that is the atomical part of AC. Each node can interact only with-in its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.
 
@@
 
-In this section, we analyse the characteristics of the aggregate computing paradigm,
+In this section, we analyse the characteristics of the AC paradigm,
  map these to relevant machine learning contributions in the literature,
- and point out different learning perspectives about the integration of Aggregate Computing with Machine Learning
+ and point out different learning perspectives about the integration of AC with Machine Learning
 
@@
 %
-In recent decades there has been an emerging trend in the use of Reinforcement Learning (RL) 
- in multi-agent settings (called Multi-Agent Reinforcement Learning), as a powerful, robust and adaptive learning paradigm.
+In recent decades there has been an emerging trend in the use of RL 
+ in multi-agent settings -- called Multi-Agent Reinforcement Learning (MARL) -- as a powerful, robust and adaptive learning paradigm.
  Progress has been considerable and a wide range of algorithms are now available.
 
-Multi-Agent Reinforcement Learning is a conjunction of Game theory and Reinforcement Learning, 
+MARL is a conjunction of Game theory and RL, 
  and there are several (even orthogonal) viewpoints on which researchers have been focused.
@@
 The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
- In Aggregate Computing instead, we need to find a technique that does not depend on the system node count.
+ In AC instead, we need to find a technique that does not depend on the system node count.
 
@@
 \subsection{Integrating Learning Mechanisms into Aggregate Computing}
-We now outline the different perspectives about Aggregate Computing with learning, 
+We now outline the different perspectives about AC with learning, 
  to clarify the current and the future research path.
@@
 % 
-In Aggregate Computing, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
+In AC, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
  be applied seamlessly.
@@
 On the other hand, Independent Learning is usually applied in Collective
- (Self-)Adaptive System \cite{csas-and-marl}, a field near to Aggregate Computing.
+ (Self-)Adaptive System \cite{csas-and-marl}.
 
@@
 Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
- such as Reinforcement Learning or Supervised Learning.
+ such as RL or Supervised Learning (SL).
 %
-Supervised learning is rare in CSAS -- or in the ``swarm" like system in general.
+SL is rare in CSAS -- or in the ``swarm" like system in general.
 %
@@
 
-In Aggregate Computing, Supervised Learning can be applied when we want to build up new building blocks from zero. 
+In AC, S) can be applied when we want to build up new building blocks from zero. 
 % 
@@
  It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
- In mixed and online cases, Reinforcement Learning is the most suitable model, because agents \textit{learn by doing}.
+ In mixed and online cases, RL is the most suitable model, because agents \textit{learn by doing}.
 
@@
   \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
-  \caption{This figure shows the main steps of RL in Aggregate Computing.
+  \caption{This figure shows the main steps of RL in AC.
   Each agent starts with his copy of a Q table. 
@@
 % 
-In this section, we briefly describe how Reinforcement Learning can be injected into the computation loop,
- and what Reinforcement Learning we used with Aggregate Computing.
+In this section, we briefly describe how RL can be injected into the computation loop,
+ and what RL we used with AC.
 
@@
 
-Speaking of Reinforcement Learning settings, we need to define the state function and the reward function.
+Speaking of RL settings, we need to define the state function and the reward function.
 %
@@
 
-As a first step towards a complete integration between Aggregate Computing and RL, we decide to set up an Independent learning technique, 
+As a first step towards a complete integration between AC and RL, we decide to set up an Independent learning technique, 
  by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}). 
@@
 
-A naive implementation of this kind of problem can be described in Aggregate Computing as:
+A naive implementation of this kind of problem can be described in AC as:
 \begin{minted}[fontsize=\scriptsize]{scala}
@@
 \section{Discussion and Concluding Remarks}\label{conclusion}
-In this article, we have discussed a preliminary example of how Aggregate Computing can be integrated with Reinforcement Learning. 
-We have then compared Aggregate Computing problems with current Machine Learning solutions in MASs systems.
+In this article, we have discussed a preliminary example of how AC can be integrated with RL. 
+We have then compared AC problems with current Machine Learning solutions in MASs systems.
 %
-We have defined a possible roadmap towards full support of learning mechanisms in Aggregate Computing. %
+We have defined a possible roadmap towards full support of learning mechanisms in AC. %
 As a proof of concept, we have discussed preliminary use case where 
- learning successfully enhances a standard Aggregate Computing program.
+ learning successfully enhances a standard AC program.
 Furthermore, we have shown that the learnt program can easily scale up to system size.

### 2021-11-26 paper-2021-woa-ac-ml.tex
fix fig 4 caption

@@
     \hspace{5mm}
+    \label{fig:simulation-a}
   \end{subfigure}
@@
   \caption{Analysis of the learning process.
-(a) shows the $total_{error}$ evolution during the learning process.
-(b), (c), and (d) shows the average error in three different simulations with 80, 360 and 760 nodes.
+\Cref{fig:simulation-a} shows the $total_{error}$ evolution during the learning process.
+\Cref{fig:simulation-b}, \Cref{fig:simulation-c}, and \Cref{fig:simulation-d} shows the average error in three different simulations with 80, 360 and 760 nodes.
   The black line is the error of the RL-based solution, 

### 2021-11-26 paper-2021-woa-ac-ml.tex
add references to autonomic and pervasive computing

@@
 
-In this work, we outline issues, opportunities and methodologies of this integration, presenting a case study where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
+In this paper, we outline issues, opportunities and methodologies of this integration, presenting a case study where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
 %
@@
 \section{Introduction}
-Recent trends like pervasive and autonomic computing 
+Recent trends like pervasive~\cite{DBLP:journals/wc/Satyanarayanan01} and autonomic~\cite{DBLP:journals/computer/KephartC03} computing 
  promote a vision of artificially designed Collective Adaptive Systems (CASs),
@@
 \Cref{fig:simulation-a} shows the $total_{error}$ evolution during the learning process.
-\Cref{fig:simulation-b}, \Cref{fig:simulation-c}, and \Cref{fig:simulation-d} shows the average error in three different simulations with 80, 360 and 760 nodes.
+\Cref{fig:simulation-b}, \Cref{fig:simulation-c}, and \Cref{fig:simulation-d} show the average error in three different simulations with 80, 360 and 760 nodes.
   The black line is the error of the RL-based solution, 

### 2021-11-26 paper-2021-woa-ac-ml.tex
minor on intro (w.r.t => w.r.t.)

@@
 
-In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}(AC) is an innovative paradigm, according to which
+In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is an innovative paradigm, according to which
  developers can express collective self-organising behaviours through a functional program specification.
@@
  For instance, it is quite difficult to craft building blocks
- that easily adapt w.r.t the node movements rapidity.
+ that easily adapt w.r.t. the node movements rapidity.
 

### 2021-11-27 paper-2021-woa-ac-ml.tex
add future works and discussion

@@
 Recent trends like pervasive~\cite{DBLP:journals/wc/Satyanarayanan01} and autonomic~\cite{DBLP:journals/computer/KephartC03} computing 
- promote a vision of artificially designed Collective Adaptive Systems (CASs),
+ promote a vision of artificially designed Collective Adaptive Systems (CASs)~\cite{DBLP:conf/huc/Ferscha15},
  namely collections of agents coordinating and adapting %as wholes 
@@
 
-Collective Adaptive Systems~\cite{DBLP:conf/huc/Ferscha15} are characterized by a very high number of nodes without a central coordinating authority that coordinate nodes.
+Collective Adaptive Systems are characterized by a very high number of nodes without a central coordinating authority that coordinate nodes.
 The overall behaviour is a consequence of local interactions.
@@
 The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
- to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.
+ to express spatio-temporal computations~\cite{DBLP:conf/coordination/AudritoBDV18}.
 The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
@@
 \section{Discussion and Concluding Remarks}\label{conclusion}
-In this article, we have discussed a preliminary example of how AC can be integrated with RL. 
-We have then compared AC problems with current Machine Learning solutions in MASs systems.
+In this article, we have discussed the combination of AC with Machine Learning techniques in order to improve adaptivity
+ and effictveness of the AC framework.
+In doing so, we compared AC problems with current Machine Learning solutions in MASs systems.
 %
-We have defined a possible roadmap towards full support of learning mechanisms in AC. %
-As a proof of concept, we have discussed preliminary use case where 
- learning successfully enhances a standard AC program.
+Finally, we explore the combination of AC with RL (in particular Hysteretic Q-Learning).
+RL application in this kind of settings is quite an explored -- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
+ area for several reason, such us the highly system dynamic, non-stationarity,  partial enviroment observability, and non-bounded system size.
+In a case study we empirically shown the effictveness of RL, by which we speed up the convergence time of
+ aggregate program taking in consideration -- the hop count.
 Furthermore, we have shown that the learnt program can easily scale up to system size.
 
-%\subsection{Roadmap}
-%The path towards full integration of Machine Learning and Aggregate Computing is long and probably we do not know the implications that it can bring. 
-%Currently, we program to:
-%\begin{enumerate}
-%  \item Enhance building blocks through learning -- \textit{offline}: 
-%  in this first step, we need to understand what kind of learning modality is most suitable for us. 
-%  To decrease the complexity, we aim at perform learning \textit{offline} and at the \textit{building block} level.
-%  \item Enhance building blocks through learning -- \textit{online}:
-%  here, the most challenging part in bringing the offline results could be in the reward function definition.
-%  Indeed, at runtime, we cannot use a global view of the system. 
-%  A novel way to describe a collective reward by Aggregate Computing might be taken into consideration.
-%  \item Create building blocks through learning: some building blocks currently lack a proper implementation -- 
-%  such as the distributed leader election -- so learning can help us also in crafting them. 
-%  Here, though, the research space increase drastically, 
-%  so probably it might be not easy to bring our previous results there.
-%  \item Improve framework aspects through learning: 
-%  when we successfully apply to learn in the first two steps, 
-%  we can also consider moving learning at the framework level, to manage non-functional aspects. 
-%  In this way, developers can focus on the "high level" behaviour aspect.
-%\end{enumerate}
+This is a first effort towards an ``hybrid'' collective programming approach, where the high-level program specification is 
+ done by a designer and the low-level matters are leaved to machine learning algorithms.
 
-\todos{discussion about the result}
-\todos{limitations: it does not easily scale with the complex application, it cannot be applied online, it cannot be used in different scenarios}
-\todos{future work: integrate deep learning solution, integrate actor-critic method (team learning)}
-\todos{future work: move learning at the framework level}
+We know that this is preliminary work and it shed light on several problems that we should handle in the future.
 
-%%
-%% Define the bibliography file to be used
+Aggregate systems are partially observable from the perspective of each node. 
+ Local observations may not be statistically significant in summarising the state of the entire aggregate. 
+ In the future, we could address this problem through recurrent networks and a CTDE learning methodology.
+
+The learning, in this case, was applied at the building block level. 
+ Another possible direction is to use it at the framework level, e.g., to select which messages to send to the neighbourhood or not. 
+ This has already been addressed in the literature, usually under the name of \textit{learn to communicate}~\cite{DBLP:journals/corr/abs-1908-03963}. 
+ A novel way to handle this problem is to use Graph Neural Networks~\cite{DBLP:journals/corr/abs-1812-08434} to learn the representation of a distributed system and to select messages based on that representation.
+ Thus, we should consider this methodology when we will extend RL at the AC framework level.
+
+In this case, we applied a variation of the independent learning approach. 
+ Agents first learn the best policy locally, and then at the end of the episode, a learner merges the Q-tables. 
+ However, this does not reduce the problems of scalability and non-stationary, since a central agent must still collect all the Q-tables of the aggregate. 
+ In particular, to handle the scalability issues, we will have to consider decentralised approaches based only on the neighbourhood information (or local leaders).
+
+Finally, we do not address the multi-agent credit assignment problem in this work. 
+ Indeed, we use a global reward function for the whole aggregate. In this way, however, learning often degenerated because it was difficult to understand the influence of an agent w.r.t. the system. 
+ To reduce this problem, we will consider novel techniques of difference reward~\cite{DBLP:conf/atal/AgoginoT04} or counterfactual baseline~\cite{DBLP:journals/corr/FoersterFANW17}.
 

### 2021-11-27 paper-2021-woa-ac-ml.tex
fit intro+abstract in one page

@@
 For this reason, over the years, various approaches have tried to handle this complexity.
-Some of them take inspiration from the natural system composed of social animals --- like bees, fish, and ants.
+Some of them take inspiration from the natural system composed of social animals.
+%--- like bees, fish, and ants.
 %
-In this case, a sort of hive mind arises -- 
- also referred to as swarm intelligence --  
+In this case, a sort of hive mind arises 
+ %-- also referred to as swarm intelligence --  
  born from local interactions between agents. 
@@
 
-Initially, researchers had tried to translate these properties to computing systems by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
- namely trying to find the right single-node-behaviour 
- to reach a global behaviour, e.g. flocking, foraging, etc.
+Initially, researchers had tried to translate these properties to computing systems by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}. %, 
+ %namely trying to find the right single-node-behaviour 
+ %to reach a global behaviour. %, e.g. flocking, foraging, etc.
 %
@@
 In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is an innovative paradigm, according to which
- developers can express collective self-organising behaviours through a functional program specification.
+ developers can express collective self-organising behaviours. % through a functional program specification.
 %
-Indeed, an aggregate program (i.e. a program written with AC) consists of the manipulation of \textit{computational fields}, a distributed
- data structure where each point in the spacetime is associated with the result of a computation
- performed by a node.
+Indeed, an aggregate program 
+ %(i.e. a program written with AC) 
+ consists of the manipulation of \textit{computational fields}, a distributed
+ data structure where each node is associated with the result of a computation
+ performed by it.
 %
@@
 
-AC captures common abstractions for collective behaviour specification. 
- These abstractions are encapsulated in so-called ``building blocks'' through which various
- applications have been built 
- (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
+AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
+ %These abstractions are encapsulated in so-called ``building blocks''.
+ %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
 However, building block synthesis concerns different stuff not only related to the behaviour itself
@@
  that easily adapt w.r.t. the node movements rapidity.
-
+%
 Ideally, we would like to continue building our application using these building blocks -- so the high-level specification remains the same -- but using some
@@
  to developers of defining the right collective specifications.
-This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
- AC abstractions and another part is distilled through learning~\cite{research}.
-
+%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
+% AC abstractions and another part is distilled through learning~\cite{research}.
 In particular, in this paper,

### 2021-11-27 paper-2021-woa-ac-ml.tex
Pass on abstract

@@
 \begin{abstract}
-Artificial intelligence is here to stay. 
- In particular, in recent years, Machine Learning techniques have outperformed 
- hand-craft human solutions in different contexts, ranging from computer vision to videogames.
+Recent trends like pervasive, autonomic, and swarm computing
+ promote 
+ a vision of large-scale collections of relatively simple agents
+ that act and coordinate 
+ with no central coordinator
+ to support distributed applications.
 %
-%Currently, though, an all-encompassing Machine learning algorithm does not exist, 
-% so researchers explore the possibility to apply these techniques in different fields.
-
-An interesting field for Machine Learning applications is the design of algorithms for Collective-Adaptive Systems (CASs). % Or do we want to use Cyber-Physical Swarm??
-Researchers have to deal with a series of challenges including dealing with massive number of nodes, global-to-local behaviour mapping, and highly environment stochasticity.
-Historically, researchers have favoured the use of self-organisation, by which a collective behaviour emerges from continuously local interaction between simple components.
-A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing.
- In particular, this paradigm enables collective program specification as a whole, 
- as it is based on a distributed data structure manipulation called \textit{computational field}.
- It is successfully used in different scenarios ranging from Swarm Robotics to Smart cities.
-Even though it is promising for expressing high-level collective behaviour system-size independent,
- practically it is difficult to write programs that work in highly dynamic networks, 
- where nodes have to deal with continuous changes in terms of environment.
-
-Therefore, with this work, we propose a solution that can be seen as the first step towards the integration between Aggregate Computing and Machine Learning, and, in particular Reinforcement Learning.
- The goal is to improve adaptivity through low-level specifications but maintaining high-level program behaviours.
-
-In this paper, we outline issues, opportunities and methodologies of this integration, presenting a case study where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
+Such Collective Adaptive Systems (CAS)
+ are hard to engineer,
+ because mechanisms for developing global behaviour
+ out of local behaviour and interaction must be devised,
+ often leveraging natural inspiration.
+%
+In particular, in the field-based Aggregate Computing approach,
+ behaviour is specified in terms of computational fields,
+ namely maps from CAS entities to computational values evolving over time, and neighbour-based interaction.
+%
+However, devising efficient aggregate computing algorithms
+ for different scenarios
+ is not an easy task; in particular,
+ a problem is determining the best local decisions based on the current context
+ for fostering collective outcomes.
+%
+Therefore, in this work we investigate
+ the possibility of using machine learning techniques
+ to automatically devise local actions
+ to improve existing aggregate algorithms.
+%
+In particular, we propose a Reinforcement Learning solution that can be seen as the first step towards the integration between Aggregate Computing and Machine Learning. 
+% The goal is to improve adaptivity through low-level specifications but maintaining high-level program behaviours. 
+Accordingly, we outline challenges and opportunities of this integration, and present preliminary results based on a case study where Aggregate Computing is successfully combined with Hysteretic Q-Learning.
 %

### 2021-11-27 paper-2021-woa-ac-ml.tex
abstract minors

@@
  that act and coordinate 
- with no central coordinator
+ with no central orchestrator
  to support distributed applications.
@@
 In particular, in the field-based Aggregate Computing approach,
- behaviour is specified in terms of computational fields,
+ collective adaptive behaviour is specified in terms of computational fields,
  namely maps from CAS entities to computational values evolving over time, and neighbour-based interaction.

### 2021-11-28 paper-2021-woa-ac-ml.tex
Ref needed on slow-rising
Correct scafi citation + remove self-citations of little relevance here

@@
  thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
- this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
+ this paradigm, such as ScaFi~\cite{DBLP:conf/isola/CasadeiVAD20}. %, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
 
@@
 \end{minted}
-However, this solution suffers from the slow-rising problem. 
+However, this solution suffers from the slow-rising problem~\cite{DBLP:conf/saso/AudritoCDV17}. 
  It consists of a slow convergence time when an equilibrium situation is broken.

### 2021-11-28 paper-2021-woa-ac-ml.tex
Use JLAMP as proxy for main results

@@
 The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
- to express spatio-temporal computations~\cite{DBLP:conf/coordination/AudritoBDV18}.
+ to express spatio-temporal computations~\cite{DBLP:journals/jlap/ViroliBDACP19}.
 The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
 using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
-By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} 
+By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency %~\cite{DBLP:journals/taas/BealVPD17} 
  (i.e. the ensemble behaviour is independent of the underlying network details)
- and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).
+ and self-stabilisation %~\cite{DBLP:journals/tomacs/ViroliABDP18} 
+ (i.e. the ability of a system to recover from arbitrary changes)~\cite{DBLP:journals/jlap/ViroliBDACP19}.
 

### 2021-11-28 paper-2021-woa-ac-ml.tex
Citation needed if we claim AC to be used in various apps

@@
 
-AC is studied in various simulated scenarios ranging from smart cities to crowds, 
+AC is studied in various simulated scenarios ranging from smart cities to crowd management~\cite{DBLP:journals/jlap/ViroliBDACP19}, 
  thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 

### 2021-11-28 paper-2021-woa-ac-ml.tex
cut Section 3 and move into motivation

@@
  the AC framework, showing the context in which
- is typically applied and the overall idea of how it works, in \Cref{aggregate-and-machine-learning} we detail the relevant properties of Aggregate
- Computing in terms of machine learning problems, in \Cref{aggregate-and-rl} we describe a combination of AC with RL and then in
+ is typically applied and the overall idea of how it works, and then the motivation of injecting Reinforcement Learning in this paradigm,
+ %in \Cref{aggregate-and-machine-learning} we detail the relevant properties of Aggregate Computing in terms of machine learning problems, 
+ in \Cref{aggregate-and-rl} we describe a combination of AC with RL and then in
  \Cref{evaluation} we practically applied it in a paradigmatic AC case study,
@@
 
-\section{Aggregate Computing and Machine Learning}\label{aggregate-and-machine-learning}
+\subsection{Motivation}
+Currently, Aggregate computing has a problem in the ability to generalise collective behaviours that are capable of working well in different scenarios. 
+ For example, a well-tested aggregate application in a static network could lead to a non-stable behaviour in a dynamic network. 
+ Typically, to alleviate this problem, designers create ad-hoc building blocks that cope with certain environmental aspects (such as the ULT block~\cite{DBLP:conf/saso/AudritoCDV17}). 
+ But they are still context-specific solutions.
 
-In this section, we analyse the characteristics of the AC paradigm,
- map these to relevant machine learning contributions in the literature,
- and point out different learning perspectives about the integration of AC with Machine Learning
+A novel way to deal with this problem is to leave the low-level aspects of AC to machine learning, 
+ training the collective in different environments and thus building general behaviours that are capable of adapting to many circumstances.
 
-\paragraph{Multi-and many-agent system}
-%
-An aggregate system is a multi-agent system
- and, often, a \emph{many}-agent system
- where a large number (hundreds or more)
- of autonomous entities are programmed to achieve 
- some collective behaviour through \emph{repeated} 
- sensing, computation, communication, and actuation steps.
-%
-Due to the high stochasticity of the environment,
- it is almost impossible to know and
- program the optimal behaviour for all agents in advance.
- This uncertainty results in the need of creating intelligent agents
- so that they can learn the optimal behaviour and adapt to environmental changes.
-%
-In recent decades there has been an emerging trend in the use of RL 
- in multi-agent settings -- called Multi-Agent Reinforcement Learning (MARL) -- as a powerful, robust and adaptive learning paradigm.
- Progress has been considerable and a wide range of algorithms are now available.
+In this case, when Reinforcement Learning is applied to several agents, we speak of Multi-Agent Reinforcement Learning (MARL)~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
+ AC has a typical cooperative and homogeneous learning setting: 
+ each agent collaborates to achieve a common goal guided by the same reward function.
+In this setting, several works have tried to apply the algorithms known in the single-agent context to the multi-agent one - the so-called Independent MARL~\cite{DBLP:journals/tsmc/BusoniuBS08}. 
+ However, several issues emerge, such as non-stationarity, multi-agent credit assignment, scalability and partial observability of the system. 
+ Therefore, several heuristics have tried to solve these problems.
+Among many Hysteric Q-Learning~\cite{hysteretic-q} uses an optimistic approach in which more weight is given to positive actions than negative ones.
 
-MARL is a conjunction of Game theory and RL, 
- and there are several (even orthogonal) viewpoints on which researchers have been focused.
+Another way to handle multi-agent learning is to use so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 
+ In this case, a central agent collects the experiences of the various agents and distils a policy that each agent can execute locally in a typical actor-critic setting. 
+ During the learning phase, the problems of partial observability and non-stationarity are reduced, since the agent has a global view. 
+ However, the problem of scalability is typically not addressed.
 
-First attempts from RL viewpoint, 
- goes towards a so-called \textit{independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
- where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}.
+In the next section, we exploit Hysteretic Q-Learning (since it empirically showed good performance) with AC, as the first step towards \textit{``hybrid"} techniques for programming aggregate systems.
+ In this way, an application part is described in terms of aggregate computations and the low-level part is left to learning.
+% \section{Aggregate Computing and Machine Learning}\label{aggregate-and-machine-learning}
 
-Examples of these are Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
- Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value~\cite{DBLP:conf/atal/KaisersT10}.
-%
-The pro of these approaches is that their complexity does not scale up with the number of agents, 
- but, unfortunately, the learning process is extremely non-stationary and unstable.
-% 
-Furthermore, they are heuristic and do not exist any convergence proof (even if, in practice, they often reach a good policy).
+% In this section, we analyse the characteristics of the AC paradigm,
+%  map these to relevant machine learning contributions in the literature,
+%  and point out different learning perspectives about the integration of AC with Machine Learning
 
-Other efforts have focussed on achieving equilibria such as Nash-equilibrium of Pareto optimality.
- The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q~\cite{DBLP:conf/icml/Littman94}.
-The problem is that it is used with few agents and scale hardly with the application complexity.
-%
-Finally, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
- agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
- Most promising approaches in this direction are MAPPO~\cite{DBLP:journals/corr/abs-2103-01955}, COMA~\cite{DBLP:journals/corr/FoersterFANW17} and MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17}.
+% \paragraph{Multi-and many-agent system}
+% %
+% An aggregate system is a multi-agent system
+%  and, often, a \emph{many}-agent system
+%  where a large number (hundreds or more)
+%  of autonomous entities are programmed to achieve 
+%  some collective behaviour through \emph{repeated} 
+%  sensing, computation, communication, and actuation steps.
+% %
+% Due to the high stochasticity of the environment,
+%  it is almost impossible to know and
+%  program the optimal behaviour for all agents in advance.
+%  This uncertainty results in the need of creating intelligent agents
+%  so that they can learn the optimal behaviour and adapt to environmental changes.
+% %
+% In recent decades there has been an emerging trend in the use of RL 
+%  in multi-agent settings -- called Multi-Agent Reinforcement Learning (MARL) -- as a powerful, robust and adaptive learning paradigm.
+%  Progress has been considerable and a wide range of algorithms are now available.
 
-The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
- In AC instead, we need to find a technique that does not depend on the system node count.
+% MARL is a conjunction of Game theory and RL, 
+%  and there are several (even orthogonal) viewpoints on which researchers have been focused.
 
-\paragraph{Neighbour-based or indirect (environment-mediated) interaction}
-%
-In aggregate systems, a device can directly interact only with its neighbours.
-%
-Data flows may be implemented
- to support indirect communication across multiple hops;
- however, information from devices far away tends to be obsolete.
-%
-The environment can also be used, via stigmergy~\cite{DBLP:journals/scp/NicolaSI20},
- for indirect communication;
- however, a device can typically only access 
-  to a very local portion of the environment.
-%
-In other words, the system state is only \emph{partially observable}.
-In practice, CTDE technique aims to solve these problems, using a \emph{global} view at training time but
- a \emph{partial view} at execution time.
+% First attempts from RL viewpoint, 
+%  goes towards a so-called \textit{independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
+%  where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}.
 
-Furthermore, when the system node size is very large, novel works of Mean-Field Reinforcement Learning~\cite{DBLP:conf/ijcnn/ZhouZX21} are able to
- handle learning efficiently, reasoning about agent and agent neighbourhoods.
-%
-\paragraph{Learning goal: collective behaviour}
-%
-The devices of an aggregate 
- must cooperatively learn the ``aggregate program'',
- namely the collective behaviour 
- that achieves a particular \emph{global goal}
- in a decentralised, resilient way.
-%
+% Examples of these are Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
+%  Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value~\cite{DBLP:conf/atal/KaisersT10}.
+% %
+% The pro of these approaches is that their complexity does not scale up with the number of agents, 
+%  but, unfortunately, the learning process is extremely non-stationary and unstable.
+% % 
+% Furthermore, they are heuristic and do not exist any convergence proof (even if, in practice, they often reach a good policy).
 
-% \paragraph{Self-organisation, swarm robotics} it is condensed it other perspective, so I think it is better to avoid this paragraph.
-%
-% \suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
+% Other efforts have focussed on achieving equilibria such as Nash-equilibrium of Pareto optimality.
+%  The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q~\cite{DBLP:conf/icml/Littman94}.
+% The problem is that it is used with few agents and scale hardly with the application complexity.
+% %
+% Finally, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
+%  agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
+%  Most promising approaches in this direction are MAPPO~\cite{DBLP:journals/corr/abs-2103-01955}, COMA~\cite{DBLP:journals/corr/FoersterFANW17} and MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17}.
 
-\paragraph{Emergent behaviour}
-%
-The collective behaviour of an aggregate system
- \emph{emerges} 
- from a dense network of computations and interactions
- in an evolving environment.
-%
+% The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
+%  In AC instead, we need to find a technique that does not depend on the system node count.
 
-Typically, in CAS, single-agent RL brings to emergent behaviour and is used in various applications~\cite{DBLP:conf/icse/DAngeloGGGNPT19}.
+% \paragraph{Neighbour-based or indirect (environment-mediated) interaction}
+% %
+% In aggregate systems, a device can directly interact only with its neighbours.
+% %
+% Data flows may be implemented
+%  to support indirect communication across multiple hops;
+%  however, information from devices far away tends to be obsolete.
+% %
+% The environment can also be used, via stigmergy~\cite{DBLP:journals/scp/NicolaSI20},
+%  for indirect communication;
+%  however, a device can typically only access 
+%   to a very local portion of the environment.
+% %
+% In other words, the system state is only \emph{partially observable}.
+% In practice, CTDE technique aims to solve these problems, using a \emph{global} view at training time but
+%  a \emph{partial view} at execution time.
 
-\paragraph{Delayed, global rewards}
-%
-In the light of the above properties,
- if we consider a learning approach based on rewards,
- we must observe that such rewards would be
- delayed and mostly global---i.e.,
- it may not be easy to solve the credit assignment problem
- to distinguish good from bad individual contributions
- to the global outcome.
+% Furthermore, when the system node size is very large, novel works of Mean-Field Reinforcement Learning~\cite{DBLP:conf/ijcnn/ZhouZX21} are able to
+%  handle learning efficiently, reasoning about agent and agent neighbourhoods.
+% %
+% \paragraph{Learning goal: collective behaviour}
+% %
+% The devices of an aggregate 
+%  must cooperatively learn the ``aggregate program'',
+%  namely the collective behaviour 
+%  that achieves a particular \emph{global goal}
+%  in a decentralised, resilient way.
+% %
 
-A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14, DBLP:conf/aaai/FoersterFANW18}.
+% % \paragraph{Self-organisation, swarm robotics} it is condensed it other perspective, so I think it is better to avoid this paragraph.
+% %
+% % \suggestions{You can talk about similar problem structure (i.e. the collective follow the same local behaviour)}
+
+% \paragraph{Emergent behaviour}
+% %
+% The collective behaviour of an aggregate system
+%  \emph{emerges} 
+%  from a dense network of computations and interactions
+%  in an evolving environment.
+% %
+
+% Typically, in CAS, single-agent RL brings to emergent behaviour and is used in various applications~\cite{DBLP:conf/icse/DAngeloGGGNPT19}.
+
+% \paragraph{Delayed, global rewards}
+% %
+% In the light of the above properties,
+%  if we consider a learning approach based on rewards,
+%  we must observe that such rewards would be
+%  delayed and mostly global---i.e.,
+%  it may not be easy to solve the credit assignment problem
+%  to distinguish good from bad individual contributions
+%  to the global outcome.
+
+% A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14, DBLP:conf/aaai/FoersterFANW18}.
 It is used to capture agents contributions to the system's global performance.
-\paragraph{Transient phase}
-%
-When an aggregate behaviour is carried out,
- we often focus on the final stable outcomes after the transitory period.
-%
-However, intermediate results are also important,
- and invariants should hold throughout the entire computation.
+% \paragraph{Transient phase}
+% %
+% When an aggregate behaviour is carried out,
+%  we often focus on the final stable outcomes after the transitory period.
+% %
+% However, intermediate results are also important,
+%  and invariants should hold throughout the entire computation.
 
-\paragraph{Dynamic topology}
-%
-If we admit \emph{mobility} and \emph{failure} (rarely considered in related works, where node population is fixed),
- then we must also consider
- the possibility of changes in neighbourhoods
- and hence the need of dealing with dynamic topologies.
+% \paragraph{Dynamic topology}
+% %
+% If we admit \emph{mobility} and \emph{failure} (rarely considered in related works, where node population is fixed),
+%  then we must also consider
+%  the possibility of changes in neighbourhoods
+%  and hence the need of dealing with dynamic topologies.
 
-\subsection{Integrating Learning Mechanisms into Aggregate Computing}
-We now outline the different perspectives about AC with learning, 
- to clarify the current and the future research path.
-%
-\paragraph{Application} Learning can be useful at different levels in our framework.
- First of all, learning can be applied to the building blocks definition.
-% 
-Indeed, most of our results and applications are based on them.
-%
-The building block improvement may consist either of learning a correction factor or by entirely learning the behaviour.
+% \subsection{Integrating Learning Mechanisms into Aggregate Computing}
+% We now outline the different perspectives about AC with learning, 
+%  to clarify the current and the future research path.
+% %
+% \paragraph{Application} Learning can be useful at different levels in our framework.
+%  First of all, learning can be applied to the building blocks definition.
+% % 
+% Indeed, most of our results and applications are based on them.
+% %
+% The building block improvement may consist either of learning a correction factor or by entirely learning the behaviour.
 
-On the other side, learning can be useful also for framework related stuff,
- mainly aiming at "non-functional" aspects. 
- Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could take advantage of using Machine Learning algorithms.
+% On the other side, learning can be useful also for framework related stuff,
+%  mainly aiming at "non-functional" aspects. 
+%  Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could take advantage of using Machine Learning algorithms.
 
-\paragraph{Learning problem} 
-The learning problem in multi-agent settings can be posed as \textit{Independent} and \textit{Team} learning. 
-In the foster, each agent perform the learning process and while in the latter,
- the learning process is performed only by one agent and then shared with the collective (in a typical CDTE setting).
-% 
-Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
- --- i.e. each agent performs the same program. 
- Here, a common reward signal exists and evaluate the overall collective behaviour.
-% 
-In AC, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
- be applied seamlessly.
+% \paragraph{Learning problem} 
+% The learning problem in multi-agent settings can be posed as \textit{Independent} and \textit{Team} learning. 
+% In the foster, each agent perform the learning process and while in the latter,
+%  the learning process is performed only by one agent and then shared with the collective (in a typical CDTE setting).
+% % 
+% Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
+%  --- i.e. each agent performs the same program. 
+%  Here, a common reward signal exists and evaluate the overall collective behaviour.
+% % 
+% In AC, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
+%  be applied seamlessly.
 
-On the other hand, Independent Learning is usually applied in Collective
- (Self-)Adaptive System \cite{csas-and-marl}.
+% On the other hand, Independent Learning is usually applied in Collective
+%  (Self-)Adaptive System \cite{csas-and-marl}.
 
-In particular, Independent Reinforcement Learning reach good results in this kind of system, 
- even if no theorem exists as in the single-agent context.
-%
-With Independent learning is still possible to reach a global collective behaviour convergence. 
- Indeed using a common reward signal, the collective may reach the same behaviour through emergence \cite{iima2008swarm, nguyen2018swarm}.
+% In particular, Independent Reinforcement Learning reach good results in this kind of system, 
+%  even if no theorem exists as in the single-agent context.
+% %
+% With Independent learning is still possible to reach a global collective behaviour convergence. 
+%  Indeed using a common reward signal, the collective may reach the same behaviour through emergence \cite{iima2008swarm, nguyen2018swarm}.
 
-\paragraph{Learning technique}
-Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
- such as RL or Supervised Learning (SL).
-%
-SL is rare in CSAS -- or in the ``swarm" like system in general.
-%
-Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
- supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).
+% \paragraph{Learning technique}
+% Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
+%  such as RL or Supervised Learning (SL).
+% %
+% SL is rare in CSAS -- or in the ``swarm" like system in general.
+% %
+% Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
+%  supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).
 
-In AC, S) can be applied when we want to build up new building blocks from zero. 
-% 
-Indeed, we eventually know the right state of the system, but we do not know how the system can reach it.
+% In AC, S) can be applied when we want to build up new building blocks from zero. 
+% % 
+% Indeed, we eventually know the right state of the system, but we do not know how the system can reach it.
 
-RL is predominantly used in CAS, 
- due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
+% RL is predominantly used in CAS, 
+%  due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
 
-In our case, RL can be used both at the building block level and the framework level.
-%
-For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.
+% In our case, RL can be used both at the building block level and the framework level.
+% %
+% For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.
 
-\paragraph{Learning time}
-Finally, learning can be performed \textit{offline}, \textit{online} or \textit{offline an then online} (\textit{mixed time}).
-% 
-The foster is the common scenario in multi-agent settings.
- Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
-Mixed-time learning can be a mid-level complexity direction instead.
- Here, simulation can always be exploited to reach a good behaviour, but at runtime, 
- the behaviour can be adapted based on the new environmental situations. 
-Finally, we can imagine that learning can be performed online. 
- It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
- In mixed and online cases, RL is the most suitable model, because agents \textit{learn by doing}.
+% \paragraph{Learning time}
+% Finally, learning can be performed \textit{offline}, \textit{online} or \textit{offline an then online} (\textit{mixed time}).
+% % 
+% The foster is the common scenario in multi-agent settings.
+%  Here indeed, we can leverage simulation to know the right computational field and then use one of the novels CTDE techniques. 
+% Mixed-time learning can be a mid-level complexity direction instead.
+%  Here, simulation can always be exploited to reach a good behaviour, but at runtime, 
+%  the behaviour can be adapted based on the new environmental situations. 
+% Finally, we can imagine that learning can be performed online. 
+%  It is a common situation where we do not know the environment and also simulation can be very hard to accomplish. 
+%  In mixed and online cases, RL is the most suitable model, because agents \textit{learn by doing}.
 
@@
 \end{figure}
-As first experiments, we leverage RL to improve building blocks. 
+%As first experiments, we leverage RL to improve building blocks. 
 % 
@@
 In this article, we have discussed the combination of AC with Machine Learning techniques in order to improve adaptivity
- and effictveness of the AC framework.
-In doing so, we compared AC problems with current Machine Learning solutions in MASs systems.
+ and effectiveness of the AC framework.
+%In doing so, we compared AC problems with current Machine Learning solutions in MASs systems.
 %
-Finally, we explore the combination of AC with RL (in particular Hysteretic Q-Learning).
-RL application in this kind of settings is quite an explored -- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
- area for several reason, such us the highly system dynamic, non-stationarity,  partial enviroment observability, and non-bounded system size.
-In a case study we empirically shown the effictveness of RL, by which we speed up the convergence time of
- aggregate program taking in consideration -- the hop count.
+In particular, we explore the combination of AC with Hysteretic Q-Learning.
+RL application in this kind of setting is quite unexplored -- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
+ area for several reasons, such as the high system dynamic, non-stationarity,  partial environment observability, and non-bounded system size.
+In a case study, we empirically show the effectiveness of RL, by which we speed up the convergence time of
+ aggregate program taking into consideration -- the hop count.
 Furthermore, we have shown that the learnt program can easily scale up to system size.
 
-This is a first effort towards an ``hybrid'' collective programming approach, where the high-level program specification is 
- done by a designer and the low-level matters are leaved to machine learning algorithms.
+This is the first effort towards an ``hybrid'' collective programming approach, where the high-level program specification is 
+ done by a designer and the low-level matters are left to machine learning algorithms.
 

### 2021-11-28 paper-2021-woa-ac-ml.tex
compact biblio
pass on Sec II and III

@@
 \usepackage{systeme}
+
 \setminted[scala]{baselinestretch=1,bgcolor=white,frame=single}
@@
 \begin{document}
-
 \title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
@@
 For this reason, over the years, various approaches have tried to handle this complexity.
-Some of them take inspiration from the natural system composed of social animals.
-%--- like bees, fish, and ants.
+Some of them take inspiration from the natural system composed of social animals --- like bees, fish, and ants.
 %
@@
 
-Initially, researchers had tried to translate these properties to computing systems by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}. %, 
- %namely trying to find the right single-node-behaviour 
- %to reach a global behaviour. %, e.g. flocking, foraging, etc.
+Initially, researchers had tried to translate these properties to computing systems by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
+ namely trying to find the right single-node-behaviour 
+ to reach a global behaviour. %, e.g. flocking, foraging, etc.
 %
@@
 Indeed, an aggregate program 
- %(i.e. a program written with AC) 
+ (i.e. a program written with AC) 
  consists of the manipulation of \textit{computational fields}, a distributed
@@
  %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
-However, building block synthesis concerns different stuff not only related to the behaviour itself
- but also to the "ensemble" specifications. 
+However, building block synthesis concerns different stuffs not only related to the behaviour itself
+ but also to the ``ensemble" specifications. 
+%
 For instance, it is quite difficult to craft building blocks
@@
 %
-Ideally, we would like to continue building our application using these building blocks -- so the high-level specification remains the same -- but using some
- under-the-hood mechanisms that adapt the behaviour according to different environmental changes.
+Ideally, we would like to continue building our application using these building blocks 
+ -- so the high-level specification remains the same -- 
+ but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
+%
 For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
@@
 Collective Adaptive Systems are characterized by a very high number of nodes without a central coordinating authority that coordinate nodes. 
-The overall behaviour is a consequence of local interactions.
-The systems expose self-* properties.
+ The overall behaviour is a consequence of local interactions.%The systems expose self-* properties.
 
@@
  to express spatio-temporal computations~\cite{DBLP:journals/jlap/ViroliBDACP19}.
+%
 The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
@@
 In the next sections, we show the field calculus operators that are essential to understand the case study presented,
- and then we underline how each node computes in order to follow the aggregate program specification.
+ and then we underline how each node computes to follow the aggregate program specification.
+ Finally, we discuss the motivation that brings us to integration Reinforcement Learning in the AC framework.
 \subsection{Field Calculus operators}
-The main constructs that capture the essential aspects for programming self-organising systems are:
+The main constructs that capture the essential aspects for programming self-organising systems with field calculus are:
 \begin{itemize}
@@
 An aggregate program execution consists of a continuously node-point execution of a \round{}, 
- that is the atomical part of AC. Each node can interact only with-in its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.
+ which is the atomical part of AC. Each node can interact only within its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.
 
-The main round steps consist of:
+The main \round{} steps consist of:
 \begin{enumerate}
@@
 Rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
- This makes it possible to scale easily with the node number size. This collective, proactive and periodical execution will eventually lead to the satisfaction of the system-wide specification.
+ This makes it possible to scale easily with the node number size. 
+ This collective, proactive and periodical execution will eventually lead to the satisfaction of the system-wide specification.
 
@@
 
-A novel way to deal with this problem is to leave the low-level aspects of AC to machine learning, 
+A novel way to deal with this problem is to leave the low-level aspects of AC to Machine Learning algorithms, 
  training the collective in different environments and thus building general behaviours that are capable of adapting to many circumstances.
@@
  each agent collaborates to achieve a common goal guided by the same reward function.
-In this setting, several works have tried to apply the algorithms known in the single-agent context to the multi-agent one - the so-called Independent MARL~\cite{DBLP:journals/tsmc/BusoniuBS08}. 
+%
+In the cooperative and homogeneous systems, several works have tried to apply the algorithms used in the single-agent context to the multi-agent one --- 
+ the so-called Independent MARL~\cite{DBLP:journals/tsmc/BusoniuBS08}. 
+%
 However, several issues emerge, such as non-stationarity, multi-agent credit assignment, scalability and partial observability of the system. 
  Therefore, several heuristics have tried to solve these problems.
+%
 Among many Hysteric Q-Learning~\cite{hysteretic-q} uses an optimistic approach in which more weight is given to positive actions than negative ones.
@@
 Another way to handle multi-agent learning is to use so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 
- In this case, a central agent collects the experiences of the various agents and distils a policy that each agent can execute locally in a typical actor-critic setting. 
- During the learning phase, the problems of partial observability and non-stationarity are reduced, since the agent has a global view. 
+ In this case, a central agent collects the experiences of the various agents and then distils a policy that each agent can execute locally in a typical actor-critic setting. 
+%
+During the learning phase, the problems of partial observability and non-stationarity are reduced, since the centralised learner has a global view. 
  However, the problem of scalability is typically not addressed.
 
-In the next section, we exploit Hysteretic Q-Learning (since it empirically showed good performance) with AC, as the first step towards \textit{``hybrid"} techniques for programming aggregate systems.
+In the next section, we exploit Hysteretic Q-Learning (since it empirically showed good performance~\cite{DBLP:journals/tsg/HadidiJ13}) with AC, 
+ as the first step towards \textit{``hybrid"} techniques for programming aggregate systems~\cite{research}.
  In this way, an application part is described in terms of aggregate computations and the low-level part is left to learning.
@@
   Each agent starts with his copy of a Q table. 
-  Then, during the simulation, they refine them Q tables changing q values according to their experience (different colours mean different values). 
+  Then, during the simulation, they refine them Q tables changing Q values according to their experience (different colours mean different values). 
   In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
@@
 
-As a first step towards a complete integration between AC and RL, we decide to set up an Independent learning technique, 
- by which each agent try to improve our current local policy (shown in \Cref{fig:aggregate-q-learning}). 
+As a first step towards a complete integration between AC and RL, we haved decided to set up an Independent RL learning technique, 
+ by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 
 % 
@@
 %
-In particular, each agent maintains a local Q table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
+In particular, each agent updates a local Q table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
 $$
@@
 
-At the end of the episodes, each agent has collected its local experience, which can vary in different zones. So, 
- we combine the N (the number of nodes) Q table selecting the maximum value for each state-action pair since we assume a homogeneous behaviour:
+At the end of each episode, each agent has collected its local experience, which can vary in different zones. 
+ So, we combine the N (the number of nodes) Q table selecting the maximum value for each state-action pair since we assume an optimistic homogeneous behaviour:
 $$
@@
 \bibliographystyle{IEEEtran}
-\bibliography{biblio}
+\bibliography{biblio-reduced}
 %%

### 2021-11-28 paper-2021-woa-ac-ml.tex
pass on Sec IV and V

@@
 
- \section{Background and Related Work}\label{background}
+ \section{Background and Motivation}\label{background}
 
@@
  aggregate program: the hop count.
-In this case, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes ((\Cref{fig:hop-count})).
+In this case, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes (\Cref{fig:hop-count}).
 
@@
 \end{minted}
-\mintinline{scala}{recentValues(data, k)} is an aggregate operator that returns a list containing the last k data perceived by the agent.
+\mintinline{scala}{recentValues(data, k)} is an aggregate operator that returns a list containing the last\textbf{\texttt{k}} data perceived by the agent.
 Then, due to the partial observability of the environment, we maintain a 
@@
   Using it, they evaluate the action to speed up the hop count convergence following a policy. 
-  This new refined output is then used to evaluate the reward and to refine the Q table (dotted line path)}
+  This new refined output is then used to evaluate the reward and to improve the Q table (dotted line path)}
   \label{fig:aggregate-learning-loop}
 \end{figure}
-%\begin{minted}[obeytabs=true,tabsize=1]{scala}
-%val inf = Double.PositiveInfinity
-%rep((initState, initPolicy, inf)) {
-%  (oldState, policy, hopCount) => 
-%      val currentState = 
-%        state(hopCount, k, w)
-%      val action = policy(state)
-%      val output = mux(source) 
-%        { 0 } 
-%        { minHood(hopCount) 
-%          + 1 
-%          + action
-%        } 
-%      }
-%      val currentReward = 
-%        reward(output)
-%      val improvedPolicy = 
-%        qLearning(oldState, 
-%          state, 
-%          action, 
-%          reward,
-%          policy
-%      (currentState, policy, output)
-%    )
-%}
-%\end{minted}
 
@@
 where \textbf{i} is the node identifier and \textbf{t} is the time when the error is evaluated.
-%
 
@@
  entire aggregate and we test the program changing the system size (80, 360 and 760 nodes).
+%
 For each population size, we run 50 episodes and then we average the error at each time step. 
@@
  and effectiveness of the AC framework.
-%In doing so, we compared AC problems with current Machine Learning solutions in MASs systems.
 %
 In particular, we explore the combination of AC with Hysteretic Q-Learning.
-RL application in this kind of setting is quite unexplored -- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
+%
+RL application in this kind of setting is quite unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
  area for several reasons, such as the high system dynamic, non-stationarity,  partial environment observability, and non-bounded system size.
@@
  Local observations may not be statistically significant in summarising the state of the entire aggregate. 
- In the future, we could address this problem through recurrent networks and a CTDE learning methodology.
+ In the future, we could address this problem through recurrent neural networks and a CTDE learning methodology.
 
@@
  Another possible direction is to use it at the framework level, e.g., to select which messages to send to the neighbourhood or not. 
- This has already been addressed in the literature, usually under the name of \textit{learn to communicate}~\cite{DBLP:journals/corr/abs-1908-03963}. 
+ This has already been addressed in the literature under the name of \textit{learn to communicate}~\cite{DBLP:journals/corr/abs-1908-03963}. 
 A novel way to handle this problem is to use Graph Neural Networks~\cite{DBLP:journals/corr/abs-1812-08434} to learn the representation of a distributed system and to select messages based on that representation.
@@
 
-In this case, we applied a variation of the independent learning approach. 
+In our experiment, we have applied a variation of the independent learning approach. 
  Agents first learn the best policy locally, and then at the end of the episode, a learner merges the Q-tables. 

### 2021-11-28 paper-2021-woa-ac-ml.tex
Pass on intro

@@
  has to emerge from local behaviours and interactions. 
-For this reason, over the years, various approaches have tried to handle this complexity.
-Some of them take inspiration from the natural system composed of social animals --- like bees, fish, and ants.
 %
-In this case, a sort of hive mind arises 
- %-- also referred to as swarm intelligence --  
- born from local interactions between agents. 
+For this reason, over the years, various approaches have tried to handle this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.).
 %
-These kinds of behaviours are accomplished with self-organisation, 
- namely a global order that emerges from continuous 
- interactions between simple entities.
+%In this case, a sort of hive mind arises 
+% %-- also referred to as swarm intelligence --  
+% born from local interactions between agents. 
+%
+Specifically, such approaches tend to exploit 
+ \emph{self-organisation}, 
+ i.e., processes by which global order emerges 
+ from continuous interactions among simple entities.
 
-Initially, researchers had tried to translate these properties to computing systems by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
- namely trying to find the right single-node-behaviour 
- to reach a global behaviour. %, e.g. flocking, foraging, etc.
+Initially, researchers had tried to achieve these properties in computing systems by miming natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
+ namely trying to find the right single-node-behaviour (local rules)
+ to pursue the intended global behaviour. %, e.g. flocking, foraging, etc.
 %
-However, this mapping cannot be easily defined, since we are dealing with complex systems.
+However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
 %
-Ideally, we would like to express collective behaviours at the ensemble level,
- abstracting over non-functional aspects such as the network topology, environment uncertainties and 
- system size.
-
-In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is an innovative paradigm, according to which
- developers can express collective self-organising behaviours. % through a functional program specification.
+A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
+%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
+In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} as proposed:
+ it is a paradigm by which
+ developers can express collective self-organising behaviours by a global perspective. % through a functional program specification.
 %
@@
  (i.e. a program written with AC) 
- consists of the manipulation of \textit{computational fields}, a distributed
- data structure where each node is associated with the result of a computation
- performed by it.
+ consists of the manipulation of \textit{computational fields}, i.e., distributed
+ data structures where each node is associated with the result of a computation performed by it,
+ and neighbourhood observation.
 %

### 2021-11-28 paper-2021-woa-ac-ml.tex
On background

@@
 AC is studied in various simulated scenarios ranging from smart cities to crowd management~\cite{DBLP:journals/jlap/ViroliBDACP19}, 
- thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
- this paradigm, such as ScaFi~\cite{DBLP:conf/isola/CasadeiVAD20}. %, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
+ thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and programming languages and tools that support 
+ this paradigm~\cite{DBLP:journals/jlap/ViroliBDACP19}, such as the ScaFi aggregate programming language~\cite{DBLP:conf/isola/CasadeiVAD20}.
+%, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.
 
-In the next sections, we show the field calculus operators that are essential to understand the case study presented,
- and then we underline how each node computes to follow the aggregate program specification.
- Finally, we discuss the motivation that brings us to integration Reinforcement Learning in the AC framework.
+In the next sections, we summarise the field calculus language and computation model, which are essential to understand the case study of \Cref{evaluation}.
+Then, we discuss the motivation that brought us to integrate Reinforcement Learning into the AC framework.
+
 \subsection{Field Calculus operators}
@@
 \end{itemize}
-For a more in-depth explanation of these operators, please refer to~\cite{DBLP:journals/computer/BealPV15}.
+%
+A full coverage of the field calculus is beyond the scope of this paper. A more comprehensive presentation is given in~\cite{DBLP:journals/jlap/ViroliBDACP19}.
 

### 2021-11-28 paper-2021-woa-ac-ml.tex
On execution model

@@
 \subsection{Computational Model}
-An aggregate program execution consists of a continuously node-point execution of a \round{}, 
- which is the atomical part of AC. Each node can interact only within its neighbourhoods sharing the \export{}, namely the result produced by the evaluation of a \round{}.
-
-The main \round{} steps consist of:
+%
+For an aggreate program to yield collective adaptive behaviour, 
+ continuous computation and communication is needed.
+%
+An individual atomic step 
+ involving context acquisition,
+ computation of the program,
+ and neighbourhood-based communication 
+ is called a \round{}.
+%
+In detail, a \round{} run at some device consists of:
 \begin{enumerate}
-  \item execution context creation: nodes create the execution context collecting information from the sensors, 
-  the neighbourhood messages, and the old program output;
-  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
-  This produces the \export{};
-  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes;
-  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...).
+  \item \emph{context acquisition}: the device collects information from the sensors, 
+  the messages from neighbours
+  (including the device itself---to model state);
+  \item \emph{program evaluation}: the aggregate program is evaluated against the acquired context,
+  yielding an \export{}, namely the message to be sent to neighbours;
+  \item \emph{export sharing}: the \export{} is sent to the neighbours;
+  \item \emph{actuations}: the \export{} also includes information that can be used to drive actuators in the local node.
 \end{enumerate}
 
-Rounds execution is completely asynchronous, so do not exist a global clock to coordinate the aggregate. 
- This makes it possible to scale easily with the node number size. 
- This collective, proactive and periodical execution will eventually lead to the satisfaction of the system-wide specification.
+Rounds execution is completely asynchronous, so there is no global clock or barrier to coordinate the aggregate. 
+%
+ This makes it possible to scale easily with the number of nodes. 
+%
+The combination of the logic of the aggregate program
+ and 
+ such a collective, proactive, and periodical execution
+ can promote the pursuing of collective adaptive behaviour.
 

### 2021-11-28 paper-2021-woa-ac-ml.tex
On motivation

@@
 \subsection{Motivation}
-Currently, Aggregate computing has a problem in the ability to generalise collective behaviours that are capable of working well in different scenarios. 
- For example, a well-tested aggregate application in a static network could lead to a non-stable behaviour in a dynamic network. 
- Typically, to alleviate this problem, designers create ad-hoc building blocks that cope with certain environmental aspects (such as the ULT block~\cite{DBLP:conf/saso/AudritoCDV17}). 
- But they are still context-specific solutions.
 
-A novel way to deal with this problem is to leave the low-level aspects of AC to Machine Learning algorithms, 
+Consider the problem of associating
+ to each node of the network
+ a value
+ representing the estimation of the minimum distance
+ from that node to the closest \emph{source} node (see \Cref{fig:hop-count}).
+%
+This is indeed a computational field 
+ and it is known as a \emph{gradient}~\cite{DBLP:conf/saso/AudritoCDV17}
+ and there exist aggregate algorithms
+ for computing it in such a way that
+ the field automatically adapts 
+ to changes induced by mobility
+  and addition/removal of sources.
+%
+The gradient is a key building block 
+ for building more complex self-organising behaviour~\cite{DBLP:journals/jlap/ViroliBDACP19}.
+
+%Currently, aggregate computing has a problem in the ability to generalise collective behaviours that are capable of working well in different scenarios. 
+%%
+% For example, a well-tested aggregate application in an environment with limited dynamics could lead to a non-stable behaviour under continuous perturbations.
+%% 
+% Typically, to alleviate this problem, designers create ad-hoc building blocks that cope with certain environmental aspects (such as the ULT block~\cite{DBLP:conf/saso/AudritoCDV17}). 
+% But they are still context-specific solutions.
+However, in practice, 
+ specific gradient algorithms
+ may be developed
+ to achieve different trade-offs 
+ in terms of reactivity to change and stability~\cite{DBLP:conf/saso/AudritoCDV17},
+ which make them more or less suitable depending on 
+ the environmental and network dynamics at hand.
+%
+The problem is that manually crafting such algorithms
+ is not an easy task, for various heuristics and assumptions typically have to be considered.
+%
+To address this problem,
+ we propose to use machine learning techniques
+ in order to learn ``good'' local decision-making.
+% 
+That is, the idea is to leave the low-level aspects of AC to Machine Learning algorithms, 
  training the collective in different environments and thus building general behaviours that are capable of adapting to many circumstances.
@@
 %
-Among many Hysteric Q-Learning~\cite{hysteretic-q} uses an optimistic approach in which more weight is given to positive actions than negative ones.
-
+Among many, Hysteric Q-Learning~\cite{hysteretic-q} uses an optimistic approach in which more weight is given to positive actions than negative ones.
+%
 Another way to handle multi-agent learning is to use so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 

### 2021-11-28 paper-2021-woa-ac-ml.tex
Shorter org summary

@@
 
-The paper is divided as follow, in \Cref{background} we describe
- the AC framework, showing the context in which
- is typically applied and the overall idea of how it works, and then the motivation of injecting Reinforcement Learning in this paradigm,
+The paper is organised as follows. 
+%
+In \Cref{background} we describe
+ the AC framework, %showing the context in which is typically applied and the overall idea of how it works, and then the motivation of injecting Reinforcement Learning in this paradigm,
+ and motivate our proposal to inject RL into this paradigm. 
  %in \Cref{aggregate-and-machine-learning} we detail the relevant properties of Aggregate Computing in terms of machine learning problems, 
- in \Cref{aggregate-and-rl} we describe a combination of AC with RL and then in
- \Cref{evaluation} we practically applied it in a paradigmatic AC case study,
- finally, \Cref{conclusion} we discuss the result showed and we underline a path toward an in-depth AC and RL integration.
+ In \Cref{aggregate-and-rl}, we describe the details of our integration of AC with RL.
+Then, in  \Cref{evaluation} we evaluate the approach on a paradigmatic AC case study.
+%
+Finally, in \Cref{conclusion} we discuss the results and the path towards deeper AC-RL integration.
 

### 2021-11-28 paper-2021-woa-ac-ml.tex
A little bit of more motivation about gradient example
Specify program execution and computation loop are those of AC

@@
 %
-The gradient is a key building block 
- for building more complex self-organising behaviour~\cite{DBLP:journals/jlap/ViroliBDACP19}.
+This example is significant as the gradient is a key building block 
+ for more complex self-organising behaviour~\cite{DBLP:journals/jlap/ViroliBDACP19}.
 
@@
 % 
-In this section, we briefly describe how RL can be injected into the computation loop,
+In this section, we briefly describe how RL can be injected into the AC computation loop,
  and what RL we used with AC.
 
-From a local view, program execution can be seen as:
+From a local view, AC program execution can be seen as:
 $$

### 2021-11-28 paper-2021-woa-ac-ml.tex
Oxford comma needed
Explain hopCount in comments + minors in code style

@@
 
-Reward function instead, might be depending on the learning technique, time and problem. 
+Reward function instead, might be depending on the learning technique, time, and problem. 
  Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
@@
 A naive implementation of this kind of problem can be described in AC as:
-\begin{minted}[fontsize=\scriptsize]{scala}
-def hopCount(): Int = {
-  rep(Double.PositiveInfinity) { 
-    hopCounts => mux(source) 
-    { 0 } 
-    { minHoodPlus(nbr{hopCounts}) + 1 }
-  }
+\begin{minted}[fontsize=\scriptsize,mathescape]{scala}
+def hopCount(): Int = 
+  // update state (initially, $\infty$; now, hopCounts) by ...
+  rep(Double.PositiveInfinity) { hopCounts => 
+    // source returns 0,
+    // non-sources minimise over neighbourdhood and add 1
+    mux(source) { 0 }{ minHoodPlus(nbr{hopCounts}) + 1 }
   }
@@
 def hopCountRl(k: Int, w: Int): Int = 
-{
-  rep(Double.PositiveInfinity) { 
-    hopCounts => 
+  rep(Double.PositiveInfinity) { hopCounts => 
       val currentState = 
@@
       hopCout() + action
-  }
   }

### 2021-11-28 paper-2021-woa-ac-ml.tex
Minor space-saving
On future work beginning

@@
 $$
-Furthermore, we use a $\epsilon$ decay policy for the $\epsilon$-greedy policy as:
+Moreover, we use a $\epsilon$ decay policy for the $\epsilon$-greedy policy as:
 \begin{equation*}
@@
 \end{equation*}
-
+%
 where \textbf{i} is the node identifier and \textbf{t} is the time when the error is evaluated.
@@
 
-This is the first effort towards an ``hybrid'' collective programming approach, where the high-level program specification is 
+To the best of our knowledge, this is the first effort towards an ``hybrid'' collective programming approach, where the high-level program specification is 
  done by a designer and the low-level matters are left to machine learning algorithms.
-
-We know that this is preliminary work and it shed light on several problems that we should handle in the future.
+%
+As such, this work shed light on several issues deserving future investigation.
 
@@
  However, this does not reduce the problems of scalability and non-stationary, since a central agent must still collect all the Q-tables of the aggregate. 
- In particular, to handle the scalability issues, we will have to consider decentralised approaches based only on the neighbourhood information (or local leaders).
+ In particular, to handle the scalability issues, we will have to consider decentralised approaches based only on neighbour data (or local leaders).
 

### 2021-11-28 paper-2021-woa-ac-ml.tex
change title
remove clickable links
fix minors

@@
 \usepackage{subcaption}
-\usepackage[hidelinks]{hyperref}
+\usepackage[hidelinks, draft]{hyperref}
 \usepackage[inline]{enumitem}
@@
 \begin{document}
-\title{Towards a Machine Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
+\title{Towards a Reinforcement Learning approach for Aggregate Computing} %% or Towards a Machine Learning approach for Aggregate Computing
 \author{
@@
 %, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
-In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} as proposed:
+In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
  it is a paradigm by which
@@
 %
-A natural way to describe this kind of problem is through a \textit{collective} view of the system.
+Instead, a natural way to describe this kind of problems should be through a \textit{collective} view of the system.
  This is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
@@
 % 
-Also, ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
+Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
  a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)
@@
 \end{minted}
-\mintinline{scala}{recentValues(data, k)} is an aggregate operator that returns a list containing the last\textbf{\texttt{k}} data perceived by the agent.
+\mintinline{scala}{recentValues(data, k)} is an aggregate operator that returns a list containing the last \textbf{\texttt{k}} data perceived by the agent.
 Then, due to the partial observability of the environment, we maintain a 
- list of velocity, to understand the hop count direction:
+ list of velocities, to understand the hop count direction:
 
@@
 \subsection{Simulation Setup}
-We verify our learning algorithm through simulation~\footnote{repository at \url{https://github.com/cric96/alpaca-2022-ac-rl}} leveraging Alchemist and ScaFi.
+We verify our learning algorithm through simulation\footnote{repository at \url{https://github.com/cric96/alpaca-2022-ac-rl}} leveraging Alchemist and ScaFi.
  To check if the aggregate eventually speeds up the hop count convergence time, we create a situation where the rising problem emerges (\Cref{fig:hop-count}):
- in the beginning, two source nodes exist in the network, ($A$ and $B$).
+ firstly, two source nodes exist in the network, ($A$ and $B$).
  At the time $t$, $B$ disappears. So the nodes near $B$ start a slow process

### 2021-11-28 paper-2021-woa-ac-ml.tex
minors: remove a comma in section III
grammars minors

@@
 \usepackage{subcaption}
-\usepackage[hidelinks, draft]{hyperref}
+\usepackage[draft]{hyperref}
 \usepackage[inline]{enumitem}
@@
  it is a paradigm by which
- developers can express collective self-organising behaviours by a global perspective. % through a functional program specification.
+ developers can express collective self-organising behaviours from a global perspective. % through a functional program specification.
 %
@@
  %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
-However, building block synthesis concerns different stuffs not only related to the behaviour itself
+However, building block synthesis concerns different stuff not only related to the behaviour itself
  but also to the ``ensemble" specifications. 
@@
 %
-For an aggreate program to yield collective adaptive behaviour, 
- continuous computation and communication is needed.
+For an aggregate program to yield collective adaptive behaviour, 
+ continuous computation and communication are needed.
 %
@@
  such a collective, proactive, and periodical execution
- can promote the pursuing of collective adaptive behaviour.
+ can promote the pursuit of collective adaptive behaviour.
 
@@
 
-As a first step towards a complete integration between AC and RL, we haved decided to set up an Independent RL learning technique, 
+As a first step towards a complete integration between AC and RL, we have decided to set up an Independent RL learning technique, 
  by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 

### 2021-11-29 paper-2021-woa-ac-ml.tex
minor: 'an hybrid' => 'a hybrid'

@@
 
-To the best of our knowledge, this is the first effort towards an ``hybrid'' collective programming approach, where the high-level program specification is 
+To the best of our knowledge, this is the first effort towards a ``hybrid'' collective programming approach, where the high-level program specification is 
  done by a designer and the low-level matters are left to machine learning algorithms.

### 2021-11-29 paper-2021-woa-ac-ml.tex
final Musolesi pass

@@
 %
-For this reason, over the years, various approaches have tried to handle this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.).
+For this reason, over the years, various approaches have addressed the problem of dealing with this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.).
 %
@@
 
-Initially, researchers had tried to achieve these properties in computing systems by miming natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
+Initially, researchers had tried to build systems that show these properties in computing systems by mimickng natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02},
  namely trying to find the right single-node-behaviour (local rules)
- to pursue the intended global behaviour. %, e.g. flocking, foraging, etc.
+ to achieve the intended global behaviour. %, e.g. flocking, foraging, etc.
 %
@@
  %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
-However, building block synthesis concerns different stuff not only related to the behaviour itself
+However, building block synthesis concerns different elements not only related to the behaviour itself
  but also to the ``ensemble" specifications. 
@@
 For instance, it is quite difficult to craft building blocks
- that easily adapt w.r.t. the node movements rapidity.
+ that easily adapt to the non stationarity due to node movements.
 %
-Ideally, we would like to continue building our application using these building blocks 
+Ideally, we would like to continue developing our applications using these building blocks 
  -- so the high-level specification remains the same -- 
@@
 %
-For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
- to developers of defining the right collective specifications.
+For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden of defining the right collective specifications to developers.
 %This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
@@
 
-Collective Adaptive Systems are characterized by a very high number of nodes without a central coordinating authority that coordinate nodes. 
+Collective Adaptive Systems are characterized by a very high number of nodes without a central authority that coordinate nodes. 
  The overall behaviour is a consequence of local interactions.%The systems expose self-* properties.
@@
 Instead, a natural way to describe this kind of problems should be through a \textit{collective} view of the system.
- This is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
- where the behaviours are designed \textit{top-down}, and then the local behaviour is generated automatically from this ``system-wide'' specification. 
+ This is the idea used by the so-called \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
+ where the behaviours are designed \textit{top-down}, and then the local one is generated automatically from this ``system-wide'' specification. 
 % 
@@
  the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
-The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
- map from any devices to their computational value (i.e. the result of a computation).
+The latter could be seen as the application of the mathematical concept of field: it consists of a 
+ map from any devices to their computational value (i.e., the result of a computation).
 The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
@@
 %
-The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
- using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
-By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency %~\cite{DBLP:journals/taas/BealVPD17} 
+The novelty of AC resides in its composable (functional-like) system-size independent program definition: 
+ using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at application level.
+By opportunistically crafting these building-blocks, it is possible to verify prominent properties such as eventual consistency %~\cite{DBLP:journals/taas/BealVPD17} 
  (i.e. the ensemble behaviour is independent of the underlying network details)
@@
   mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
-  For instance, \mintinline{scala}{minHood} return the minimum value 
+  For instance, \mintinline{scala}{minHood} returns the minimum value 
   within the neighbourhood.
@@
 %
-A full coverage of the field calculus is beyond the scope of this paper. A more comprehensive presentation is given in~\cite{DBLP:journals/jlap/ViroliBDACP19}.
+A full coverage of field calculus is beyond the scope of this paper. A more comprehensive presentation is given in~\cite{DBLP:journals/jlap/ViroliBDACP19}.
 
@@
 %
-Another way to handle multi-agent learning is to use so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 
+Another way to handle multi-agent learning is to use the so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 
  In this case, a central agent collects the experiences of the various agents and then distils a policy that each agent can execute locally in a typical actor-critic setting. 
@@
   \caption{This figure shows the main steps of RL in AC.
-  Each agent starts with his copy of a Q table. 
-  Then, during the simulation, they refine them Q tables changing Q values according to their experience (different colours mean different values). 
+  Each agent starts with his copy of a Q-table. 
+  Then, during the simulation, they refine them Q-tables changing Q values according to their experience (different colours mean different values). 
   In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
-  Finally, a new episode starts using this $Q_{max}$ as a starting Q table.
+  Finally, a new episode starts using this $Q_{max}$ as a starting Q-table.
   }
@@
 
-From a local view, AC program execution can be seen as:
+From a local point of view, AC program execution can be seen as:
 $$
@@
 
-Speaking of RL settings, we need to define the state function and the reward function.
+As far as RL settings are concerned, we need to define the state function and the reward function.
 %
@@
 
-Reward function instead, might be depending on the learning technique, time, and problem. 
+Instead, reward function might be dependent on the learning technique, time, and problem. 
  Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
@@
 $$
-As remarked before, the policy $\pi$ need to be distributed, so each agent can choose the action that will be performed:
+As remarked before, the policy $\pi$ needs to be distributed, so that each agent can choose the action that will be performed:
 $$
@@
 
-As a first step towards a complete integration between AC and RL, we have decided to set up an Independent RL learning technique, 
+As a first step towards a complete integration between AC and RL, we have decided to adopt an Independent RL learning technique, 
  by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 
@@
 %
-In particular, each agent updates a local Q table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
+In particular, each agent updates a local Q-table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
 $$
@@
 
-At the end of each episode, each agent has collected its local experience, which can vary in different zones. 
- So, we combine the N (the number of nodes) Q table selecting the maximum value for each state-action pair since we assume an optimistic homogeneous behaviour:
+At the end of each episode, each agent collects its local experience, which can vary in different zones. 
+ So, we combine the $N$ (the number of nodes) Q-table selecting the maximum value for each state-action pair since we assume an optimistic homogeneous behaviour:
 $$
@@
 The idea is to maintain the nodes in a stable situation or to
- fasten the time needed to reach a stable one.
+ speed up the time needed to reach a stable one.
 %
@@
   each agent computes its current state. 
-  Using it, they evaluate the action to speed up the hop count convergence following a policy. 
-  This new refined output is then used to evaluate the reward and to improve the Q table (dotted line path)}
+  Then, an agent evaluates the action to speed up the hop count convergence following a policy. 
+  This new refined output is then used to evaluate the reward and to improve the Q-table (dotted line path)}
   \label{fig:aggregate-learning-loop}
@@
 We use 80 ($N$) nodes displaced into an irregular grid with a width of 100 meters and a height of 20 meters. 
- Each node has a neighbourhoods range equal to 8 meters --- on average, each node has 7 neighbours. 
+ Each node has a neighbourhood range equal to 8 meters --- on average, each node has 7 neighbours. 
  The time $t$ is fixed to 20.
- Each simulation episode lasts 60-time unit. Finally, we set $k = 3$ and $w = 2$.
+ Each simulation episode lasts 60 time-unit. Finally, we set $k = 3$ and $w = 2$.
 % 
@@
 
-Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b,c,d). To do that, we share the learnt Q-table to the 
+Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b, c, d). To do that, we share the learnt Q-table to the 
  entire aggregate and we test the program changing the system size (80, 360 and 760 nodes).
@@
 For each population size, we run 50 episodes and then we average the error at each time step. 
- As a result, the aggregate speed-up convergence in each new situation.
+The aggregate speed-up convergence is observed for each scenario taken into consideration.
 
@@
  and effectiveness of the AC framework.
-%
 In particular, we explore the combination of AC with Hysteretic Q-Learning.
-%
-RL application in this kind of setting is quite unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
- area for several reasons, such as the high system dynamic, non-stationarity,  partial environment observability, and non-bounded system size.
-In a case study, we empirically show the effectiveness of RL, by which we speed up the convergence time of
- aggregate program taking into consideration -- the hop count.
+The application of RL to this kind of setting is a largely unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
+ research area. There are a variety of challenges, such as non-stationarity,  partial environment observability, and non-bounded system size.
+In a case study, we have empirically showed the effectiveness of RL by showing a speed up of the convergence time of
+ aggregate program we took into consideration, namely the hop count.
 Furthermore, we have shown that the learnt program can easily scale up to system size.
@@
 To the best of our knowledge, this is the first effort towards a ``hybrid'' collective programming approach, where the high-level program specification is 
- done by a designer and the low-level matters are left to machine learning algorithms.
+ done by a designer and the low-level implementation to building blocks based on machine learning.
 %
-As such, this work shed light on several issues deserving future investigation.
+As such, this work has shed light on several issues deserving future investigation.
 
@@
  In the future, we could address this problem through recurrent neural networks and a CTDE learning methodology.
-
-The learning, in this case, was applied at the building block level. 
- Another possible direction is to use it at the framework level, e.g., to select which messages to send to the neighbourhood or not. 
+ Learning algorithms were implemented at building block level. 
+ Another possible direction is to use them at the framework level, e.g., to select which messages to send to the neighbourhood or not. 
  This has already been addressed in the literature under the name of \textit{learn to communicate}~\cite{DBLP:journals/corr/abs-1908-03963}. 
@@
 
-In our experiment, we have applied a variation of the independent learning approach. 
+In our experiments, we have applied a variation of the independent learning approach. 
  Agents first learn the best policy locally, and then at the end of the episode, a learner merges the Q-tables. 
@@
 Finally, we do not address the multi-agent credit assignment problem in this work. 
- Indeed, we use a global reward function for the whole aggregate. In this way, however, learning often degenerated because it was difficult to understand the influence of an agent w.r.t. the system. 
- To reduce this problem, we will consider novel techniques of difference reward~\cite{DBLP:conf/atal/AgoginoT04} or counterfactual baseline~\cite{DBLP:journals/corr/FoersterFANW17}.
+ Indeed, we use a global reward function for the whole aggregate. In this way, however, learning often degenerated because it was difficult to understand the influence of an agent on the system. 
+ To reduce this problem, we plan to study the application of difference rewards~\cite{DBLP:conf/atal/AgoginoT04} or counterfactual baselines~\cite{DBLP:journals/corr/FoersterFANW17}.
 

### 2021-11-29 paper-2021-woa-ac-ml.tex
minors

@@
 %
-For this reason, over the years, various approaches have addressed the problem of dealing with this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.).
+For this reason, over the years, various approaches have addressed the problem of dealing with this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, beehives, flocks of birds etc.).
 %
@@
 
-Initially, researchers had tried to build systems that show these properties in computing systems by mimickng natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02},
+Initially, researchers had tried to build systems that show these properties in computing systems by mimicking natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02},
  namely trying to find the right single-node-behaviour (local rules)
@@
 For instance, it is quite difficult to craft building blocks
- that easily adapt to the non stationarity due to node movements.
+ that easily adapt to the non-stationarity due to node movements.
 %
@@
 
-Instead, reward function might be dependent on the learning technique, time, and problem. 
+Instead, the reward function might be dependent on the learning technique, time, and problem. 
  Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
@@
  research area. There are a variety of challenges, such as non-stationarity,  partial environment observability, and non-bounded system size.
-In a case study, we have empirically showed the effectiveness of RL by showing a speed up of the convergence time of
- aggregate program we took into consideration, namely the hop count.
+In a case study, we have empirically shown the effectiveness of RL by showing a speedup of the convergence time of
+ the aggregate program we took into consideration, namely the hop count.
 Furthermore, we have shown that the learnt program can easily scale up to system size.
